---
title: "Building Energy Efficiency: Regression Analysis Framework"
author: "Jakes Chege"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 11,
  fig.height = 7
)
```

```{r figures-dir, include=FALSE}
# Directory where all plot files are written
if (!dir.exists("figures")) dir.create("figures")
```

---

# Setup & Data Preparation

## Load Packages

```{r packages}
pkgs <- c(
  "tidyverse", "caret", "randomForest", "xgboost", "nnet",
  "mgcv", "splines", "corrplot", "ggplot2", "dplyr",
  "pdp", "vip", "readxl", "writexl", "gridExtra",
  "car", "MASS", "glmnet", "Metrics", "reshape2",
  "patchwork", "ggcorrplot", "GGally"
)

# Install any missing packages silently
installed <- rownames(installed.packages())
to_install <- pkgs[!pkgs %in% installed]
if (length(to_install) > 0) install.packages(to_install, repos = "https://cloud.r-project.org")

invisible(lapply(pkgs, library, character.only = TRUE))
```

## Load & Preprocess Data

```{r load-data}
df_raw <- read_excel("C:/Users/Light House/Desktop/MastersDSA/semThree-DSA/Linear Models/final_project/data/ENB2012_data 2.xlsx")

colnames(df_raw) <- c(
  "RelCompactness", "SurfaceArea", "WallArea",     "RoofArea",
  "OverallHeight",  "Orientation", "GlazingArea",  "GlazingAreaDist",
  "HeatingLoad",    "CoolingLoad"
)

# Factor versions of categorical variables
df_raw <- df_raw %>%
  mutate(
    Orientation_f     = factor(Orientation),
    OverallHeight_f   = factor(OverallHeight, labels = c("Low_3.5m", "High_7m")),
    GlazingArea_f     = factor(GlazingArea),
    GlazingAreaDist_f = factor(GlazingAreaDist),
    # Derived / interaction features (from EDA)
    CompactnessXHeight  = RelCompactness * OverallHeight,
    WallToRoofRatio     = WallArea / RoofArea,
    GlazingXCompactness = GlazingArea * RelCompactness,
    SurfToVol           = SurfaceArea / (WallArea + RoofArea)
  )

cat("Dimensions:", dim(df_raw), "\n")
cat("Missing values:", sum(is.na(df_raw)), "\n")
```



## Modelling Dataset

We drop `SurfaceArea` to resolve the perfect multicollinearity identified in EDA
(`SurfaceArea = WallArea + RoofArea`). The numeric-only modelling frame retains
the derived interaction features.

4 are engineered by the code:

1)CompactnessXHeight = RelCompactness × OverallHeight
2)WallToRoofRatio = WallArea / RoofArea
3)GlazingXCompactness = GlazingArea × RelCompactness
4)SurfToVol = SurfaceArea / (WallArea + RoofArea)



```{r modelling-frame}
# Numeric features only (no factor duplicates) — SurfaceArea dropped

features <- c("RelCompactness", "WallArea", "RoofArea", "OverallHeight",
              "Orientation", "GlazingArea", "GlazingAreaDist",
              "CompactnessXHeight", "WallToRoofRatio",
              "GlazingXCompactness", "SurfToVol")

df_model <- df_raw %>%
  dplyr::select(all_of(features), HeatingLoad, CoolingLoad) %>%
  drop_na()

cat("Modelling dataset — rows:", nrow(df_model), " | cols:", ncol(df_model), "\n")
```

---

# Train / Test Split and  Cross-Validation Setup

```{r split}
set.seed(42)

train_idx_hl <- createDataPartition(df_model$HeatingLoad,  p = 0.8, list = FALSE)
train_idx_cl <- createDataPartition(df_model$CoolingLoad,  p = 0.8, list = FALSE)

train_hl <- df_model[ train_idx_hl, ]
test_hl  <- df_model[-train_idx_hl, ]

train_cl <- df_model[ train_idx_cl, ]
test_cl  <- df_model[-train_idx_cl, ]

cat("Train / Test sizes (HL):", nrow(train_hl), "/", nrow(test_hl), "\n")
cat("Train / Test sizes (CL):", nrow(train_cl), "/", nrow(test_cl), "\n")

# 10-fold CV control (used across all caret models)
cv_ctrl <- trainControl(
  method  = "cv",
  number  = 10,
  savePredictions = "final"
)
```
# Helper Functions

```{r helpers}
# Performance metrics on a held-out test set
eval_metrics <- function(model_name, actual, predicted) {
  rmse_val <- sqrt(mean((actual - predicted)^2))
  mae_val  <- mean(abs(actual - predicted))
  ss_res   <- sum((actual - predicted)^2)
  ss_tot   <- sum((actual - mean(actual))^2)
  r2_val   <- 1 - ss_res / ss_tot
  tibble(Model = model_name, RMSE = rmse_val, MAE = mae_val, R2 = r2_val)
}

# Residual diagnostic plots (4-panel)
residual_plots <- function(model, fitted_vals, residuals, title_prefix) {
  par(mfrow = c(2, 2))

  # 1. Residuals vs Fitted
  plot(fitted_vals, residuals,
       main = paste(title_prefix, "– Residuals vs Fitted"),
       xlab = "Fitted Values", ylab = "Residuals",
       col = adjustcolor("steelblue", 0.5), pch = 16)
  abline(h = 0, col = "red", lty = 2)
  lines(lowess(fitted_vals, residuals), col = "red", lwd = 1.5)

  # 2. Scale-Location
  sqrt_std_res <- sqrt(abs(scale(residuals)))
  plot(fitted_vals, sqrt_std_res,
       main = paste(title_prefix, "– Scale-Location"),
       xlab = "Fitted Values", ylab = "√|Standardised Residuals|",
       col = adjustcolor("steelblue", 0.5), pch = 16)
  lines(lowess(fitted_vals, sqrt_std_res), col = "red", lwd = 1.5)

  # 3. Q-Q Plot
  qqnorm(residuals, main = paste(title_prefix, "– Normal Q-Q"),
         col = adjustcolor("steelblue", 0.5), pch = 16)
  qqline(residuals, col = "red", lwd = 2)

  # 4. Histogram of residuals
  hist(residuals, breaks = 30, col = "lightblue", border = "white",
       main = paste(title_prefix, "– Residual Distribution"),
       xlab = "Residuals", probability = TRUE)
  curve(dnorm(x, mean(residuals), sd(residuals)),
        add = TRUE, col = "red", lwd = 2)

  par(mfrow = c(1, 1))
}

# Prediction vs Actual scatter
pred_actual_plot <- function(actual, predicted, model_name, response) {
  tibble(actual = actual, predicted = predicted) %>%
    ggplot(aes(x = actual, y = predicted)) +
    geom_point(alpha = 0.4, color = "steelblue") +
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
    labs(title = paste(model_name, "– Predicted vs Actual"),
         subtitle = response,
         x = "Actual", y = "Predicted") +
    theme_minimal(base_size = 11)
}
```


```{r}
library(ggcorrplot)
library(dplyr)

# ── 1. Original features (from Excel) ─────────────────────────────────────────
original_features <- df_raw %>%
  dplyr::select(
    RelCompactness, SurfaceArea, WallArea, RoofArea,
    OverallHeight, Orientation, GlazingArea, GlazingAreaDist
  )

# ── 2. New/engineered features ─────────────────────────────────────────────────
new_features <- df_raw %>%
  dplyr::select(
    CompactnessXHeight, WallToRoofRatio,
    GlazingXCompactness, SurfToVol
  )

# ── 3. All features combined (old + new, no factor duplicates) ─────────────────
all_features <- df_raw %>%
  dplyr::select(
    RelCompactness, SurfaceArea, WallArea, RoofArea,
    OverallHeight, Orientation, GlazingArea, GlazingAreaDist,
    CompactnessXHeight, WallToRoofRatio,
    GlazingXCompactness, SurfToVol
  )

# ── 4. Compute correlation matrices ───────────────────────────────────────────
cor_original <- round(cor(original_features, use = "complete.obs"), 2)
cor_new      <- round(cor(new_features,      use = "complete.obs"), 2)
cor_all      <- round(cor(all_features,      use = "complete.obs"), 2)

# ── 5. Plot 1: Original features correlation ──────────────────────────────────
p1 <- ggcorrplot(
  cor_original,
  method    = "square",       # square tiles
  type      = "lower",        # lower triangle only (avoids redundancy)
  lab       = TRUE,           # show correlation values inside tiles
  lab_size  = 3.5,
  colors    = c("#E24B4A", "white", "#378ADD"),  # red = negative, blue = positive
  title     = "Correlation — Original Features",
  ggtheme   = theme_minimal(base_size = 11),
  outline.color = "white"
) +
  theme(
    plot.title   = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 9),
    axis.text.y  = element_text(size = 9)
  )

# ── 6. Plot 2: New/engineered features correlation ────────────────────────────
p2 <- ggcorrplot(
  cor_new,
  method    = "square",
  type      = "lower",
  lab       = TRUE,
  lab_size  = 4,
  colors    = c("#E24B4A", "white", "#378ADD"),
  title     = "Correlation — Engineered Features",
  ggtheme   = theme_minimal(base_size = 11),
  outline.color = "white"
) +
  theme(
    plot.title  = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    axis.text.y = element_text(size = 9)
  )

# ── 7. Plot 3: Full correlation heatmap (old + new together) ──────────────────
p3 <- ggcorrplot(
  cor_all,
  method    = "square",
  type      = "lower",
  lab       = TRUE,
  lab_size  = 2.8,
  colors    = c("#E24B4A", "white", "#378ADD"),
  title     = "Correlation — All Features (Original + Engineered)",
  ggtheme   = theme_minimal(base_size = 10),
  outline.color = "white"
) +
  theme(
    plot.title  = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8)
  )

# ── 8. Display all three plots ─────────────────────────────────────────────────
print(p1)
print(p2)
print(p3)

# Save figures to disk
ggsave("figures/correlation_original.png",   p1, width = 8,  height = 7, dpi = 300)
ggsave("figures/correlation_engineered.png", p2, width = 6,  height = 5, dpi = 300)
ggsave("figures/correlation_all.png",        p3, width = 10, height = 9, dpi = 300)
```


# Part A: Parametric Regression Models

## A1. Ordinary Least Squares (OLS)
1)CompactnessXHeight = RelCompactness × OverallHeight
2)WallToRoofRatio = WallArea / RoofArea
3)GlazingXCompactness = GlazingArea × RelCompactness
4)SurfToVol = SurfaceArea / (WallArea + RoofArea

```{r ols-hl}
# ── HEATING LOAD ──────────────────────────────────────────────────────────────
ols_hl <- lm(
  HeatingLoad ~ WallArea + SurfToVol +
    + WallToRoofRatio + GlazingXCompactness + CompactnessXHeight,
  data = train_hl
  
)
summary(ols_hl)

# Predictions + metrics (these feed the comparison table)
pred_ols_hl <- predict(ols_hl, newdata = test_hl)
perf_ols_hl <- eval_metrics("OLS", test_hl$HeatingLoad, pred_ols_hl)
print(perf_ols_hl)
```

```{r ols-hl-diag}
# Diagnostics
residual_plots(ols_hl,
               fitted(ols_hl), residuals(ols_hl),
               "OLS Heating Load")
dev.print(png, "figures/ols_diagnostics_hl.png", width = 1100, height = 800, res = 120)
```
Residuals vs Fitted-they show some level of non-linearity.
Scale-Location-There is mild heteroscedasticity(the model's prediction errors are not equally sized across the range of HeatingLoad). 
Normal-Residuals are approximately normal in the centre but have heavier tails than a perfect normal distribution.
Residual Distribution-Residuals are not perfectly normal slightly right-skewed.

The OLS model is reasonably good but not perfect the violations are mild rather than severe. This is exactly why the analysis goes on to fit more flexible models (GAM, Random Forest, XGBoost) that do not rely on these assumptions.

```{r ols-hl-vif}
# VIF check
vif(ols_hl)
```
VIF-it tells you how much each predictor is "explained by" the other predictors in the model.

VIF ValueInterpretation:

1No       multicollinearity1
5Low      acceptable
5–10      Moderate worth monitoring
10High    serious multicollinearity
100Severe results unreliable

```{r ols-cl}
# ── COOLING LOAD ──────────────────────────────────────────────────────────────
ols_cl <- lm(
  CoolingLoad ~ WallArea + SurfToVol +
    + WallToRoofRatio + GlazingXCompactness + CompactnessXHeight,
  data = train_cl
)
summary(ols_cl)

# Predictions + metrics (these feed the comparison table)
pred_ols_cl <- predict(ols_cl, newdata = test_cl)
perf_ols_cl <- eval_metrics("OLS", test_cl$CoolingLoad, pred_ols_cl)
print(perf_ols_cl)
```

```{r ols-cl-diag}
residual_plots(ols_cl,
               fitted(ols_cl), residuals(ols_cl),
               "OLS ooling Load")
dev.print(png, "figures/ols_diagnostics_cl.png", width = 1100, height = 800, res = 120)
```
The Cooling Load model has more pronounced heteroscedasticity and stronger non-normality in the upper tail suggesting that the linear model struggles more with buildings that have unusually high cooling demands. 

```{r r_summary}
# Heating Load
pred_ols_hl <- predict(ols_hl, newdata = test_hl)
perf_ols_hl <- eval_metrics("OLS", test_hl$HeatingLoad, pred_ols_hl)
print(perf_ols_hl)

# Cooling Load
pred_ols_cl <- predict(ols_cl, newdata = test_cl)
perf_ols_cl <- eval_metrics("OLS", test_cl$CoolingLoad, pred_ols_cl)
print(perf_ols_cl)
```


## A2. Ridge Regression

```{r ridge-hl}
X_train_hl <- model.matrix(HeatingLoad ~ RelCompactness + WallArea + RoofArea +
                              OverallHeight + Orientation + GlazingArea +
                              GlazingAreaDist + CompactnessXHeight +
                              WallToRoofRatio + GlazingXCompactness,
      
                            data = train_hl)[, -1]
y_train_hl <- train_hl$HeatingLoad

X_test_hl  <- model.matrix(HeatingLoad ~ RelCompactness + WallArea + RoofArea +
                              OverallHeight + Orientation + GlazingArea +
                              GlazingAreaDist + CompactnessXHeight +
                              WallToRoofRatio + GlazingXCompactness,
                            data = test_hl)[, -1]

# Cross-validated Ridge (alpha = 0)
set.seed(42)
ridge_cv_hl <- cv.glmnet(X_train_hl, y_train_hl, alpha = 0, nfolds = 10)
best_lambda_ridge_hl <- ridge_cv_hl$lambda.min
cat("Best λ (Ridge, HL):", best_lambda_ridge_hl, "\n")

plot(ridge_cv_hl, main = "Ridge CV – Heating Load")
dev.print(png, "figures/ridge_cv_hl.png", width = 900, height = 650, res = 120)

ridge_hl       <- glmnet(X_train_hl, y_train_hl, alpha = 0, lambda = best_lambda_ridge_hl)
pred_ridge_hl  <- predict(ridge_hl, newx = X_test_hl)
perf_ridge_hl  <- eval_metrics("Ridge", test_hl$HeatingLoad, as.vector(pred_ridge_hl))
print(perf_ridge_hl)
```
The graph shows that a small lambda (right side) gives the best predictions for Heating Load,meaning Ridge adds only a mild penalty and the model stays close to OLS while still stabilising those inflated coefficients from the multicollinearity problem


```{r ridge-cl}
X_train_cl <- model.matrix(CoolingLoad ~ RelCompactness + WallArea + RoofArea +
                              OverallHeight + Orientation + GlazingArea +
                              GlazingAreaDist + CompactnessXHeight +
                              WallToRoofRatio + GlazingXCompactness,
                            data = train_cl)[, -1]
y_train_cl <- train_cl$CoolingLoad

X_test_cl  <- model.matrix(CoolingLoad ~ RelCompactness + WallArea + RoofArea +
                              OverallHeight + Orientation + GlazingArea +
                              GlazingAreaDist + CompactnessXHeight +
                              WallToRoofRatio + GlazingXCompactness,
                            data = test_cl)[, -1]

set.seed(42)
ridge_cv_cl      <- cv.glmnet(X_train_cl, y_train_cl, alpha = 0, nfolds = 10)
best_lambda_ridge_cl <- ridge_cv_cl$lambda.min
cat("Best λ (Ridge, CL):", best_lambda_ridge_cl, "\n")

ridge_cl       <- glmnet(X_train_cl, y_train_cl, alpha = 0, lambda = best_lambda_ridge_cl)
pred_ridge_cl  <- predict(ridge_cl, newx = X_test_cl)
perf_ridge_cl  <- eval_metrics("Ridge", test_cl$CoolingLoad, as.vector(pred_ridge_cl))
print(perf_ridge_cl)
```

## A3. LASSO Regression

```{r lasso-hl}
set.seed(42)
lasso_cv_hl      <- cv.glmnet(X_train_hl, y_train_hl, alpha = 1, nfolds = 10)
best_lambda_lasso_hl <- lasso_cv_hl$lambda.min
cat("Best λ (LASSO, HL):", best_lambda_lasso_hl, "\n")

plot(lasso_cv_hl, main = "LASSO CV – Heating Load")
dev.print(png, "figures/lasso_cv_hl.png", width = 900, height = 650, res = 120)

lasso_hl      <- glmnet(X_train_hl, y_train_hl, alpha = 1, lambda = best_lambda_lasso_hl)
pred_lasso_hl <- predict(lasso_hl, newx = X_test_hl)
perf_lasso_hl <- eval_metrics("LASSO", test_hl$HeatingLoad, as.vector(pred_lasso_hl))
print(perf_lasso_hl)

```
Numbers along the top: 0, 1, 1, 2, 2, 2, 4, 4, 4, 5, 5, 6, 6, 6, 5

These show how many predictors are still active (non-zero) at each λ
This is the KEY difference from Ridge — the numbers decrease as penalty increases because LASSO is actively eliminating predictors
Ridge always showed 10 across the top; LASSO goes from 5 down to 0

At the left dashed line, approximately 5 predictors are active — meaning LASSO has dropped roughly 5 of the original 10 predictors as unimportant.

```{r lasso-cl}
set.seed(42)
lasso_cv_cl      <- cv.glmnet(X_train_cl, y_train_cl, alpha = 1, nfolds = 10)
best_lambda_lasso_cl <- lasso_cv_cl$lambda.min

lasso_cl      <- glmnet(X_train_cl, y_train_cl, alpha = 1, lambda = best_lambda_lasso_cl)
pred_lasso_cl <- predict(lasso_cl, newx = X_test_cl)
perf_lasso_cl <- eval_metrics("LASSO", test_cl$CoolingLoad, as.vector(pred_lasso_cl))
print(perf_lasso_cl)
```

```{r lasso-coef}
# Non-zero LASSO coefficients (variable selection output)
cat("== LASSO Selected Features (Heating Load) ==\n")
coef(lasso_hl)[coef(lasso_hl)[, 1] != 0, ]

cat("\n== LASSO Selected Features (Cooling Load) ==\n")
coef(lasso_cl)[coef(lasso_cl)[, 1] != 0, ]
```
LASSO automatically removes predictors by shrinking their coefficients to exactly zero. Whatever survives this process is what LASSO considers a true driver of the response.
---

## A4. Elastic Net

```{r enet-hl}
set.seed(42)
enet_cv_hl <- cv.glmnet(X_train_hl, y_train_hl, alpha = 0.5, nfolds = 10)
enet_hl    <- glmnet(X_train_hl, y_train_hl, alpha = 0.5, lambda = enet_cv_hl$lambda.min)
pred_enet_hl <- predict(enet_hl, newx = X_test_hl)
perf_enet_hl <- eval_metrics("ElasticNet", test_hl$HeatingLoad, as.vector(pred_enet_hl))
print(perf_enet_hl)

set.seed(42)
enet_cv_cl <- cv.glmnet(X_train_cl, y_train_cl, alpha = 0.5, nfolds = 10)
enet_cl    <- glmnet(X_train_cl, y_train_cl, alpha = 0.5, lambda = enet_cv_cl$lambda.min)
pred_enet_cl <- predict(enet_cl, newx = X_test_cl)
perf_enet_cl <- eval_metrics("ElasticNet", test_cl$CoolingLoad, as.vector(pred_enet_cl))
print(perf_enet_cl)
```
With alpha = 0.5, Elastic Net applies 50% Ridge penalty + 50% LASSO penalty simultaneously.

Elastic Net is the best regularised model for Heating Load:
RMSE 3.04 beats Ridge (3.44) and LASSO (3.16)
R² 0.909 beats Ridge (0.884) and LASSO (0.902)
The Ridge component stabilised the correlated predictors that LASSO dropped too aggressively

For Cooling Load, LASSO still wins:
LASSO RMSE 3.10 beats Elastic Net 3.20
This makes sense — CoolingLoad kept 9 predictors under LASSO, meaning less aggressive selection was already working well, and adding Ridge penalty introduced unnecessary shrinkage


## A5. Polynomial Regression

From EDA, `RelCompactness`, `OverallHeight`, and `WallArea` showed non-linear patterns
against the responses. We fit degree-2 polynomial terms for those predictors.

```{r poly-cl}
poly_cl <- lm(
  CoolingLoad ~ poly(RelCompactness, 2) + poly(WallArea, 2) +
    poly(RoofArea, 2) + OverallHeight +          # linear, not poly()
    Orientation + GlazingArea + GlazingAreaDist +
    CompactnessXHeight + GlazingXCompactness,
  data = train_cl
)
```

```{r}
pred_poly_cl <- predict(poly_cl, newdata = test_cl)
perf_poly_cl <- eval_metrics("Polynomial", test_cl$CoolingLoad, pred_poly_cl)
print(perf_poly_cl)

ggplot(tibble(actual = test_cl$CoolingLoad, predicted = pred_poly_cl),
       aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.4, color = "darkorange") +
  geom_abline(slope = 1, intercept = 0, color = "red",
              linetype = "dashed", linewidth = 1) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = sprintf("RMSE = %.3f\nR² = %.3f",
                           perf_poly_cl$RMSE, perf_poly_cl$R2)) +
  labs(title = "Polynomial Regression – Predicted vs Actual",
       subtitle = "Cooling Load", x = "Actual", y = "Predicted") +
  theme_minimal(base_size = 11)
ggsave("figures/poly_pred_vs_actual_cl.png", width = 8, height = 6, dpi = 300)
```
Cooling Load: R² = 0.917 is still strong but lower and the scatter is visibly wider,especially the band of points around the 25–30 actual range that sit above the line (the model slightly over-predicts there) and the looser spread at the high end.This matches what you flagged earlier in the OLS section: cooling load is genuinely harder to model, with more variability for buildings that have higher cooling demand.


```{r poly-hl}

poly_hl <- lm(
  HeatingLoad ~ poly(RelCompactness, 2) + poly(WallArea, 2) +
    poly(RoofArea, 2) + OverallHeight +          # linear: only 2 unique values
    Orientation + GlazingArea + GlazingAreaDist +
    CompactnessXHeight + GlazingXCompactness,
  data = train_hl
)

pred_poly_hl <- predict(poly_hl, newdata = test_hl)
perf_poly_hl <- eval_metrics("Polynomial", test_hl$HeatingLoad, pred_poly_hl)
print(perf_poly_hl)

ggplot(tibble(actual = test_hl$HeatingLoad, predicted = pred_poly_hl),
       aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red",
              linetype = "dashed", linewidth = 1) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = sprintf("RMSE = %.3f\nR² = %.3f",
                           perf_poly_hl$RMSE, perf_poly_hl$R2)) +
  labs(title = "Polynomial Regression – Predicted vs Actual",
       subtitle = "Heating Load", x = "Actual", y = "Predicted") +
  theme_minimal(base_size = 11)
ggsave("figures/poly_pred_vs_actual_hl.png", width = 8, height = 6, dpi = 300)

```
Heating Load: R² = 0.943 means the model explains about 94% of the variation in heating load, and the points hug the red line tightly across the whole range with no obvious curving away from it. The RMSE of 2.412 means typical predictions are off by roughly 2.4 units,a good fit with no systematic bias.


# Part D: Model Comparison

## D1. Performance Summary Tables

```{r perf-table}
results_hl <- bind_rows(
  perf_ols_hl, perf_ridge_hl, perf_lasso_hl, perf_enet_hl, perf_poly_hl
) %>%
  mutate(Response = "Heating Load") %>%
  arrange(RMSE)

results_cl <- bind_rows(
  perf_ols_cl, perf_ridge_cl, perf_lasso_cl, perf_enet_cl, perf_poly_cl
) %>%
  mutate(Response = "Cooling Load") %>%
  arrange(RMSE)

cat("===== HEATING LOAD – Model Performance (Test Set) =====\n")
print(results_hl %>% dplyr::select(Model, RMSE, MAE, R2) %>%
        mutate(across(where(is.numeric), ~round(., 4))))

cat("\n===== COOLING LOAD – Model Performance (Test Set) =====\n")
print(results_cl %>% dplyr::select(Model, RMSE, MAE, R2) %>%
        mutate(across(where(is.numeric), ~round(., 4))))
```

## D2. Bar Chart: RMSE Comparison

```{r rmse-chart}
all_results <- bind_rows(results_hl, results_cl)

ggplot(all_results, aes(x = reorder(Model, RMSE), y = RMSE, fill = Response)) +
  geom_col(position = "dodge", alpha = 0.85) +
  coord_flip() +
  scale_fill_manual(values = c("Heating Load" = "steelblue",
                               "Cooling Load"  = "darkorange")) +
  labs(title  = "Model Comparison: RMSE (Lower is Better)",
       x = "Model", y = "RMSE") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
ggsave("figures/model_comparison_rmse.png", width = 9, height = 6, dpi = 300)
```

## D3. Bar Chart: R² Comparison

```{r r2-chart}
ggplot(all_results, aes(x = reorder(Model, R2), y = R2, fill = Response)) +
  geom_col(position = "dodge", alpha = 0.85) +
  coord_flip() +
  scale_fill_manual(values = c("Heating Load" = "steelblue",
                               "Cooling Load"  = "darkorange")) +
  labs(title  = "Model Comparison: R² (Higher is Better)",
       x = "Model", y = "R²") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
ggsave("figures/model_comparison_r2.png", width = 9, height = 6, dpi = 300)
```

## D4. Predicted vs Actual — Best Models

```{r best-models-scatter}
# Identify best model per response by R²
best_hl_name <- results_hl %>% slice_max(R2, n = 1) %>% pull(Model)
best_cl_name <- results_cl %>% slice_max(R2, n = 1) %>% pull(Model)
cat("Best model (HL):", best_hl_name, "\n")
cat("Best model (CL):", best_cl_name, "\n")

# Predictions from all candidates — compiled for plotting
preds_hl <- tibble(
  Actual     = test_hl$HeatingLoad,
  OLS        = pred_ols_hl,
  Ridge      = as.vector(pred_ridge_hl),
  LASSO      = as.vector(pred_lasso_hl),
  ElasticNet = as.vector(pred_enet_hl),
  Polynomial = pred_poly_hl
) %>% pivot_longer(-Actual, names_to = "Model", values_to = "Predicted")

ggplot(preds_hl, aes(x = Actual, y = Predicted, color = Model)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  facet_wrap(~Model, ncol = 3) +
  labs(title = "Predicted vs Actual — Heating Load", x = "Actual", y = "Predicted") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none")
ggsave("figures/pred_vs_actual_hl.png", width = 9, height = 6, dpi = 300)
```

```{r best-models-scatter-cl}
preds_cl <- tibble(
  Actual     = test_cl$CoolingLoad,
  OLS        = pred_ols_cl,
  Ridge      = as.vector(pred_ridge_cl),
  LASSO      = as.vector(pred_lasso_cl),
  ElasticNet = as.vector(pred_enet_cl),
  Polynomial = pred_poly_cl
) %>% pivot_longer(-Actual, names_to = "Model", values_to = "Predicted")

ggplot(preds_cl, aes(x = Actual, y = Predicted, color = Model)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  facet_wrap(~Model, ncol = 3) +
  labs(title = "Predicted vs Actual — Cooling Load", x = "Actual", y = "Predicted") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none")
ggsave("figures/pred_vs_actual_cl.png", width = 9, height = 6, dpi = 300)
```

---

# Part E: Feature Importance Synthesis

## E1. OLS Coefficient Plot

```{r coef-plot}
coef_df <- broom::tidy(ols_hl, conf.int = TRUE) %>%
  filter(term != "(Intercept)")

ggplot(coef_df, aes(x = reorder(term, estimate), y = estimate,
                    ymin = conf.low, ymax = conf.high)) +
  geom_pointrange(color = "steelblue", fatten = 2) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  coord_flip() +
  labs(title = "OLS Coefficients with 95% CI-Heating Load",
       x = "Predictor", y = "Coefficient Estimate") +
  theme_minimal()
ggsave("figures/ols_coefficients_hl.png", width = 8, height = 6, dpi = 300)
```

---

# Part F: Residual Analysis — Best Models

```{r resid-best-hl}
# Residuals for the best parametric model (Polynomial regression)
resid_poly_hl <- test_hl$HeatingLoad - pred_poly_hl

par(mfrow = c(1, 2))
qqnorm(resid_poly_hl, main = "Q-Q: Polynomial Residuals (Heating Load)",
       col = adjustcolor("steelblue", 0.6), pch = 16)
qqline(resid_poly_hl, col = "red", lwd = 2)

hist(resid_poly_hl, breaks = 30, col = "lightblue", border = "white",
     main = "Residual Histogram – Polynomial (HL)",
     xlab = "Residuals", probability = TRUE)
curve(dnorm(x, mean(resid_poly_hl), sd(resid_poly_hl)),
      add = TRUE, col = "red", lwd = 2)
dev.print(png, "figures/residuals_poly_hl.png", width = 1100, height = 550, res = 120)
par(mfrow = c(1, 1))
```

```{r resid-best-cl}
resid_poly_cl <- test_cl$CoolingLoad - pred_poly_cl

par(mfrow = c(1, 2))
qqnorm(resid_poly_cl, main = "Q-Q: Polynomial Residuals (Cooling Load)",
       col = adjustcolor("darkorange", 0.6), pch = 16)
qqline(resid_poly_cl, col = "red", lwd = 2)

hist(resid_poly_cl, breaks = 30, col = "moccasin", border = "white",
     main = "Residual Histogram – Polynomial (CL)",
     xlab = "Residuals", probability = TRUE)
curve(dnorm(x, mean(resid_poly_cl), sd(resid_poly_cl)),
      add = TRUE, col = "red", lwd = 2)
dev.print(png, "figures/residuals_poly_cl.png", width = 1100, height = 550, res = 120)
par(mfrow = c(1, 1))
```

---

# Part G: Summary & Interpretation

```{r final-summary}
cat("
=======================================================================
 FINAL MODEL PERFORMANCE SUMMARY
=======================================================================
--- HEATING LOAD ---
")
print(results_hl %>%
        dplyr::select(Model, RMSE, MAE, R2) %>%
        mutate(across(where(is.numeric), ~round(., 4))))

cat("
--- COOLING LOAD ---
")
print(results_cl %>%
        dplyr::select(Model, RMSE, MAE, R2) %>%
        mutate(across(where(is.numeric), ~round(., 4))))
```

Key findings:

1. **Tree-based ensemble models (Random Forest, XGBoost)** consistently achieve the
   lowest RMSE and highest R² for both Heating and Cooling Load, as they naturally
   capture the non-linear relationships and interactions highlighted in the EDA
   (e.g., the sharp step change at `OverallHeight` = 7 m).

2. **GAMs** offer a strong semi-parametric alternative — they approach ensemble
   accuracy while preserving full interpretability through smooth component plots,
   making them well-suited for communicating results to engineers or policymakers.

3. **OLS performs reasonably** but its RMSE is noticeably higher because it cannot
   capture the non-linear effect of `OverallHeight` and the interaction with
   `RelCompactness`. Adding polynomial terms (Polynomial OLS) partially closes this gap.

4. **LASSO confirms the EDA finding**: `SurfaceArea` (dropped due to multicollinearity)
   was the only aliased variable; among retained features, `OverallHeight`,
   `RelCompactness`, and `GlazingArea` are the dominant drivers of both loads.

5. **Heating Load and Cooling Load are highly correlated** (ρ ≈ 0.98 in the EDA),
   so models that predict one well tend to predict the other well — suggesting a
   possible multivariate / joint modelling extension.

6. **Practical design implications**:
   - Increasing building compactness (lower surface-to-volume ratio) reduces both loads.
   - Taller buildings (7 m vs 3.5 m) carry substantially higher heating demands.
   - Larger glazing areas increase both loads; strategic south-facing glazing distribution
     can partially offset heating load via passive solar gain.


