# ============================================================
# 10.2  Cubist Regression
# ============================================================

# Define workflow for Cubist model
cubist_model <- cubist::cubist()
cubist_workflow <- workflow() %>%
  add_recipe(reg_recipe) %>%
  add_model(cubist_model)

# Train Cubist model
set.seed(42)
cubist_fit <- cubist_workflow %>% fit(data = train_reg)

# Predict on test set
cubist_pred <- predict(cubist_fit, new_data = test_reg)
cubist_metrics <- metrics(test_reg$HeatingLoad, cubist_pred$.It seems there was an issue with the previous `write_to_file` operation. Let's try appending the Cubist model implementation to the existing content in `classification.Rmd` using a different approach.

### Step 7: Append Cubist Model Implementation

Here is the updated content for `classification.Rmd`:

```{r}
# ============================================================
# 10.2  Cubist Regression
# ============================================================

# Define workflow for Cubist model
cubist_model <- cubist::cubist()
cubist_workflow <- workflow() %>%
  add_recipe(reg_recipe) %>%
  add_model(cubist_model)

# Train Cubist model
set.seed(42)
cubist_fit <- cubist_workflow %>% fit(data = train_reg)

# Predict on test set
cubist_pred <- predict(cubist_fit, new_data = test_reg)
cubist_metrics <- metrics(test_reg$HeatingLoad, cubist_pred$.pred_HeatingLoad) %>%
  bind_cols(metrics(test_reg$CoolingLoad, cubist_pred$.pred_CoolingLoad))

# Print Cubist model metrics
print(cubist_metrics)
```

Now, I will use the `write_to_file` tool to append this content to the existing file.

<write_to_file>
<path>classification.Rmd</path>
<content>
```{r}
# ============================================================
# 10.2  Cubist Regression
# ============================================================

# Define workflow for Cubist model
cubist_model <- cubist::cubist()
cubist_workflow <- workflow() %>%
  add_recipe(reg_recipe) %>%
  add_model(cubist_model)

# Train Cubist model
set.seed(42)
cubist_fit <- cubist_workflow %>% fit(data = train_reg)

# Predict on test set
cubist_pred <- predict(cubist_fit, new_data = test_reg)
cubist_metrics <- metrics(test_reg$HeatingLoad, cubist_pred$.pred_HeatingLoad) %>%
  bind_cols(metrics(test_reg$CoolingLoad, cubist_pred$.pred_CoolingLoad))

# Print Cubist model metrics
print(cubist_metrics)
```
<task_progress>
- [x] Remove classification targets and splitting code
- [x] Create regression targets and split data
- [x] Implement MARS model using `earth` package
- [x] Implement Cubist model using `Cubist` package
- [ ] Implement Random Forest model using `ranger` package
- [ ] Implement XGBoost or LightGBM model using `tidymodels`
- [ ] Evaluate models using RMSE, MAE, R-squared metrics
</task_progress>
</write_to_file>