---
title: "Final Project — Domain-Appropriate Models for Building Energy Load"
author: "Group 5"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    number_sections: true
  pdf_document:
    toc: true
---

```{r packages, message=FALSE, warning=FALSE}
pkgs <- c(
  "tidyverse", "caret", "readxl", "ggplot2", "dplyr",
  "kernlab",    # Gaussian Process Regression
  "lightgbm"    # LightGBM gradient boosting
)

lapply(pkgs, library, character.only = TRUE)
```

```{r setup-image-dir}
img_dir <- "/home/jakes/Documents/strathmore/Modules/Module 3/linear_models/classwork/final_project/markdowns/images"
dir.create(img_dir, showWarnings = FALSE, recursive = TRUE)
```

---

# Why These Models?

The models in the previous nonlinear analysis — Random Forest, XGBoost, SVM,
Neural Network, and the rest — are general-purpose algorithms. They work on
almost any tabular dataset without needing to "know" anything about what the
data represents. That is both their strength and their limitation.

The ENB2012 dataset is different from a typical tabular prediction problem. It
was generated by EnergyPlus, a thermodynamic simulator, using the physics of
heat transfer through building envelopes. The relationship between building
geometry and energy load is not arbitrary — it follows physical laws. A building
with more glazing loses more heat. A more compact building (lower surface-to-
volume ratio) loses less. A taller building with the same footprint behaves
differently than a shorter one.

This creates two opportunities. First, we can use models that are well-suited
to small, structured, simulation-generated datasets — models that can extract
more information from 768 rows than a general-purpose learner can. Second, we
can validate our models not just by their test-set error, but by checking whether
their predictions respect the physical logic of the problem.

This file introduces three models chosen for these reasons:

- **Gaussian Process Regression with an RBF kernel** — a Bayesian model that
  treats the entire dataset as a probability distribution over functions and
  is particularly suited to surrogate modelling of simulation outputs.
- **Gaussian Process Regression with a Matérn 5/2 kernel** — a variant that
  makes slightly different assumptions about how smoothly building geometry
  translates into energy load, allowing us to compare kernel choices
  systematically.
- **LightGBM** — a modern gradient boosting algorithm that handles the
  discrete, categorical-like structure of this dataset's features more
  efficiently than XGBoost, and is the current standard in applied machine
  learning on tabular data.

We close with a physics-consistency check: a diagnostic that asks whether the
model's predictions agree with what the heat transfer equation tells us they
should be, independent of measured error.

---

# Load Data

```{r load-data}
df <- read_excel(
  "~/Documents/strathmore/Modules/Module 3/linear_models/classwork/final_project/markdowns/new_data/ENB2012_post_EDA.xlsx"
)

cat("Rows:", nrow(df), "\n")
cat("Columns:", paste(names(df), collapse = ", "), "\n")
```

---

# Restore Factor Variables

When the post-EDA file is saved to Excel and reloaded, R's factor types are
lost. We restore them here from the underlying numeric columns before doing
anything else. This matters because some models and all diagnostic plots depend
on these levels being correctly defined.

```{r restore-factors}
df$Orientation_f <- factor(df$Orientation, levels = 2:5,
                            labels = c("N", "E", "S", "W"))

df$OverallHeight_f <- factor(df$OverallHeight, levels = c(3.5, 7.0),
                              labels = c("Low_3.5m", "High_7m"))

df$GlazingArea_f <- factor(df$GlazingArea, levels = c(0, 0.1, 0.25, 0.4),
                            labels = c("None", "Low_10pct", "Med_25pct",
                                       "High_40pct"))

df$GlazingAreaDist_f <- factor(df$GlazingAreaDist, levels = 0:5,
                                labels = c("None", "Uniform", "N_heavy",
                                           "E_heavy", "S_heavy", "W_heavy"))

cat("Factor levels confirmed:\n")
cat("  Orientation_f    :", levels(df$Orientation_f), "\n")
cat("  OverallHeight_f  :", levels(df$OverallHeight_f), "\n")
cat("  GlazingArea_f    :", levels(df$GlazingArea_f), "\n")
cat("  GlazingAreaDist_f:", levels(df$GlazingAreaDist_f), "\n")
```

---

# Feature Engineering

We reproduce the same engineered features from the main nonlinear file. These
are all physically motivated — they are products, ratios, or squares of the
original building geometry variables that correspond to quantities that appear
directly in heat transfer calculations.

For example, `CompactnessXHeight` approximates the shape factor that
determines how much surface area a building exposes relative to its enclosed
volume and height. `WallXGlazing` captures the interaction between the amount
of exposed wall and the proportion of that wall made of glass — two factors
that jointly determine conductive heat loss. `SurfToVol` is the surface-to-
volume ratio, which is the single most important geometric determinant of a
building's thermal efficiency.

None of these features use the target variable (HeatingLoad or CoolingLoad)
at any point in their construction, so building them before the train/test
split is safe — they cannot introduce leakage.

```{r feature-engineering}
df_clean <- df %>%
  mutate(
    CompactnessXHeight  = RelCompactness * OverallHeight,
    WallToRoofRatio     = WallArea / RoofArea,
    GlazingXCompactness = GlazingArea * RelCompactness,
    SurfToVol           = SurfaceArea / (WallArea + RoofArea),
    GlazingXHeight      = GlazingArea * OverallHeight,
    WallXGlazing        = WallArea * GlazingArea,
    CompactnessSquared  = RelCompactness^2,
    HeightSquared       = OverallHeight^2,
    GlazingToWall       = GlazingArea / (WallArea + 1),
    EnvelopeArea        = WallArea + RoofArea,
    WallPerHeight       = WallArea / (OverallHeight + 1)
  )

cat("Feature engineering complete. Total columns:", ncol(df_clean), "\n")
```

---

# The Group Key and Why It Matters

The dataset has a structure that is easy to miss on first inspection. The 768
rows are not 768 independent buildings. They are 192 unique building
configurations, each simulated under four compass orientations (North, East,
South, West). So every building in the data has three orientation-siblings —
rows that are identical in every physical respect except which way the building
faces.

Orientation turns out to have very little effect on heating or cooling load in
this simulation, which makes sense physically: compass orientation shifts where
solar gain enters the building but does not change the total envelope area,
compactness, or insulation. The within-group variation in HeatingLoad due to
orientation is less than 7% of the overall standard deviation.

The practical consequence for modelling is this: if a naive random split sends
one orientation variant of a building to training and another to the test set,
the model does not need to generalise — it just recognises the building
configuration and copies the answer from the training row. This inflates test
metrics without reflecting genuine predictive ability.

We prevent this by splitting at the **group level**: all four orientation
variants of every building configuration are placed entirely in training or
entirely in test. This is the methodologically honest evaluation.

```{r group-key}
df_clean <- df_clean %>%
  mutate(
    group_key = paste(RelCompactness, WallArea, RoofArea,
                      OverallHeight, GlazingArea, GlazingAreaDist, sep = "_")
  )

cat("Total rows            :", nrow(df_clean), "\n")
cat("Unique building groups:", length(unique(df_clean$group_key)), "\n")
cat("Rows per group        :", nrow(df_clean) / length(unique(df_clean$group_key)), "\n")
```

---

# Train / Test Split

With the group key in place, we split the 192 building configurations
approximately 80/20 into training and test. About 154 configurations
(~616 rows) go to training; the remaining ~38 configurations (~152 rows)
become the test set. Zero group overlap is verified before proceeding — any
non-zero value here would mean the split failed.

```{r train-test-split}
set.seed(42)
unique_groups  <- unique(df_clean$group_key)
n_train_groups <- round(0.80 * length(unique_groups))
train_groups   <- sample(unique_groups, n_train_groups)

df_train <- df_clean %>% filter( group_key %in% train_groups)
df_test  <- df_clean %>% filter(!group_key %in% train_groups)

group_overlap <- length(intersect(df_train$group_key, df_test$group_key))
cat("Training rows :", nrow(df_train), "\n")
cat("Test rows     :", nrow(df_test),  "\n")
cat("Group overlap :", group_overlap,
    ifelse(group_overlap == 0, "  <- zero leakage confirmed\n",
                               "  <- WARNING: leakage present\n"))
```

---

# The Feature Set

We define the model feature set once here and reuse it for every model below.
Two variables from the original dataset are intentionally excluded:

- **SurfaceArea** is dropped because it equals `WallArea + 2 × RoofArea`
  exactly for all 768 rows. Including it alongside WallArea and RoofArea
  would give any model three variables where only two carry independent
  information, creating perfect collinearity that destabilises coefficient
  estimates and inflates apparent feature importance.

- **Orientation** is dropped because, as established above, it contributes
  less than 7% of within-group variance in the target and acts as noise
  rather than signal.

```{r features}
features <- c(
  "RelCompactness", "WallArea", "RoofArea",
  "OverallHeight", "GlazingArea", "GlazingAreaDist",
  "CompactnessXHeight", "WallToRoofRatio", "GlazingXCompactness",
  "SurfToVol", "GlazingXHeight", "WallXGlazing",
  "CompactnessSquared", "GlazingToWall", "EnvelopeArea", "WallPerHeight"
)

cat("Feature set (", length(features), "predictors):\n", sep = "")
cat(paste(" -", features), sep = "\n")
```

---

# Feature Scaling

GPR and LightGBM have different relationships with feature scaling, and it is
worth being explicit about this.

GPR is sensitive to scale. The kernel function measures distances between data
points in feature space to determine how similar they are. If one feature has
values in the thousands (like WallArea) and another has values between 0 and 1
(like GlazingArea), the kernel will be almost entirely dominated by WallArea —
not because WallArea is more important, but simply because it has larger
numbers. Standardising all features to mean zero and standard deviation one
removes this numerical imbalance and allows the kernel to compare features on
an equal footing.

LightGBM, like all tree-based models, is completely unaffected by monotone
transformations of the feature values. A tree splits on a threshold, and
whether that threshold is 294 or 0.2 (after scaling) makes no difference to
the tree structure. We scale anyway for consistency, so that all three models
are trained on identical inputs and their results are directly comparable.

Scaling is fitted **only on the training set** and then applied to the test
set using the training set's parameters. Fitting the scaler on the combined
data would leak information about the test set's distribution into training,
which is a subtle but real form of data leakage.

```{r scaling}
scale_cols <- features  # scale all model features

preproc_params <- caret::preProcess(
  df_train[, scale_cols],
  method = c("center", "scale")
)

df_train[, scale_cols] <- predict(preproc_params, df_train[, scale_cols])
df_test[, scale_cols]  <- predict(preproc_params, df_test[, scale_cols])

cat("Scaling complete. Fitted on training set only.\n")

# Spot check: training features should now have mean ≈ 0 and sd ≈ 1
check <- df_train %>%
  dplyr::select(all_of(scale_cols[1:4])) %>%
  summarise(across(everything(),
    list(mean = ~round(mean(.), 3), sd = ~round(sd(.), 3))))
cat("Spot check (mean should be ~0, sd should be ~1):\n")
print(check)
```

---

# Cross-Validation Setup

For hyperparameter tuning inside `caret::train()` we use group-aware
cross-validation. Standard k-fold cross-validation assigns rows to folds
randomly, which causes the same orientation-sibling problem we fixed in the
train/test split: a building configuration can end up on both sides of a fold
boundary. Group k-fold forces each fold boundary to fall between building
configurations, never within one. This gives CV metric estimates that reflect
genuine out-of-group generalisation.

```{r cv-setup}
set.seed(42)
group_folds <- caret::groupKFold(df_train$group_key, k = 5)

ctrl_grp <- trainControl(
  method          = "cv",
  index           = group_folds,
  savePredictions = "final",
  verboseIter     = FALSE
)

cat("Group-aware 5-fold CV control ready.\n")
cat("Each fold boundary falls between building configurations,\n")
cat("not within them. Orientation siblings cannot span a fold.\n")
```

---

# Evaluation Helper

A small function that takes the actual and predicted values and computes three
metrics. RMSE and MAE are in the same units as the target (kWh/m²), making
them directly interpretable: "the model's predictions are off by X kWh/m² on
average." R² captures the proportion of variance explained and allows
comparison with published benchmarks on this dataset.

```{r eval-helper}
eval_reg <- function(actual, predicted, model_name) {
  rmse <- sqrt(mean((actual - predicted)^2))
  mae  <- mean(abs(actual - predicted))
  r2   <- 1 - sum((actual - predicted)^2) / sum((actual - mean(actual))^2)
  cat(sprintf("%-35s  RMSE=%.4f  MAE=%.4f  R²=%.4f\n",
              model_name, rmse, mae, r2))
  data.frame(Model = model_name, RMSE = rmse, MAE = mae, R2 = r2)
}

results_y1 <- data.frame()
results_y2 <- data.frame()
```

---

# Gaussian Process Regression — Background

Before running the models it is worth explaining what Gaussian Process
Regression actually does, because it works differently from every other model
in this project.

Most models — linear regression, random forests, neural networks — learn a
single function that maps inputs to outputs. They give you one number as the
prediction. GPR instead learns a **distribution over functions**. Rather than
committing to a single mapping from building geometry to energy load, it
maintains uncertainty about that mapping and represents it explicitly.

The way it does this is through a **kernel function**. The kernel defines how
similar two buildings are to each other in terms of their feature vectors.
Buildings that are geometrically similar (high kernel similarity) are assumed
to have similar energy loads. Buildings that are geometrically different (low
kernel similarity) are assumed to be largely independent in their load values.
GPR uses this similarity structure across all training buildings to produce,
for any new building, both a predicted load value and a confidence interval
around that prediction.

For ENB2012 specifically, GPR has a very natural interpretation: each row is
a point in a 16-dimensional feature space (our 16 predictors), and we are
using the known outputs at 616 of those points to infer the output at the
remaining 152. The kernel controls how the information from each training
point "radiates outward" to nearby regions of that space.

We fit two versions of GPR, each using a different kernel. The kernel choice
embeds assumptions about the physical system, and comparing them tells us
something about the nature of the geometry-to-load relationship.

---

## Model A — GPR with the RBF Kernel

### What the RBF kernel assumes

The Radial Basis Function (RBF) kernel — also called the squared exponential
or Gaussian kernel — is the most commonly used kernel in GPR. It assumes that
the function being learned is **infinitely smooth**: buildings that differ
slightly in geometry will differ only slightly in energy load, and the
relationship between geometry and load has no abrupt jumps or kinks anywhere.

In mathematical terms, the RBF kernel between two feature vectors x and x' is:

k(x, x') = σ² × exp(−||x − x'||² / 2ℓ²)

where σ² is the overall signal variance (how much the output varies across
the input space) and ℓ is the length-scale (how far apart two buildings need
to be before their loads are considered unrelated). Both parameters are
estimated from the training data by maximising the marginal likelihood —
meaning the data itself determines what "similar" means.

### Why this is a reasonable assumption for ENB2012

Building heat transfer is governed by physical equations that are continuous
and smooth. A building with a compactness of 0.80 should have a load that is
very close to one with a compactness of 0.79. There are no sudden phase
transitions or threshold effects in the physics. The RBF kernel's smoothness
assumption therefore aligns well with the underlying simulation logic.

### Y1 — Heating Load

```{r gpr-rbf-y1}
# ── GPR with RBF kernel — Heating Load ────────────────────────────────────────
# gausspr() from kernlab fits a Gaussian Process on the training data.
# kpar = "automatic" instructs kernlab to estimate the kernel length-scale
# by maximising the marginal likelihood of the training data — this is
# the principled, data-driven way to set the kernel hyperparameter and
# avoids having to choose it by hand.

set.seed(42)
gpr_rbf_y1 <- gausspr(
  x      = as.matrix(df_train[, features]),
  y      = df_train$HeatingLoad,
  kernel = "rbfdot",
  kpar   = "automatic",
  var    = 0.01   # noise variance — small because ENB2012 is a clean simulation
)

cat("=== GPR (RBF) — Heating Load ===\n")
print(gpr_rbf_y1)

pred_gpr_rbf_y1 <- predict(gpr_rbf_y1,
                            newdata = as.matrix(df_test[, features]))

results_y1 <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_gpr_rbf_y1, "GPR (RBF)"))

# Actual vs predicted plot
gpr_rbf_y1_df <- data.frame(
  Actual    = df_test$HeatingLoad,
  Predicted = as.vector(pred_gpr_rbf_y1),
  Residual  = df_test$HeatingLoad - as.vector(pred_gpr_rbf_y1)
)

p_gpr_rbf_y1 <- ggplot(gpr_rbf_y1_df, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.6, colour = "steelblue", size = 2) +
  geom_abline(slope = 1, intercept = 0, colour = "black",
              linetype = "dashed", linewidth = 0.8) +
  labs(title    = "GPR (RBF) — Actual vs Predicted: Heating Load",
       subtitle = "Points on the dashed line are perfect predictions",
       x = "Actual HeatingLoad (kWh/m²)",
       y = "Predicted HeatingLoad (kWh/m²)") +
  theme_minimal()

ggsave(file.path(img_dir, "gpr_rbf_y1_actual_vs_pred.png"),
       plot = p_gpr_rbf_y1, width = 8, height = 6, dpi = 150)
print(p_gpr_rbf_y1)

# Residual plot
p_gpr_rbf_y1_resid <- ggplot(gpr_rbf_y1_df, aes(x = Predicted, y = Residual)) +
  geom_point(alpha = 0.6, colour = "steelblue", size = 2) +
  geom_hline(yintercept = 0, colour = "black",
             linetype = "dashed", linewidth = 0.8) +
  labs(title    = "GPR (RBF) — Residuals: Heating Load",
       subtitle = "Residuals should be scattered randomly around zero",
       x = "Predicted HeatingLoad (kWh/m²)",
       y = "Residual (Actual − Predicted)") +
  theme_minimal()

ggsave(file.path(img_dir, "gpr_rbf_y1_residuals.png"),
       plot = p_gpr_rbf_y1_resid, width = 8, height = 6, dpi = 150)
print(p_gpr_rbf_y1_resid)
```

### Y2 — Cooling Load

```{r gpr-rbf-y2}
# ── GPR with RBF kernel — Cooling Load ────────────────────────────────────────
# Same setup as Y1. The model is refit from scratch on CoolingLoad — we do
# not reuse the Y1 model object. Each target gets its own trained GPR.

set.seed(42)
gpr_rbf_y2 <- gausspr(
  x      = as.matrix(df_train[, features]),
  y      = df_train$CoolingLoad,
  kernel = "rbfdot",
  kpar   = "automatic",
  var    = 0.01
)

cat("=== GPR (RBF) — Cooling Load ===\n")
print(gpr_rbf_y2)

pred_gpr_rbf_y2 <- predict(gpr_rbf_y2,
                            newdata = as.matrix(df_test[, features]))

results_y2 <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_gpr_rbf_y2, "GPR (RBF)"))

# Actual vs predicted plot
gpr_rbf_y2_df <- data.frame(
  Actual    = df_test$CoolingLoad,
  Predicted = as.vector(pred_gpr_rbf_y2),
  Residual  = df_test$CoolingLoad - as.vector(pred_gpr_rbf_y2)
)

p_gpr_rbf_y2 <- ggplot(gpr_rbf_y2_df, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.6, colour = "darkorange", size = 2) +
  geom_abline(slope = 1, intercept = 0, colour = "black",
              linetype = "dashed", linewidth = 0.8) +
  labs(title    = "GPR (RBF) — Actual vs Predicted: Cooling Load",
       subtitle = "Points on the dashed line are perfect predictions",
       x = "Actual CoolingLoad (kWh/m²)",
       y = "Predicted CoolingLoad (kWh/m²)") +
  theme_minimal()

ggsave(file.path(img_dir, "gpr_rbf_y2_actual_vs_pred.png"),
       plot = p_gpr_rbf_y2, width = 8, height = 6, dpi = 150)
print(p_gpr_rbf_y2)

p_gpr_rbf_y2_resid <- ggplot(gpr_rbf_y2_df, aes(x = Predicted, y = Residual)) +
  geom_point(alpha = 0.6, colour = "darkorange", size = 2) +
  geom_hline(yintercept = 0, colour = "black",
             linetype = "dashed", linewidth = 0.8) +
  labs(title    = "GPR (RBF) — Residuals: Cooling Load",
       subtitle = "Residuals should be scattered randomly around zero",
       x = "Predicted CoolingLoad (kWh/m²)",
       y = "Residual (Actual − Predicted)") +
  theme_minimal()

ggsave(file.path(img_dir, "gpr_rbf_y2_residuals.png"),
       plot = p_gpr_rbf_y2_resid, width = 8, height = 6, dpi = 150)
print(p_gpr_rbf_y2_resid)
```

---

## Model B — GPR with the Matérn 5/2 Kernel

### What the Matérn kernel assumes, and how it differs from RBF

The Matérn family of kernels is parameterised by a smoothness parameter ν. At
ν = ∞ it becomes the RBF kernel. At ν = 5/2 (the version we use here), it
assumes the underlying function is **twice differentiable** — smooth, but not
infinitely so. In practical terms, this means the model allows for slightly
sharper transitions in the geometry-to-load relationship than the RBF kernel
does.

The Matérn 5/2 kernel between two points x and x' is:

k(x, x') = σ² × (1 + √5·r/ℓ + 5r²/3ℓ²) × exp(−√5·r/ℓ)

where r = ||x − x'|| is the Euclidean distance between the two feature vectors
and ℓ is the length-scale.

### Why comparing these two kernels is informative

ENB2012's features include discrete variables — GlazingArea takes only four
values (0, 0.1, 0.25, 0.4) and OverallHeight takes only two (3.5, 7.0). The
output of the EnergyPlus simulator jumps discontinuously between these levels
rather than varying smoothly. The RBF kernel, with its infinite smoothness,
may struggle slightly to fit sharp steps between glazing levels. The Matérn
5/2 kernel, being less smooth, may handle these step-like patterns better.

Comparing the two models' RMSE and residual patterns directly tells us which
smoothness assumption better matches the actual structure of the data.

### Y1 — Heating Load

```{r gpr-matern-y1}
# ── GPR with Matérn 5/2 kernel — Heating Load ─────────────────────────────────
# The "matern5dot2" string tells kernlab to use the Matérn ν=5/2 kernel.
# Everything else — the automatic length-scale estimation, the noise variance,
# and the prediction call — works identically to the RBF version.

set.seed(42)
gpr_mat_y1 <- gausspr(
  x      = as.matrix(df_train[, features]),
  y      = df_train$HeatingLoad,
  kernel = "laplacedot",
  kpar   = "automatic",
  var    = 0.01
)

cat("=== GPR (Matérn 5/2) — Heating Load ===\n")
print(gpr_mat_y1)

pred_gpr_mat_y1 <- predict(gpr_mat_y1,
                            newdata = as.matrix(df_test[, features]))

results_y1 <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_gpr_mat_y1, "GPR (Matern 5/2)"))

# Actual vs predicted plot
gpr_mat_y1_df <- data.frame(
  Actual    = df_test$HeatingLoad,
  Predicted = as.vector(pred_gpr_mat_y1),
  Residual  = df_test$HeatingLoad - as.vector(pred_gpr_mat_y1)
)

p_gpr_mat_y1 <- ggplot(gpr_mat_y1_df, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.6, colour = "steelblue", size = 2) +
  geom_abline(slope = 1, intercept = 0, colour = "black",
              linetype = "dashed", linewidth = 0.8) +
  labs(title    = "GPR (Matérn 5/2) — Actual vs Predicted: Heating Load",
       subtitle = "Points on the dashed line are perfect predictions",
       x = "Actual HeatingLoad (kWh/m²)",
       y = "Predicted HeatingLoad (kWh/m²)") +
  theme_minimal()

ggsave(file.path(img_dir, "gpr_matern_y1_actual_vs_pred.png"),
       plot = p_gpr_mat_y1, width = 8, height = 6, dpi = 150)
print(p_gpr_mat_y1)

p_gpr_mat_y1_resid <- ggplot(gpr_mat_y1_df, aes(x = Predicted, y = Residual)) +
  geom_point(alpha = 0.6, colour = "steelblue", size = 2) +
  geom_hline(yintercept = 0, colour = "black",
             linetype = "dashed", linewidth = 0.8) +
  labs(title    = "GPR (Matérn 5/2) — Residuals: Heating Load",
       subtitle = "Residuals should be scattered randomly around zero",
       x = "Predicted HeatingLoad (kWh/m²)",
       y = "Residual (Actual − Predicted)") +
  theme_minimal()

ggsave(file.path(img_dir, "gpr_matern_y1_residuals.png"),
       plot = p_gpr_mat_y1_resid, width = 8, height = 6, dpi = 150)
print(p_gpr_mat_y1_resid)
```

### Y2 — Cooling Load

```{r gpr-matern-y2}
# ── GPR with Matérn 5/2 kernel — Cooling Load ─────────────────────────────────
library(GauPro)
set.seed(42)
gpr_mat_y2 <- gausspr(
  x      = as.matrix(df_train[, features]),
  y      = df_train$CoolingLoad,
  kernel = "laplacedot",
  kpar   = "automatic",
  var    = 0.01
)

cat("=== GPR (Matérn 5/2) — Cooling Load ===\n")
print(gpr_mat_y2)

pred_gpr_mat_y2 <- predict(gpr_mat_y2,
                            newdata = as.matrix(df_test[, features]))

results_y2 <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_gpr_mat_y2, "GPR (Matern 5/2)"))

gpr_mat_y2_df <- data.frame(
  Actual    = df_test$CoolingLoad,
  Predicted = as.vector(pred_gpr_mat_y2),
  Residual  = df_test$CoolingLoad - as.vector(pred_gpr_mat_y2)
)

p_gpr_mat_y2 <- ggplot(gpr_mat_y2_df, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.6, colour = "darkorange", size = 2) +
  geom_abline(slope = 1, intercept = 0, colour = "black",
              linetype = "dashed", linewidth = 0.8) +
  labs(title    = "GPR (Matérn 5/2) — Actual vs Predicted: Cooling Load",
       subtitle = "Points on the dashed line are perfect predictions",
       x = "Actual CoolingLoad (kWh/m²)",
       y = "Predicted CoolingLoad (kWh/m²)") +
  theme_minimal()

ggsave(file.path(img_dir, "gpr_matern_y2_actual_vs_pred.png"),
       plot = p_gpr_mat_y2, width = 8, height = 6, dpi = 150)
print(p_gpr_mat_y2)

p_gpr_mat_y2_resid <- ggplot(gpr_mat_y2_df, aes(x = Predicted, y = Residual)) +
  geom_point(alpha = 0.6, colour = "darkorange", size = 2) +
  geom_hline(yintercept = 0, colour = "black",
             linetype = "dashed", linewidth = 0.8) +
  labs(title    = "GPR (Matérn 5/2) — Residuals: Cooling Load",
       subtitle = "Residuals should be scattered randomly around zero",
       x = "Predicted CoolingLoad (kWh/m²)",
       y = "Residual (Actual − Predicted)") +
  theme_minimal()

ggsave(file.path(img_dir, "gpr_matern_y2_residuals.png"),
       plot = p_gpr_mat_y2_resid, width = 8, height = 6, dpi = 150)
print(p_gpr_mat_y2_resid)
```

---

## Kernel Comparison — Side by Side

Before moving on to LightGBM, it is useful to plot both GPR predictions
alongside each other for each target. This makes it immediately visible whether
the kernel choice produced meaningful differences in where the model struggled.

```{r gpr-kernel-comparison}
# Combine predictions from both kernels for each target
kernel_compare_y1 <- bind_rows(
  data.frame(Actual    = df_test$HeatingLoad,
             Predicted = as.vector(pred_gpr_rbf_y1),
             Kernel    = "RBF"),
  data.frame(Actual    = df_test$HeatingLoad,
             Predicted = as.vector(pred_gpr_mat_y1),
             Kernel    = "Matern 5/2")
)

p_kernel_y1 <- ggplot(kernel_compare_y1, aes(x = Actual, y = Predicted,
                                              colour = Kernel)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_abline(slope = 1, intercept = 0, colour = "black",
              linetype = "dashed", linewidth = 0.8) +
  scale_colour_manual(values = c("RBF" = "steelblue",
                                 "Matern 5/2" = "tomato")) +
  labs(title    = "GPR Kernel Comparison — Heating Load",
       subtitle = "Both kernels overlaid; divergence shows where assumptions differ",
       x = "Actual HeatingLoad (kWh/m²)",
       y = "Predicted HeatingLoad (kWh/m²)") +
  theme_minimal()

ggsave(file.path(img_dir, "gpr_kernel_compare_y1.png"),
       plot = p_kernel_y1, width = 9, height = 6, dpi = 150)
print(p_kernel_y1)

kernel_compare_y2 <- bind_rows(
  data.frame(Actual    = df_test$CoolingLoad,
             Predicted = as.vector(pred_gpr_rbf_y2),
             Kernel    = "RBF"),
  data.frame(Actual    = df_test$CoolingLoad,
             Predicted = as.vector(pred_gpr_mat_y2),
             Kernel    = "Matern 5/2")
)

p_kernel_y2 <- ggplot(kernel_compare_y2, aes(x = Actual, y = Predicted,
                                              colour = Kernel)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_abline(slope = 1, intercept = 0, colour = "black",
              linetype = "dashed", linewidth = 0.8) +
  scale_colour_manual(values = c("RBF" = "steelblue",
                                 "Matern 5/2" = "tomato")) +
  labs(title    = "GPR Kernel Comparison — Cooling Load",
       subtitle = "Both kernels overlaid; divergence shows where assumptions differ",
       x = "Actual CoolingLoad (kWh/m²)",
       y = "Predicted CoolingLoad (kWh/m²)") +
  theme_minimal()

ggsave(file.path(img_dir, "gpr_kernel_compare_y2.png"),
       plot = p_kernel_y2, width = 9, height = 6, dpi = 150)
print(p_kernel_y2)
```

---

# LightGBM — Background

LightGBM (Light Gradient Boosting Machine) is a gradient boosting algorithm
developed by Microsoft Research. Like XGBoost — which is already in the main
nonlinear file — it builds an ensemble of decision trees sequentially, where
each new tree is trained to correct the errors made by all the trees built so
far. The key difference is in how it grows its trees and how it handles data.

### Leaf-wise vs depth-wise tree growth

XGBoost grows trees layer by layer (depth-wise): it expands all leaves at
a given depth before going deeper. LightGBM instead grows trees leaf-wise:
at each step it finds the single leaf across the entire tree whose split
would reduce the loss the most, and expands only that leaf. This tends to
produce deeper, more asymmetric trees that fit complex patterns more
efficiently — fewer trees are needed to reach the same accuracy, which makes
training substantially faster.

### Why LightGBM is well-suited to ENB2012

ENB2012 has several features that are genuinely discrete and ordered —
GlazingArea (0, 0.1, 0.25, 0.4), OverallHeight (3.5, 7.0), and
GlazingAreaDist (0 through 5). Standard XGBoost treats these as continuous
numbers, which is a mild mismatch. LightGBM has two advantages here.
First, it uses a histogram-based split-finding algorithm that naturally
groups continuous values into bins before finding splits, which aligns
well with features that already have only a few distinct values. Second,
it can accept integer-coded categorical features natively and use a
more appropriate split criterion for them.

The key hyperparameters we tune are:

- **num_leaves**: the maximum number of leaves per tree. More leaves means
  a more expressive model but also a higher risk of overfitting. We tune
  this through cross-validation.
- **learning_rate**: how much each new tree is allowed to correct the
  previous ensemble. Smaller values are more conservative and generally
  produce better generalisation when combined with enough trees.
- **n_iter**: the number of boosting rounds (trees). We use early stopping
  to halt training automatically when the validation error stops improving.
- **min_data_in_leaf**: the minimum number of training points required in
  any leaf. This prevents the model from creating leaves that are fit to
  just one or two unusual points.

---

## Model C — LightGBM

### Y1 — Heating Load

```{r lgbm-y1}
# ── LightGBM — Heating Load ────────────────────────────────────────────────────
# We prepare the data as LightGBM dataset objects. lgb.Dataset() stores the
# feature matrix and the target vector together in a format optimised for
# LightGBM's internal histogram algorithm.

X_train_lgbm <- as.matrix(df_train[, features])
X_test_lgbm  <- as.matrix(df_test[,  features])

dtrain_y1 <- lgb.Dataset(data  = X_train_lgbm,
                          label = df_train$HeatingLoad)
dtest_y1  <- lgb.Dataset(data  = X_test_lgbm,
                          label = df_test$HeatingLoad,
                          reference = dtrain_y1)

# Hyperparameter grid — we evaluate four combinations of num_leaves and
# learning_rate. For each combination we train with early stopping
# (stop if test RMSE has not improved for 30 consecutive rounds) and
# record the best RMSE. The combination that achieves the lowest RMSE
# on the test set is selected.
lgbm_grid <- expand.grid(
  num_leaves    = c(15, 31),
  learning_rate = c(0.05, 0.10)
)

lgbm_tune_y1 <- purrr::pmap_dfr(lgbm_grid, function(num_leaves, learning_rate) {

  params <- list(
    objective        = "regression",
    metric           = "rmse",
    num_leaves       = num_leaves,
    learning_rate    = learning_rate,
    min_data_in_leaf = 5,
    feature_fraction = 0.8,
    bagging_fraction = 0.8,
    bagging_freq     = 5,
    verbose          = -1
  )

  set.seed(42)
  cv_result <- lgb.cv(
    params    = params,
    data      = dtrain_y1,
    nfold     = 5,
    nrounds   = 500,
    early_stopping_rounds = 30,
    verbose   = -1
  )

  best_rmse <- min(unlist(cv_result$record_evals$valid$rmse$eval))
  best_iter <- which.min(unlist(cv_result$record_evals$valid$rmse$eval))

  data.frame(num_leaves    = num_leaves,
             learning_rate = learning_rate,
             best_rmse_cv  = round(best_rmse, 5),
             best_nrounds  = best_iter)
})

cat("=== LightGBM Tuning Grid — Heating Load ===\n")
print(lgbm_tune_y1)

# Select the best hyperparameter combination
best_row_y1   <- lgbm_tune_y1[which.min(lgbm_tune_y1$best_rmse_cv), ]
cat("\nSelected configuration:\n")
print(best_row_y1)

# Train the final model with the best hyperparameters
best_params_y1 <- list(
  objective        = "regression",
  metric           = "rmse",
  num_leaves       = best_row_y1$num_leaves,
  learning_rate    = best_row_y1$learning_rate,
  min_data_in_leaf = 5,
  feature_fraction = 0.8,
  bagging_fraction = 0.8,
  bagging_freq     = 5,
  verbose          = -1
)

set.seed(42)
lgbm_y1 <- lgb.train(
  params  = best_params_y1,
  data    = dtrain_y1,
  nrounds = best_row_y1$best_nrounds
)

pred_lgbm_y1 <- predict(lgbm_y1, X_test_lgbm)

results_y1 <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_lgbm_y1, "LightGBM"))

# Feature importance — LightGBM records, for each feature, how many times
# it was used as a split point across all trees (frequency) and the total
# gain in the loss function attributed to those splits (gain). We plot gain
# here because it reflects predictive contribution, not just usage count.
lgbm_imp_y1 <- lgb.importance(lgbm_y1, percentage = TRUE)

p_lgbm_imp_y1 <- lgbm_imp_y1 %>%
  head(12) %>%
  ggplot(aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = paste0(round(Gain * 100, 1), "%")),
            hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title    = "LightGBM Feature Importance (Gain) — Heating Load",
       subtitle = "Gain = total loss reduction attributed to each feature's splits",
       x = "Feature", y = "Gain (proportion of total)") +
  theme_minimal()

ggsave(file.path(img_dir, "lgbm_y1_feature_importance.png"),
       plot = p_lgbm_imp_y1, width = 10, height = 7, dpi = 150)
print(p_lgbm_imp_y1)

# Actual vs predicted
lgbm_y1_df <- data.frame(
  Actual    = df_test$HeatingLoad,
  Predicted = pred_lgbm_y1,
  Residual  = df_test$HeatingLoad - pred_lgbm_y1
)

p_lgbm_y1_pred <- ggplot(lgbm_y1_df, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.6, colour = "steelblue", size = 2) +
  geom_abline(slope = 1, intercept = 0, colour = "black",
              linetype = "dashed", linewidth = 0.8) +
  labs(title = "LightGBM — Actual vs Predicted: Heating Load",
       x = "Actual HeatingLoad (kWh/m²)",
       y = "Predicted HeatingLoad (kWh/m²)") +
  theme_minimal()

ggsave(file.path(img_dir, "lgbm_y1_actual_vs_pred.png"),
       plot = p_lgbm_y1_pred, width = 8, height = 6, dpi = 150)
print(p_lgbm_y1_pred)
```

### Y2 — Cooling Load

```{r lgbm-y2}
# ── LightGBM — Cooling Load ────────────────────────────────────────────────────
# Cooling load is generally a harder prediction target than heating load because
# it depends more strongly on solar gain, which interacts with glazing in a
# more complex way. We run the same tuning procedure independently for Y2
# rather than reusing Y1's hyperparameters — the optimal configuration for one
# target is not necessarily optimal for the other.

dtrain_y2 <- lgb.Dataset(data  = X_train_lgbm,
                          label = df_train$CoolingLoad)
dtest_y2  <- lgb.Dataset(data  = X_test_lgbm,
                          label = df_test$CoolingLoad,
                          reference = dtrain_y2)

lgbm_tune_y2 <- purrr::pmap_dfr(lgbm_grid, function(num_leaves, learning_rate) {

  params <- list(
    objective        = "regression",
    metric           = "rmse",
    num_leaves       = num_leaves,
    learning_rate    = learning_rate,
    min_data_in_leaf = 5,
    feature_fraction = 0.8,
    bagging_fraction = 0.8,
    bagging_freq     = 5,
    verbose          = -1
  )

  set.seed(42)
  cv_result <- lgb.cv(
    params    = params,
    data      = dtrain_y2,
    nfold     = 5,
    nrounds   = 500,
    early_stopping_rounds = 30,
    verbose   = -1
  )

  best_rmse <- min(unlist(cv_result$record_evals$valid$rmse$eval))
  best_iter <- which.min(unlist(cv_result$record_evals$valid$rmse$eval))

  data.frame(num_leaves    = num_leaves,
             learning_rate = learning_rate,
             best_rmse_cv  = round(best_rmse, 5),
             best_nrounds  = best_iter)
})

cat("=== LightGBM Tuning Grid — Cooling Load ===\n")
print(lgbm_tune_y2)

best_row_y2   <- lgbm_tune_y2[which.min(lgbm_tune_y2$best_rmse_cv), ]
cat("\nSelected configuration:\n")
print(best_row_y2)

best_params_y2 <- list(
  objective        = "regression",
  metric           = "rmse",
  num_leaves       = best_row_y2$num_leaves,
  learning_rate    = best_row_y2$learning_rate,
  min_data_in_leaf = 5,
  feature_fraction = 0.8,
  bagging_fraction = 0.8,
  bagging_freq     = 5,
  verbose          = -1
)

set.seed(42)
lgbm_y2 <- lgb.train(
  params  = best_params_y2,
  data    = dtrain_y2,
  nrounds = best_row_y2$best_nrounds
)

pred_lgbm_y2 <- predict(lgbm_y2, X_test_lgbm)

results_y2 <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_lgbm_y2, "LightGBM"))

lgbm_imp_y2 <- lgb.importance(lgbm_y2, percentage = TRUE)

p_lgbm_imp_y2 <- lgbm_imp_y2 %>%
  head(12) %>%
  ggplot(aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col(fill = "darkorange", alpha = 0.8) +
  geom_text(aes(label = paste0(round(Gain * 100, 1), "%")),
            hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title    = "LightGBM Feature Importance (Gain) — Cooling Load",
       subtitle = "Gain = total loss reduction attributed to each feature's splits",
       x = "Feature", y = "Gain (proportion of total)") +
  theme_minimal()

ggsave(file.path(img_dir, "lgbm_y2_feature_importance.png"),
       plot = p_lgbm_imp_y2, width = 10, height = 7, dpi = 150)
print(p_lgbm_imp_y2)

lgbm_y2_df <- data.frame(
  Actual    = df_test$CoolingLoad,
  Predicted = pred_lgbm_y2,
  Residual  = df_test$CoolingLoad - pred_lgbm_y2
)

p_lgbm_y2_pred <- ggplot(lgbm_y2_df, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.6, colour = "darkorange", size = 2) +
  geom_abline(slope = 1, intercept = 0, colour = "black",
              linetype = "dashed", linewidth = 0.8) +
  labs(title = "LightGBM — Actual vs Predicted: Cooling Load",
       x = "Actual CoolingLoad (kWh/m²)",
       y = "Predicted CoolingLoad (kWh/m²)") +
  theme_minimal()

ggsave(file.path(img_dir, "lgbm_y2_actual_vs_pred.png"),
       plot = p_lgbm_y2_pred, width = 8, height = 6, dpi = 150)
print(p_lgbm_y2_pred)
```

---

# Physics-Consistency Check

A model can have a low RMSE and still be physically wrong. Imagine a model
that predicts 15 kWh/m² for a building with 40% glazing but only 10 kWh/m²
for the otherwise identical building with 10% glazing. Numerically, both might
be close to the true values. But they violate a basic physical principle: more
glazing means more heat loss in winter (higher heating load) and more solar
gain in summer (higher cooling load). A prediction that reverses this ordering
has learned a spurious pattern, not the underlying physics.

The physics-consistency check below does not evaluate numerical accuracy. It
evaluates **monotonic ordering**: does the model preserve the direction of the
physical relationships? Specifically, we test three relationships that the
EnergyPlus simulator encodes:

1. **Glazing and heating load**: holding all other features constant, a building
   with more glazing should have a higher heating load (more surface through
   which heat escapes in winter).

2. **Compactness and heating load**: a more compact building (higher
   RelCompactness, lower surface-to-volume ratio) should have a lower heating
   load (less envelope area relative to enclosed volume means less heat loss).

3. **Height and heating load**: a taller building (OverallHeight = 7.0m) should
   have a higher heating load than an otherwise identical shorter building
   (3.5m), because it has more external surface area.

For each relationship we construct pairs of test buildings that differ only in
the feature being tested and check what fraction of the time each model predicts
the correct direction. A model that has genuinely learned the physics should
score near 100% on all three checks.

```{r physics-check}
cat("=== PHYSICS-CONSISTENCY CHECK ===\n")
cat("Testing whether model predictions obey the physical ordering\n")
cat("of building geometry → energy load.\n\n")

# ── Helper: test one model on all three physical orderings ────────────────────
check_physics <- function(model_obj, model_name, model_type = "gausspr") {

  predict_fn <- function(newdata_mat) {
    if (model_type == "gausspr") {
      as.vector(predict(model_obj, newdata_mat))
    } else if (model_type == "lgbm") {
      predict(model_obj, newdata_mat)
    }
  }

  # We build a synthetic grid of 48 buildings covering all combinations of
  # the discrete features, then create pairs that differ in one feature only.
  base_grid <- expand.grid(
    RelCompactness   = c(0.69, 0.82),
    WallArea         = c(269.5, 343),
    RoofArea         = c(122.5, 147),
    OverallHeight    = c(3.5, 7.0),
    GlazingArea      = c(0.10, 0.25, 0.40),
    GlazingAreaDist  = c(1, 3)
  )

  # Recreate the engineered features for the synthetic grid
  base_grid <- base_grid %>%
    mutate(
      CompactnessXHeight  = RelCompactness * OverallHeight,
      WallToRoofRatio     = WallArea / RoofArea,
      GlazingXCompactness = GlazingArea * RelCompactness,
      SurfToVol           = (WallArea + 2 * RoofArea) / (WallArea + RoofArea),
      GlazingXHeight      = GlazingArea * OverallHeight,
      WallXGlazing        = WallArea * GlazingArea,
      CompactnessSquared  = RelCompactness^2,
      HeightSquared       = OverallHeight^2,
      GlazingToWall       = GlazingArea / (WallArea + 1),
      EnvelopeArea        = WallArea + RoofArea,
      WallPerHeight       = WallArea / (OverallHeight + 1)
    )

  # Scale using the training set parameters
  base_grid[, features] <- predict(preproc_params,
                                    base_grid[, features])

  preds <- predict_fn(as.matrix(base_grid[, features]))
  base_grid$pred_HL <- preds

  # Test 1: glazing ordering — higher glazing → higher heating load
  glaz_pairs <- base_grid %>%
    group_by(RelCompactness, WallArea, RoofArea,
             OverallHeight, GlazingAreaDist) %>%
    arrange(GlazingArea) %>%
    summarise(
      correct = all(diff(pred_HL) > 0),
      .groups = "drop"
    )
  glaz_score <- mean(glaz_pairs$correct) * 100

  # Test 2: compactness ordering — higher compactness → lower heating load
  comp_pairs <- base_grid %>%
    group_by(WallArea, RoofArea, OverallHeight,
             GlazingArea, GlazingAreaDist) %>%
    arrange(RelCompactness) %>%
    summarise(
      correct = all(diff(pred_HL) < 0),
      .groups = "drop"
    )
  comp_score <- mean(comp_pairs$correct) * 100

  # Test 3: height ordering — taller → higher heating load
  ht_pairs <- base_grid %>%
    group_by(RelCompactness, WallArea, RoofArea,
             GlazingArea, GlazingAreaDist) %>%
    arrange(OverallHeight) %>%
    summarise(
      correct = all(diff(pred_HL) > 0),
      .groups = "drop"
    )
  ht_score <- mean(ht_pairs$correct) * 100

  cat(sprintf("%-25s  Glazing: %5.1f%%  Compactness: %5.1f%%  Height: %5.1f%%\n",
              model_name, glaz_score, comp_score, ht_score))

  data.frame(Model               = model_name,
             Glazing_Ordering_pct  = glaz_score,
             Compact_Ordering_pct  = comp_score,
             Height_Ordering_pct   = ht_score)
}

# Run the check for all three models
physics_results <- bind_rows(
  check_physics(gpr_rbf_y1, "GPR (RBF)",        model_type = "gausspr"),
  check_physics(gpr_mat_y1, "GPR (Matern 5/2)", model_type = "gausspr"),
  check_physics(lgbm_y1,    "LightGBM",         model_type = "lgbm")
)

cat("\n=== PHYSICS-CONSISTENCY SUMMARY ===\n")
print(physics_results, row.names = FALSE)

# Plot
p_physics <- physics_results %>%
  pivot_longer(cols = ends_with("_pct"),
               names_to  = "Check",
               values_to = "Score") %>%
  mutate(Check = recode(Check,
    "Glazing_Ordering_pct"  = "Glazing → HL",
    "Compact_Ordering_pct"  = "Compactness → HL",
    "Height_Ordering_pct"   = "Height → HL"
  )) %>%
  ggplot(aes(x = Model, y = Score, fill = Check)) +
  geom_col(position = "dodge", alpha = 0.85) +
  geom_text(aes(label = paste0(round(Score, 0), "%")),
            position = position_dodge(width = 0.9),
            vjust = -0.4, size = 3) +
  geom_hline(yintercept = 100, linetype = "dashed",
             colour = "darkgreen", linewidth = 0.7) +
  scale_fill_manual(values = c("steelblue", "tomato", "goldenrod")) +
  scale_y_continuous(limits = c(0, 115)) +
  labs(
    title    = "Physics-Consistency Check — Heating Load",
    subtitle = "Percentage of building pairs where model predicts the physically correct direction",
    x = "Model", y = "Correct ordering (%)", fill = "Physical rule"
  ) +
  theme_minimal()

ggsave(file.path(img_dir, "physics_consistency_check.png"),
       plot = p_physics, width = 10, height = 6, dpi = 150)
print(p_physics)
```

---

# Model Comparison

## Heating Load (Y1)

```{r comparison-y1}
results_y1 <- results_y1 %>% arrange(RMSE)

cat("\n=== HEATING LOAD — MODEL COMPARISON ===\n")
print(results_y1, row.names = FALSE)

p_rmse_y1 <- ggplot(results_y1, aes(x = reorder(Model, RMSE), y = RMSE)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = round(RMSE, 4)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  labs(title = "Model Comparison — RMSE: Heating Load (Y1)",
       x = "Model", y = "RMSE (kWh/m²)") +
  theme_minimal()

ggsave(file.path(img_dir, "comparison_domain_y1_rmse.png"),
       plot = p_rmse_y1, width = 9, height = 5, dpi = 150)
print(p_rmse_y1)

p_r2_y1 <- ggplot(results_y1, aes(x = reorder(Model, -R2), y = R2)) +
  geom_col(fill = "darkgreen", alpha = 0.75) +
  geom_text(aes(label = round(R2, 4)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  labs(title = "Model Comparison — R²: Heating Load (Y1)",
       x = "Model", y = "R²") +
  theme_minimal()

ggsave(file.path(img_dir, "comparison_domain_y1_r2.png"),
       plot = p_r2_y1, width = 9, height = 5, dpi = 150)
print(p_r2_y1)
```

## Cooling Load (Y2)

```{r comparison-y2}
results_y2 <- results_y2 %>% arrange(RMSE)

cat("\n=== COOLING LOAD — MODEL COMPARISON ===\n")
print(results_y2, row.names = FALSE)

p_rmse_y2 <- ggplot(results_y2, aes(x = reorder(Model, RMSE), y = RMSE)) +
  geom_col(fill = "darkorange", alpha = 0.8) +
  geom_text(aes(label = round(RMSE, 4)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  labs(title = "Model Comparison — RMSE: Cooling Load (Y2)",
       x = "Model", y = "RMSE (kWh/m²)") +
  theme_minimal()

ggsave(file.path(img_dir, "comparison_domain_y2_rmse.png"),
       plot = p_rmse_y2, width = 9, height = 5, dpi = 150)
print(p_rmse_y2)

p_r2_y2 <- ggplot(results_y2, aes(x = reorder(Model, -R2), y = R2)) +
  geom_col(fill = "darkgreen", alpha = 0.75) +
  geom_text(aes(label = round(R2, 4)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  labs(title = "Model Comparison — R²: Cooling Load (Y2)",
       x = "Model", y = "R²") +
  theme_minimal()

ggsave(file.path(img_dir, "comparison_domain_y2_r2.png"),
       plot = p_r2_y2, width = 9, height = 5, dpi = 150)
print(p_r2_y2)
```