---
title: "new"
author: "Jakes Chege"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 11, fig.height = 7)
```

```{r packages}
pkgs <- c("tidyverse", "caret", "randomForest", "xgboost", "nnet",
          "mgcv", "splines", "corrplot", "ggplot2", "ggplot", "dplyr",
          "pdp", "vip", "readxl")
```

Description of the variables in the data
* $X_1$ 	Relative Compactness
* $X_2$ 	Surface Area
* $X_3$ 	Wall Area
* $X_4$ 	Roof Area
* $X_5$ 	Overall Height
* $X_6$ 	Orientation
* $X_7$ 	Glazing Area
* $X_8$ 	Glazing Area Distribution
* $Y_1$ 	Heating Load
* $Y_2$ 	Cooling Load


The variables labeled 
* $X_i$: attributes (covariates) 
* $Y_i$: supposed to be outcome variables 

The aim is to use the features to predict each of the outcome responses. Can you think of classification models, regression models (parametric, semi-parametric, non-linear) where possible and be able to perform analysis that can assess the heating load and cooling load requirements of buildings. 

I expect to see very robust analysis and we should have a very well done presentation to assess the overall understanding of this unit as we approach the examination date. 

```{r load-data}
df <- read_excel("~/Documents/strathmore/Modules/Module 3/linear_models/classwork/final_project/data/ENB2012_data 2.xlsx")

head(df)
```


```{r}
colnames(df) <- c(
  "RelCompactness",   # X1
  "SurfaceArea",      # X2
  "WallArea",         # X3
  "RoofArea",         # X4
  "OverallHeight",    # X5
  "Orientation",      # X6
  "GlazingArea",      # X7
  "GlazingAreaDist",  # X8
  "HeatingLoad",      # Y1
  "CoolingLoad"       # Y2
)
```

```{r}
summary(df)
```
```{r}
df_orig <- read_excel("~/Documents/strathmore/Modules/Module 3/linear_models/classwork/final_project/data/ENB2012_data 2.xlsx")
```

```{r}
cat("Dimensions:", dim(df), "\n")
str(df)
```

```{r}
head(df, 10)
tail(df, 10)
```

```{r}
library(psych)
psych::describe(df)
```

```{r}
cat("\nMissing values per column:\n")
colSums(is.na(df))

cat("\nTotal missing:", sum(is.na(df)), "\n")
```

```{r}
cat("Duplicated rows:", sum(duplicated(df)), "\n")

# View some duplicates if they exist
if (sum(duplicated(df)) > 0) {
  df[duplicated(df) | duplicated(df, fromLast = TRUE), ] %>% 
    arrange(RelCompactness, SurfaceArea) %>% 
    head(20)
}
```
# 2.7  Unique value counts per variable
```{r}
sapply(df, function(x) length(unique(x)))
```
# 2.8  Data types check
```{r}
sapply(df, class)
```
# Note: OverallHeight has only 2 levels (3.5, 7.0) — essentially binary
# Orientation has 4 levels (2,3,4,5) — categorical
# GlazingArea has 4 levels (0, 0.1, 0.25, 0.4) — ordinal/categorical
# GlazingAreaDist has 6 levels (0–5) — categorical
```{r}
# Convert appropriate variables to factors for certain analyses
df$Orientation_f     <- factor(df$Orientation)
df$OverallHeight_f   <- factor(df$OverallHeight, labels = c("Low_3.5m", "High_7m"))
df$GlazingArea_f     <- factor(df$GlazingArea)
df$GlazingAreaDist_f <- factor(df$GlazingAreaDist)
```

```{r}
sapply(df, class)
```

```{r}
head(df)
```
# 3.1  Histograms for all continuous variables
```{r}
cont_vars <- c("RelCompactness", "SurfaceArea", "WallArea", "RoofArea",
               "HeatingLoad", "CoolingLoad")

p_hist <- list()
for (v in cont_vars) {
  p_hist[[v]] <- ggplot(df, aes(x = .data[[v]])) +
    geom_histogram(aes(y = after_stat(density)), bins = 30,
                   fill = "steelblue", color = "white", alpha = 0.7) +
    geom_density(color = "red", linewidth = 0.8) +
    labs(title = paste("Distribution of", v), x = v, y = "Density") +
    theme_minimal(base_size = 11)
}
do.call(grid.arrange, c(p_hist, ncol = 3))
```
# 3.2  Boxplots for all numeric variables
```{r}
library(tidyverse)
df_long <- df %>%
  select(RelCompactness:CoolingLoad) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value")

ggplot(df_long, aes(x = Variable, y = Value, fill = Variable)) +
  geom_boxplot(outlier.colour = "red", outlier.size = 1.5) +
  facet_wrap(~ Variable, scales = "free", ncol = 5) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none",
        axis.text.x = element_blank()) +
  labs(title = "Boxplots of All Variables")
```
# 3.3  Bar charts for categorical/discrete variables
```{r}
cat_vars <- c("Orientation", "OverallHeight", "GlazingArea", "GlazingAreaDist")

p_bar <- list()
for (v in cat_vars) {
  p_bar[[v]] <- ggplot(df, aes(x = factor(.data[[v]]))) +
    geom_bar(fill = "coral", color = "white", alpha = 0.8) +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.3) +
    labs(title = paste("Frequency of", v), x = v, y = "Count") +
    theme_minimal(base_size = 11)
}
do.call(grid.arrange, c(p_bar, ncol = 2))
```

# 3.4  QQ-plots for normality assessment of Y1 and Y2
```{r}
par(mfrow = c(1, 2))
qqnorm(df$HeatingLoad, main = "Q-Q Plot: Heating Load (Y1)")
qqline(df$HeatingLoad, col = "red", lwd = 2)

qqnorm(df$CoolingLoad, main = "Q-Q Plot: Cooling Load (Y2)")
qqline(df$CoolingLoad, col = "red", lwd = 2)
par(mfrow = c(1, 1))
```
# 3.5  Formal normality tests on the responses
```{r}
cat("Shapiro-Wilk test for Y1 (Heating Load):\n")
shapiro.test(df$HeatingLoad[1:5000])  # Shapiro max 5000 obs, sample if needed
# If n > 5000, use Anderson-Darling:
nortest::ad.test(df$HeatingLoad)

cat("\nShapiro-Wilk test for Y2 (Cooling Load):\n")
shapiro.test(df$CoolingLoad)
nortest::ad.test(df$CoolingLoad)
```
# 3.6  Ridgeline plots — distribution by OverallHeight
```{r}
library(ggridges)
ggplot(df, aes(x = HeatingLoad, y = OverallHeight_f, fill = OverallHeight_f)) +
  geom_density_ridges(alpha = 0.7) +
  labs(title = "Heating Load Distribution by Overall Height",
       x = "Heating Load", y = "Overall Height") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(df, aes(x = CoolingLoad, y = OverallHeight_f, fill = OverallHeight_f)) +
  geom_density_ridges(alpha = 0.7) +
  labs(title = "Cooling Load Distribution by Overall Height",
       x = "Cooling Load", y = "Overall Height") +
  theme_minimal() +
  theme(legend.position = "none")
```
# 4.1  Scatter plots: each X vs Y1 and Y2
```{r}
feature_vars <- c("RelCompactness", "SurfaceArea", "WallArea", "RoofArea",
                  "OverallHeight", "Orientation", "GlazingArea", "GlazingAreaDist")

# --- Against Heating Load ---
p_scat_y1 <- list()
for (v in feature_vars) {
  p_scat_y1[[v]] <- ggplot(df, aes(x = .data[[v]], y = HeatingLoad)) +
    geom_point(alpha = 0.4, color = "steelblue", size = 1) +
    geom_smooth(method = "loess", color = "red", se = TRUE, linewidth = 0.8) +
    labs(title = paste(v, "vs Heating Load"), x = v, y = "Heating Load") +
    theme_minimal(base_size = 9)
}
do.call(grid.arrange, c(p_scat_y1, ncol = 4))

# --- Against Cooling Load ---
p_scat_y2 <- list()
for (v in feature_vars) {
  p_scat_y2[[v]] <- ggplot(df, aes(x = .data[[v]], y = CoolingLoad)) +
    geom_point(alpha = 0.4, color = "darkorange", size = 1) +
    geom_smooth(method = "loess", color = "blue", se = TRUE, linewidth = 0.8) +
    labs(title = paste(v, "vs Cooling Load"), x = v, y = "Cooling Load") +
    theme_minimal(base_size = 9)
}
do.call(grid.arrange, c(p_scat_y2, ncol = 4))
```

The heating load is the amount of heat energy the building loses to the outside environment. To maintain a steady indoor temperature (e.g., 20°C), the heating system must produce enough heat to replace what is lost.

The cooling load is the amount of heat energy the building gains from the outside environment and internal sources. The air conditioning system must remove this excess heat to keep the indoor temperature at a comfortable level.

Physical Design Variables

    RelCompactness (Relative Compactness): This measures how "compact" the building shape is. A more compact building (like a cube) has a lower ratio of surface area to volume, which generally helps keep heat inside.

    SurfaceArea: The total exterior area of the building (walls, roof, and floor). Generally, a larger surface area allows more heat to escape to the outside environment.

    WallArea: The total square footage of the exterior walls. Larger wall areas usually result in higher heat loss because walls are a primary path for heat transfer.

    RoofArea: The total area of the roof. Like walls, the roof is a significant surface where heat can escape, especially in colder climates.

    OverallHeight: How tall the building is. Taller buildings often have more exposed surface area, which can increase the heating load.

Orientation and Window Variables

    Orientation: The direction the building faces (e.g., North, East, South, West). This determines how much direct sunlight the house receives, which can help warm it naturally (passive solar gain).

    GlazingArea: The percentage of the floor area that is covered by glass (windows). More glass generally means more potential for heat loss in winter, though it also allows for more solar heating.

    GlazingAreaDist (Glazing Area Distribution): This describes where the windows are located on the building (e.g., uniformly distributed, or concentrated on one side like the South side to maximize solar heat).
    
# 4.2  Scatter with colour by Overall Height
```{r}
ggplot(df, aes(x = RelCompactness, y = HeatingLoad, color = OverallHeight_f)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Relative Compactness vs Heating Load by Height",
       color = "Overall Height") +
  theme_minimal()

ggplot(df, aes(x = WallArea, y = CoolingLoad, color = OverallHeight_f)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Wall Area vs Cooling Load by Height",
       color = "Overall Height") +
  theme_minimal()
```

# 4.3  Boxplots of Y1/Y2 across categorical features
```{r}
# Heating Load by Orientation
ggplot(df, aes(x = Orientation_f, y = HeatingLoad, fill = Orientation_f)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Heating Load by Orientation", x = "Orientation") +
  theme_minimal() + theme(legend.position = "none")

# Heating Load by Glazing Area
ggplot(df, aes(x = GlazingArea_f, y = HeatingLoad, fill = GlazingArea_f)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Heating Load by Glazing Area", x = "Glazing Area") +
  theme_minimal() + theme(legend.position = "none")

# Cooling Load by Glazing Area Distribution
ggplot(df, aes(x = GlazingAreaDist_f, y = CoolingLoad, fill = GlazingAreaDist_f)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Cooling Load by Glazing Area Distribution", x = "Glazing Area Distribution") +
  theme_minimal() + theme(legend.position = "none")

# Heating Load by Overall Height
ggplot(df, aes(x = OverallHeight_f, y = HeatingLoad, fill = OverallHeight_f)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Heating Load by Overall Height") +
  theme_minimal() + theme(legend.position = "none")
```
The more the number of windows, the more the heat the house is likely to lose to the outside environment.
The less the distance between windows, the more the heat gained from the outside environment.
The taller the building, the more the heat lost to the outside environment.

# 4.4  Violin plots (richer than boxplots)
```{r}
ggplot(df, aes(x = GlazingArea_f, y = HeatingLoad, fill = GlazingArea_f)) +
  geom_violin(alpha = 0.6) +
  geom_boxplot(width = 0.15, fill = "white") +
  labs(title = "Heating Load by Glazing Area (Violin + Boxplot)") +
  theme_minimal() + theme(legend.position = "none")
```
# 4.5  Interaction plots
```{r}
# Two-way interaction: OverallHeight × GlazingArea on HeatingLoad
df %>%
  group_by(OverallHeight_f, GlazingArea_f) %>%
  summarise(mean_HL = mean(HeatingLoad), .groups = "drop") %>%
  ggplot(aes(x = GlazingArea_f, y = mean_HL, 
             color = OverallHeight_f, group = OverallHeight_f)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(title = "Interaction: Overall Height × Glazing Area on Heating Load",
       x = "Glazing Area", y = "Mean Heating Load", color = "Height") +
  theme_minimal()
```

# 5.1  Pearson correlation matrix
```{r}
cor_matrix <- cor(df %>% select(RelCompactness:CoolingLoad), method = "pearson")
round(cor_matrix, 3)
```
```{r}
# ============================================================
# 5.2  Correlation heatmap (corrplot)
# ============================================================
library(corrplot)
corrplot(cor_matrix, method = "color", type = "upper",
         addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         col = colorRampPalette(c("#6D9EC1", "white", "#E46726"))(200),
         title = "Pearson Correlation Matrix",
         mar = c(0, 0, 2, 0))
```
Overall Height seems to be influential in determining the heating and cooling loads of a building. Heating and cooling load are also highly correlated. There is also a high negative correlation between Relative compactness and surface area.

```{r}
# ============================================================
# 5.3  ggcorrplot version (publication quality)
# ============================================================
library(ggcorrplot)
ggcorrplot(cor_matrix, type = "lower", lab = TRUE, lab_size = 3,
           colors = c("#2166AC", "white", "#B2182B"),
           title = "Correlation Heatmap",
           ggtheme = theme_minimal())
```

```{r}
# ============================================================
# 5.6  Multicollinearity check (VIF) with a preliminary linear model
# ============================================================
# Because SurfaceArea = WallArea + RoofArea, the model cannot mathematically solve the equation because the columns are perfectly correlated.
# Option: Remove SurfaceArea to resolve the aliasing
prelim_model_fixed <- lm(HeatingLoad ~ RelCompactness + WallArea + RoofArea + 
                           OverallHeight + Orientation + GlazingArea + 
                           GlazingAreaDist, data = df)

# Now check the VIF again
vif(prelim_model_fixed)

# VIF > 10 signals severe multicollinearity
# VIF > 5  signals moderate concern
```
When VIF scores are this high (especially those over 100), it tells us that your independent variables are not truly independent.

```{r}
# ============================================================
# 6.1  Boxplot-based outlier identification
# ============================================================
detect_outliers <- function(x) {
  q1 <- quantile(x, 0.25)
  q3 <- quantile(x, 0.75)
  iqr <- q3 - q1
  lower <- q1 - 1.5 * iqr
  upper <- q3 + 1.5 * iqr
  return(sum(x < lower | x > upper))
}

cat("Outlier counts (IQR method):\n")
sapply(df %>% select(RelCompactness:CoolingLoad), detect_outliers)
```


```{r}
# ============================================================
# 6.3  Cook's Distance from the linear model
# ============================================================
cooks_d <- cooks.distance(prelim_model)
plot(cooks_d, type = "h", main = "Cook's Distance", ylab = "Cook's D")
abline(h = 4 / nrow(df), col = "red", lty = 2)
cat("Influential points (Cook's D > 4/n):", sum(cooks_d > 4/nrow(df)), "\n")
```
