---
title: "new"
author: "Jakes Chege"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 11, fig.height = 7)
```

```{r packages}
pkgs <- c("tidyverse", "caret", "randomForest", "xgboost", "nnet",
          "mgcv", "splines", "corrplot", "ggplot2", "dplyr",
          "pdp", "vip", "readxl")
```

Description of the variables in the data
* $X_1$ 	Relative Compactness
* $X_2$ 	Surface Area
* $X_3$ 	Wall Area
* $X_4$ 	Roof Area
* $X_5$ 	Overall Height
* $X_6$ 	Orientation
* $X_7$ 	Glazing Area
* $X_8$ 	Glazing Area Distribution
* $Y_1$ 	Heating Load
* $Y_2$ 	Cooling Load


The variables labeled 
* $X_i$: attributes (covariates) 
* $Y_i$: supposed to be outcome variables 

The aim is to use the features to predict each of the outcome responses. Can you think of classification models, regression models (parametric, semi-parametric, non-linear) where possible and be able to perform analysis that can assess the heating load and cooling load requirements of buildings. 

I expect to see very robust analysis and we should have a very well done presentation to assess the overall understanding of this unit as we approach the examination date. 

```{r load-data}
data <- read_excel("~/Documents/strathmore/Modules/Module 3/linear_models/classwork/notes/ENB2012_data 2.xlsx")
head(data)
```


```{r}
summary(data)
```

