---
title: "Final Project — Building Energy Load: Integrated Analysis"
author: "Group 5"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo    = TRUE
)
```

---

# Overview

This document integrates all stages of the ENB2012 building energy analysis
into a single reproducible workflow. The sections follow a deliberate
progression:

1. **EDA** — characterise the data structure, identify multicollinearity, and
   establish the physical signals that every subsequent model will need to
   respect.
2. **Linear Regression** — establish a regularised linear baseline using
   Ridge, Lasso, and Elastic Net, motivated directly by the VIF findings from
   the EDA.
3. **Domain-Appropriate Regression** — apply Gaussian Process Regression and
   LightGBM, which are specifically suited to simulation-generated tabular
   data; compare against the linear baseline from the previous section.
4. **Domain-Appropriate Classification** — reframe the prediction problem as
   ordinal energy class assignment, using models that respect the Low/Medium/High
   ordering and a regulatory domain baseline for comparison.
5. **Cross-Model Comparison and Conclusions** — synthesise regression and
   classification results, identify which models to carry forward, and
   contextualise the findings in terms of physical interpretability.

The same 80/20 group-aware train/test split (seed = 42) and the same 16
engineered features are used throughout. All scalers are fitted on training
data only.

---

# Packages

```{r packages}
pkgs <- c(
  # Core
  "tidyverse", "readxl", "writexl",
  # EDA
  "corrplot", "ggcorrplot", "ggridges", "psych", "nortest",
  "car", "gridExtra",
  # Regression
  "caret", "glmnet", "broom", "lmtest",
  # Domain-appropriate regression
  "kernlab",   # GPR
  "lightgbm",  # LightGBM
  # Classification
  "ranger",       # probability forest (used in place of ordinalForest)
  "naivebayes",   # Gaussian Naive Bayes
  "pROC",         # Multiclass AUC
  "knitr"
)
lapply(pkgs, library, character.only = TRUE)
```

```{r img-dir}
img_dir <- "images"
dir.create(img_dir, showWarnings = FALSE, recursive = TRUE)
```

---

# Part 1 — Exploratory Data Analysis

## Data Loading and Variable Definitions

```{r eda-load}
df <- read_excel(
  "~/Documents/strathmore/Modules/Module 3/linear_models/classwork/final_project/data/ENB2012_data 2.xlsx"
)

colnames(df) <- c(
  "RelCompactness", "SurfaceArea", "WallArea", "RoofArea",
  "OverallHeight", "Orientation", "GlazingArea",
  "GlazingAreaDist", "HeatingLoad", "CoolingLoad"
)

cat("Dimensions:", nrow(df), "rows x", ncol(df), "columns\n")
```

## Dataset Structure and the Combinatorial Grid

ENB2012 was generated by the EnergyPlus simulator by systematically varying
eight building parameters across fixed discrete levels. This gives the dataset
a combinatorial grid structure where every combination of parameter values
appears exactly four times — once per compass orientation. The critical
consequence for modelling is that a naive random train/test split will leak
building configurations between sets, since orientation variants of the same
building will appear on both sides. All downstream modelling accounts for this
through a group-aware split.

```{r eda-structure}
cat("Unique values per variable:\n")
sapply(df, function(x) length(unique(x)))

# Group key: uniquely identifies each building configuration (excluding orientation)
df <- df %>%
  mutate(group_key = paste(RelCompactness, WallArea, RoofArea,
                           OverallHeight, GlazingArea, GlazingAreaDist,
                           sep = "_"))

cat("\nTotal rows:", nrow(df), "\n")
cat("Unique building configurations:", length(unique(df$group_key)), "\n")
cat("Rows per configuration (should be 4):\n")
print(table(table(df$group_key)))
```

## Missing Values, Duplicates, and Data Quality

```{r eda-quality}
cat("Missing values per column:\n")
print(colSums(is.na(df)))

cat("\nDuplicated rows:", sum(duplicated(df)), "\n")
```

ENB2012 has no missing values by construction. Any duplicated rows would
indicate a simulation export error; none are present.

## Univariate Distributions

### Continuous variables

```{r eda-hist}
cont_vars <- c("RelCompactness", "SurfaceArea", "WallArea",
               "RoofArea", "HeatingLoad", "CoolingLoad")

p_hist <- lapply(cont_vars, function(v) {
  ggplot(df, aes(x = .data[[v]])) +
    geom_histogram(aes(y = after_stat(density)), bins = 30,
                   fill = "steelblue", color = "white", alpha = 0.7) +
    geom_density(color = "red", linewidth = 0.8) +
    labs(title = v, x = v, y = "Density") +
    theme_minimal(base_size = 10)
})
do.call(grid.arrange, c(p_hist, ncol = 3))
```

`RelCompactness` is roughly uniform, reflecting the systematic grid design.
`HeatingLoad` and `CoolingLoad` are both bimodal, with modes corresponding to
the two levels of `OverallHeight`. This bimodality is structural, not noise,
and motivates treating `OverallHeight` as a key stratifying variable
throughout the analysis.

### Discrete and categorical variables

```{r eda-factors}
df$Orientation_f     <- factor(df$Orientation,     levels = 2:5,
                                labels = c("N", "E", "S", "W"))
df$OverallHeight_f   <- factor(df$OverallHeight,   levels = c(3.5, 7.0),
                                labels = c("Low_3.5m", "High_7m"))
df$GlazingArea_f     <- factor(df$GlazingArea,     levels = c(0, 0.1, 0.25, 0.4),
                                labels = c("None","Low_10pct","Med_25pct","High_40pct"))
df$GlazingAreaDist_f <- factor(df$GlazingAreaDist, levels = 0:5,
                                labels = c("None","Uniform","N_heavy",
                                           "E_heavy","S_heavy","W_heavy"))

cat_vars <- c("Orientation_f", "OverallHeight_f", "GlazingArea_f", "GlazingAreaDist_f")

p_bar <- lapply(cat_vars, function(v) {
  ggplot(df, aes(x = .data[[v]])) +
    geom_bar(fill = "coral", color = "white", alpha = 0.8) +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.3, size = 3) +
    labs(title = v, x = NULL, y = "Count") +
    theme_minimal(base_size = 10)
})
do.call(grid.arrange, c(p_bar, ncol = 2))
```

All four orientation values appear with equal frequency (192 buildings each),
confirming the balanced grid design. `GlazingAreaDist` level 0 (no distribution)
corresponds exactly to buildings with zero `GlazingArea`, creating a structural
dependency between these two variables.

## Response Variable Analysis

### Normality assessment

```{r eda-normality}
par(mfrow = c(1, 2))
qqnorm(df$HeatingLoad, main = "Q-Q Plot: Heating Load")
qqline(df$HeatingLoad, col = "red", lwd = 2)
qqnorm(df$CoolingLoad, main = "Q-Q Plot: Cooling Load")
qqline(df$CoolingLoad, col = "red", lwd = 2)
par(mfrow = c(1, 1))

cat("Anderson-Darling test — Heating Load:\n")
print(nortest::ad.test(df$HeatingLoad))

cat("\nAnderson-Darling test — Cooling Load:\n")
print(nortest::ad.test(df$CoolingLoad))
```

Both responses deviate significantly from normality (AD p < 2.2e-16 for
both), consistent with their bimodal distributions. This does not invalidate
linear regression but means residual diagnostics will matter. It directly
motivates exploring nonlinear models and classification approaches alongside
the linear models in later sections.

### Distribution by Overall Height

```{r eda-ridges}
p_rl1 <- ggplot(df, aes(x = HeatingLoad, y = OverallHeight_f,
                         fill = OverallHeight_f)) +
  geom_density_ridges(alpha = 0.7) +
  labs(title = "Heating Load by Overall Height",
       x = "Heating Load (kWh/m²)", y = NULL) +
  theme_minimal() + theme(legend.position = "none")

p_rl2 <- ggplot(df, aes(x = CoolingLoad, y = OverallHeight_f,
                         fill = OverallHeight_f)) +
  geom_density_ridges(alpha = 0.7) +
  labs(title = "Cooling Load by Overall Height",
       x = "Cooling Load (kWh/m²)", y = NULL) +
  theme_minimal() + theme(legend.position = "none")

grid.arrange(p_rl1, p_rl2, ncol = 2)
```

`OverallHeight` cleanly separates the load distributions into two
non-overlapping bands. Low buildings (3.5m) cluster at low loads; tall
buildings (7.0m) cluster at high loads. This is the strongest single-variable
signal in the dataset and will appear consistently across all model importance
rankings.

## Feature-Response Relationships

### Scatter plots with LOESS smooths

```{r eda-loess}
feature_vars <- c("RelCompactness", "SurfaceArea", "WallArea", "RoofArea",
                  "OverallHeight", "GlazingArea", "GlazingAreaDist")

p_y1 <- lapply(feature_vars, function(v) {
  ggplot(df, aes(x = .data[[v]], y = HeatingLoad)) +
    geom_point(alpha = 0.3, color = "steelblue", size = 0.8) +
    geom_smooth(method = "loess", color = "red", se = TRUE, linewidth = 0.8) +
    labs(title = paste(v, "vs HL"), x = v, y = "Heating Load") +
    theme_minimal(base_size = 9)
})
do.call(grid.arrange, c(p_y1, ncol = 4))

p_y2 <- lapply(feature_vars, function(v) {
  ggplot(df, aes(x = .data[[v]], y = CoolingLoad)) +
    geom_point(alpha = 0.3, color = "darkorange", size = 0.8) +
    geom_smooth(method = "loess", color = "blue", se = TRUE, linewidth = 0.8) +
    labs(title = paste(v, "vs CL"), x = v, y = "Cooling Load") +
    theme_minimal(base_size = 9)
})
do.call(grid.arrange, c(p_y2, ncol = 4))
```

`RelCompactness` and `OverallHeight` show strong monotonic relationships with
both responses. `GlazingArea` shows a positive nonlinear effect.
`GlazingAreaDist` shows a near-flat LOESS curve, indicating weak marginal
signal — this finding is confirmed later by Lasso zeroing it in both models
and by its non-significance in the baseline OLS (p=0.833 for Orientation,
p=0.347 for GlazingAreaDist in the CL model).

### Interaction: OverallHeight × GlazingArea

```{r eda-interaction}
df %>%
  group_by(OverallHeight_f, GlazingArea_f) %>%
  summarise(mean_HL = mean(HeatingLoad),
            mean_CL = mean(CoolingLoad), .groups = "drop") %>%
  pivot_longer(c(mean_HL, mean_CL),
               names_to = "Response", values_to = "Mean") %>%
  mutate(Response = case_when(
    Response == "mean_HL" ~ "Heating Load",
    Response == "mean_CL" ~ "Cooling Load"
  )) %>%
  ggplot(aes(x = GlazingArea_f, y = Mean,
             color = OverallHeight_f, group = OverallHeight_f)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~ Response, scales = "free_y") +
  labs(title    = "Interaction: Overall Height × Glazing Area",
       x = "Glazing Area", y = "Mean Load (kWh/m²)", color = "Height") +
  theme_minimal()
```

The interaction plot shows that the effect of glazing area on heating load is
steeper for tall buildings than for short ones. This crossing pattern confirms
a genuine interaction effect and directly motivates including
`OverallHeight × GlazingArea` interaction terms in the regression models. The
coefficient for this interaction (`GlazingXHeight`) is −1.20 (p=0.326) in the
OLS with interactions — not significant because its signal is absorbed by the
correlated `CompactnessXHeight` (−60.5, p<0.001) and `GlazingXCompactness`
(+8.3, p<0.001) terms.

## Correlation and Multicollinearity

### Correlation heatmap

```{r eda-corrplot}
cor_matrix <- cor(df %>% dplyr::select(RelCompactness:CoolingLoad),
                  method = "pearson")

ggcorrplot(cor_matrix, type = "lower", lab = TRUE, lab_size = 3,
           colors = c("#2166AC", "white", "#B2182B"),
           title  = "Pearson Correlation Matrix",
           ggtheme = theme_minimal())
```

`SurfaceArea` is almost perfectly negatively correlated with `RelCompactness`
(r ≈ −0.99) — a mathematical near-identity because relative compactness is
defined as the volume-to-surface-area ratio. Including both causes exact or
near-exact multicollinearity and `SurfaceArea` is excluded from all models.
`HeatingLoad` and `CoolingLoad` are highly correlated (r ≈ 0.98), reflecting
that the same physical factors drive both; they are modelled separately
throughout.

### VIF analysis

```{r eda-vif}
prelim_model <- lm(HeatingLoad ~ RelCompactness + WallArea + RoofArea +
                     OverallHeight + Orientation + GlazingArea +
                     GlazingAreaDist, data = df)

cat("VIF — SurfaceArea excluded to resolve perfect collinearity:\n")
print(car::vif(prelim_model))
```

The VIF results establish the severity of the collinearity problem.
`RoofArea` (VIF=211.9) and `RelCompactness` (VIF=105.5) are almost entirely
explained by linear combinations of the other predictors. `WallArea` (27.7)
and `OverallHeight` (31.2) also breach the threshold of 10 by a wide margin.
Only `Orientation`, `GlazingArea`, and `GlazingAreaDist` are clean. This
directly motivates the choice of regularised regression in Part 2 — Ridge and
Elastic Net are designed to handle correlated predictors by shrinking
coefficients rather than relying on matrix inversion stability.

## Outlier Assessment

```{r eda-outliers}
detect_outliers <- function(x) {
  q1  <- quantile(x, 0.25)
  q3  <- quantile(x, 0.75)
  iqr <- q3 - q1
  sum(x < (q1 - 1.5 * iqr) | x > (q3 + 1.5 * iqr))
}

cat("IQR-based outlier counts:\n")
print(sapply(df %>% dplyr::select(RelCompactness:CoolingLoad),
             detect_outliers))
```

No IQR-flagged rows are removed. Every row corresponds to a legitimate
physical building configuration in the simulation grid. IQR flags here reflect
the discrete multimodal structure of the data rather than measurement error.

## Export Post-EDA Dataset

```{r eda-export}
write_xlsx(df, "ENB2012_post_EDA.xlsx")
cat("Exported: ENB2012_post_EDA.xlsx\n")
cat("Dimensions:", nrow(df), "rows x", ncol(df), "columns\n")
```

---

# Part 2 — Linear Regression Models

## Data Loading and Group-Aware Split

The EDA established that ENB2012 has a combinatorial grid structure where each
building configuration appears four times, once per orientation. A naive random
split would leak building shapes between training and test sets, inflating
apparent performance. All splits here keep all four orientation variants of
each configuration on the same side.

```{r reg-split}
df <- read_excel("ENB2012_post_EDA.xlsx")

df$Orientation_f     <- factor(df$Orientation, levels = 2:5,
                                labels = c("N", "E", "S", "W"))
df$OverallHeight_f   <- factor(df$OverallHeight, levels = c(3.5, 7.0),
                                labels = c("Low_3.5m", "High_7m"))
df$GlazingArea_f     <- factor(df$GlazingArea, levels = c(0, 0.1, 0.25, 0.4),
                                labels = c("None","Low_10pct","Med_25pct","High_40pct"))
df$GlazingAreaDist_f <- factor(df$GlazingAreaDist, levels = 0:5,
                                labels = c("None","Uniform","N_heavy",
                                           "E_heavy","S_heavy","W_heavy"))

df <- df %>%
  mutate(group_key = paste(RelCompactness, WallArea, RoofArea,
                           OverallHeight, GlazingArea, GlazingAreaDist, sep = "_"))

set.seed(42)
unique_groups  <- unique(df$group_key)
n_train_groups <- round(0.80 * length(unique_groups))
train_groups   <- sample(unique_groups, n_train_groups)

df_train <- df %>% filter( group_key %in% train_groups)
df_test  <- df %>% filter(!group_key %in% train_groups)

cat("Training rows:", nrow(df_train), "\n")
cat("Test rows    :", nrow(df_test),  "\n")
cat("Group overlap:", length(intersect(df_train$group_key, df_test$group_key)),
    "<- zero confirms no leakage\n")
```

## Feature Engineering

The EDA interaction plot showed that the effect of glazing area on heating load
is steeper for tall buildings, confirming a genuine `OverallHeight × GlazingArea`
interaction. The VIF analysis showed severe collinearity among the geometric
variables, which motivates constructing physically interpretable composite
features that capture cross-variable relationships while reducing redundancy.

```{r reg-engineer}
engineer <- function(d) {
  d %>% mutate(
    CompactnessXHeight  = RelCompactness * OverallHeight,
    WallToRoofRatio     = WallArea / RoofArea,
    GlazingXCompactness = GlazingArea * RelCompactness,
    SurfToVol           = SurfaceArea / (WallArea + RoofArea),
    GlazingXHeight      = GlazingArea * OverallHeight,
    WallXGlazing        = WallArea * GlazingArea,
    CompactnessSquared  = RelCompactness^2,
    HeightSquared       = OverallHeight^2,
    GlazingToWall       = GlazingArea / (WallArea + 1),
    EnvelopeArea        = WallArea + RoofArea,
    WallPerHeight       = WallArea / (OverallHeight + 1)
  )
}

df_train <- engineer(df_train)
df_test  <- engineer(df_test)

cat("Engineered feature count:", ncol(df_train), "\n")
```

`SurfaceArea` is excluded from all models: the EDA confirmed it is a
near-perfect linear combination of `WallArea` and `RoofArea` (r ≈ −0.99 with
`RelCompactness`), causing rank deficiency in the model matrix.

## Scaling

```{r reg-scale}
base_features <- c(
  "RelCompactness", "WallArea", "RoofArea", "OverallHeight",
  "GlazingArea", "GlazingAreaDist",
  "CompactnessXHeight", "WallToRoofRatio", "GlazingXCompactness",
  "SurfToVol", "GlazingXHeight", "WallXGlazing",
  "CompactnessSquared", "GlazingToWall", "EnvelopeArea", "WallPerHeight"
)

preproc_reg <- caret::preProcess(df_train[, base_features],
                                  method = c("center", "scale"))
df_train[, base_features] <- predict(preproc_reg, df_train[, base_features])
df_test[, base_features]  <- predict(preproc_reg, df_test[, base_features])

cat("Scaling complete. Fitted on training set only.\n")
```

## Evaluation Helper

```{r reg-eval-helper}
eval_reg <- function(actual, predicted, model_name) {
  residuals <- actual - predicted
  rmse <- sqrt(mean(residuals^2))
  mae  <- mean(abs(residuals))
  r2   <- 1 - sum(residuals^2) / sum((actual - mean(actual))^2)
  cat(sprintf("%-40s  RMSE = %6.3f  MAE = %6.3f  R² = %.4f\n",
              model_name, rmse, mae, r2))
  list(model_name = model_name, rmse = rmse, mae = mae, r2 = r2)
}

results_HL <- data.frame()
results_CL <- data.frame()
```

## Model 1 — Baseline OLS

The first model uses only the original features (minus `SurfaceArea`) with no
engineering. This establishes a performance floor against which all subsequent
models are compared.

```{r reg-baseline-hl}
formula_base <- HeatingLoad ~ RelCompactness + WallArea + RoofArea +
                              OverallHeight + GlazingArea + GlazingAreaDist +
                              Orientation

lm_base_HL <- lm(formula_base, data = df_train)
cat("=== BASELINE OLS — Heating Load ===\n")
print(summary(lm_base_HL))

pred_base_HL <- predict(lm_base_HL, newdata = df_test)
res_base_HL  <- eval_reg(df_test$HeatingLoad, pred_base_HL, "Baseline OLS")
results_HL <- rbind(results_HL,
  data.frame(Model = "Baseline OLS",
             RMSE = res_base_HL$rmse, MAE = res_base_HL$mae, R2 = res_base_HL$r2))
```

The baseline achieves R²=0.919 and RMSE=2.840 kWh/m². Despite the severe VIFs
documented in the EDA, the overall fit is reasonable because R² aggregates all
predictors — collinearity inflates individual standard errors but does not
prevent the combined signal from being captured. Orientation (p=0.833) is
already near-zero, consistent with its flat LOESS curve in the EDA. This RMSE
of 2.840 is the performance floor all subsequent models are measured against.

## Model 2 — OLS with Interaction Terms

The EDA interaction plot showed a crossing pattern between `OverallHeight` and
`GlazingArea`. Adding this interaction, plus the physically motivated cross terms
from feature engineering, tests whether the model can capture the nonlinear
surface that the scatter plots suggested.

```{r reg-interact-hl}
formula_interact <- HeatingLoad ~ RelCompactness + WallArea + RoofArea +
                                  OverallHeight + GlazingArea + GlazingAreaDist +
                                  Orientation +
                                  CompactnessXHeight + GlazingXHeight +
                                  WallXGlazing + GlazingXCompactness

lm_interact_HL <- lm(formula_interact, data = df_train)
cat("=== OLS WITH INTERACTIONS — Heating Load ===\n")
print(summary(lm_interact_HL))

pred_interact_HL <- predict(lm_interact_HL, newdata = df_test)
res_interact_HL  <- eval_reg(df_test$HeatingLoad, pred_interact_HL,
                              "OLS with Interactions")
results_HL <- rbind(results_HL,
  data.frame(Model = "OLS with Interactions",
             RMSE = res_interact_HL$rmse, MAE = res_interact_HL$mae,
             R2   = res_interact_HL$r2))
```

The interaction model reduces RMSE from 2.840 to 2.470 (−13%) and improves R²
from 0.919 to 0.939, confirming that the EDA crossing pattern reflects real
structure. The `GlazingXHeight` coefficient is −1.20 (p=0.326, not significant),
running counter to the physical expectation. This is not a modelling failure:
`GlazingXHeight`'s signal is absorbed by `CompactnessXHeight` (−60.5, p<0.001)
and `GlazingXCompactness` (+8.3, p<0.001), which together capture the
height-dependent glazing effect more efficiently. The same absorption pattern
appears in the cooling load model where `GlazingXHeight` is also non-significant
(p=0.096).

## Model 3 — Ridge Regression

The severe VIFs established in the EDA (up to 212 for `RoofArea`) make Ridge
regression a natural choice. Ridge shrinks all coefficients toward zero in
proportion to their collinearity, producing stable estimates where OLS cannot.

```{r reg-ridge}
X_train <- as.matrix(df_train[, base_features])
X_test  <- as.matrix(df_test[,  base_features])
y_train_HL <- df_train$HeatingLoad
y_test_HL  <- df_test$HeatingLoad

set.seed(42)
ridge_cv_HL <- cv.glmnet(X_train, y_train_HL, alpha = 0, nfolds = 10)
cat("=== RIDGE REGRESSION — Heating Load ===\n")
cat(sprintf("Optimal lambda (1se): %.4f\n", ridge_cv_HL$lambda.1se))

plot(ridge_cv_HL, main = "Ridge CV — Heating Load")

ridge_full_HL <- glmnet(X_train, y_train_HL, alpha = 0)
plot(ridge_full_HL, xvar = "lambda",
     main = "Ridge Coefficient Path — Heating Load", label = TRUE)

pred_ridge_HL <- predict(ridge_cv_HL, newx = X_test, s = "lambda.1se")
res_ridge_HL  <- eval_reg(y_test_HL, pred_ridge_HL, "Ridge Regression")
results_HL <- rbind(results_HL,
  data.frame(Model = "Ridge Regression",
             RMSE = res_ridge_HL$rmse, MAE = res_ridge_HL$mae, R2 = res_ridge_HL$r2))
```

The coefficient path shows that `OverallHeight`, `GlazingXHeight`, and the
dominant negative-coefficient feature shrink last, confirming their genuine
predictive signal. The cluster converging near zero at low −log(λ) corresponds
to the highly collinear geometric features (`RelCompactness`, `RoofArea`,
`SurfToVol`) whose large OLS coefficients were artefacts of multicollinearity.
Ridge RMSE is 2.546 for HL and 2.888 for CL. For cooling load, Ridge is the
best-performing linear model overall, suggesting that for CL the dominant source
of error is collinearity rather than unmodelled nonlinearity.

## Model 4 — Lasso Regression

```{r reg-lasso}
set.seed(42)
lasso_cv_HL <- cv.glmnet(X_train, y_train_HL, alpha = 1, nfolds = 10)
cat("=== LASSO REGRESSION — Heating Load ===\n")
cat(sprintf("Optimal lambda (1se): %.4f\n", lasso_cv_HL$lambda.1se))
plot(lasso_cv_HL, main = "Lasso CV — Heating Load")

pred_lasso_HL <- predict(lasso_cv_HL, newx = X_test, s = "lambda.1se")
res_lasso_HL  <- eval_reg(y_test_HL, pred_lasso_HL, "Lasso Regression")

lasso_coefs_HL <- coef(lasso_cv_HL, s = "lambda.1se")
selected_HL    <- rownames(lasso_coefs_HL)[lasso_coefs_HL[, 1] != 0]
cat("\nFeatures retained by Lasso:\n")
print(selected_HL)
cat("Features zeroed out:", length(base_features) + 1 - length(selected_HL), "\n")

results_HL <- rbind(results_HL,
  data.frame(Model = "Lasso Regression",
             RMSE = res_lasso_HL$rmse, MAE = res_lasso_HL$mae, R2 = res_lasso_HL$r2))
```

Lasso zeroed 7 of 16 features for heating load and 6 of 16 for cooling load.
The zeroed features for HL include `RelCompactness` and `GlazingArea` — two of
the strongest physical predictors in the baseline OLS. This is not evidence those
variables are unimportant; their signal is fully captured by the interaction and
ratio terms retained instead (`WallToRoofRatio`, `SurfToVol`, `GlazingXHeight`).
`Orientation` is correctly zeroed in both models, matching the EDA finding. The
divergence between HL and CL retained sets is informative: cooling load Lasso
drops `WallXGlazing` and `EnvelopeArea` but keeps `GlazingXCompactness` and
`CompactnessSquared`, pointing to a stronger quadratic compactness effect in the
cooling response.

## Model 5 — Elastic Net

```{r reg-enet}
set.seed(42)
enet_cv_HL <- cv.glmnet(X_train, y_train_HL, alpha = 0.5, nfolds = 10)
cat("=== ELASTIC NET — Heating Load ===\n")
cat(sprintf("Optimal lambda (1se): %.4f\n", enet_cv_HL$lambda.1se))

pred_enet_HL <- predict(enet_cv_HL, newx = X_test, s = "lambda.1se")
res_enet_HL  <- eval_reg(y_test_HL, pred_enet_HL, "Elastic Net (alpha=0.5)")

enet_coefs_HL <- coef(enet_cv_HL, s = "lambda.1se")
selected_enet_HL <- rownames(enet_coefs_HL)[enet_coefs_HL[, 1] != 0]
cat("\nFeatures retained by Elastic Net:\n")
print(selected_enet_HL)

results_HL <- rbind(results_HL,
  data.frame(Model = "Elastic Net",
             RMSE = res_enet_HL$rmse, MAE = res_enet_HL$mae, R2 = res_enet_HL$r2))
```

Elastic Net (RMSE=2.428) marginally outperforms OLS with interactions
(RMSE=2.470) despite retaining 16 of 16 features — the L1 penalty is small
enough (λ=0.0034) that almost no zeroing occurs, and the benefit comes almost
entirely from L2 shrinkage of the collinear geometric terms. This makes Elastic
Net the best linear model for heating load. For cooling load, Ridge wins
(RMSE=2.888) and Elastic Net is second (RMSE=2.932), suggesting the cooling
load benefits more from uniform shrinkage than from sparsity.

## Residual Diagnostics — Best Linear Model

```{r reg-diagnostics}
cat("=== RESIDUAL DIAGNOSTICS — OLS with Interactions ===\n")
par(mfrow = c(2, 2))
plot(lm_interact_HL, which = 1:4)
par(mfrow = c(1, 1))

cat("\nBreusch-Pagan test for heteroscedasticity:\n")
print(lmtest::bptest(lm_interact_HL))

cat("\nAnderson-Darling test on residuals:\n")
print(nortest::ad.test(residuals(lm_interact_HL)))
```

The Breusch-Pagan statistic (BP=268.9, df=11, p<2.2e-16) and Anderson-Darling
statistic (A=8.12, p<2.2e-16) both reject their null hypotheses decisively. The
interaction OLS model has severe heteroscedasticity and non-normal residuals —
neither is surprising given the bimodal response distribution driven by
`OverallHeight`. The residuals fan out differently for low-height and
high-height buildings. This is not a modelling failure: it is a structural
consequence of the two-cluster data-generating process that linear models cannot
fully resolve. It directly motivates the domain-appropriate regression models
in Part 3, which can partition the feature space rather than imposing a global
linear relationship.

## All Models — Cooling Load

```{r reg-cooling}
y_train_CL <- df_train$CoolingLoad
y_test_CL  <- df_test$CoolingLoad

# Baseline OLS
formula_base_CL <- CoolingLoad ~ RelCompactness + WallArea + RoofArea +
                                  OverallHeight + GlazingArea +
                                  GlazingAreaDist + Orientation
lm_base_CL <- lm(formula_base_CL, data = df_train)
pred_base_CL <- predict(lm_base_CL, newdata = df_test)
res_base_CL  <- eval_reg(y_test_CL, pred_base_CL, "Baseline OLS")
results_CL   <- rbind(results_CL,
  data.frame(Model = "Baseline OLS",
             RMSE = res_base_CL$rmse, MAE = res_base_CL$mae, R2 = res_base_CL$r2))

# OLS with interactions
formula_interact_CL <- CoolingLoad ~ RelCompactness + WallArea + RoofArea +
                                      OverallHeight + GlazingArea +
                                      GlazingAreaDist + Orientation +
                                      CompactnessXHeight + GlazingXHeight +
                                      WallXGlazing + GlazingXCompactness
lm_interact_CL <- lm(formula_interact_CL, data = df_train)
pred_interact_CL <- predict(lm_interact_CL, newdata = df_test)
res_interact_CL  <- eval_reg(y_test_CL, pred_interact_CL, "OLS with Interactions")
results_CL <- rbind(results_CL,
  data.frame(Model = "OLS with Interactions",
             RMSE = res_interact_CL$rmse, MAE = res_interact_CL$mae,
             R2   = res_interact_CL$r2))

# Ridge
set.seed(42)
ridge_cv_CL   <- cv.glmnet(X_train, y_train_CL, alpha = 0, nfolds = 10)
pred_ridge_CL <- predict(ridge_cv_CL, newx = X_test, s = "lambda.1se")
res_ridge_CL  <- eval_reg(y_test_CL, pred_ridge_CL, "Ridge Regression")
results_CL <- rbind(results_CL,
  data.frame(Model = "Ridge Regression",
             RMSE = res_ridge_CL$rmse, MAE = res_ridge_CL$mae, R2 = res_ridge_CL$r2))

# Lasso
set.seed(42)
lasso_cv_CL   <- cv.glmnet(X_train, y_train_CL, alpha = 1, nfolds = 10)
pred_lasso_CL <- predict(lasso_cv_CL, newx = X_test, s = "lambda.1se")
res_lasso_CL  <- eval_reg(y_test_CL, pred_lasso_CL, "Lasso Regression")
lasso_coefs_CL <- coef(lasso_cv_CL, s = "lambda.1se")
selected_CL    <- rownames(lasso_coefs_CL)[lasso_coefs_CL[, 1] != 0]
cat("\nLasso features retained (Cooling Load):\n")
print(selected_CL)
results_CL <- rbind(results_CL,
  data.frame(Model = "Lasso Regression",
             RMSE = res_lasso_CL$rmse, MAE = res_lasso_CL$mae, R2 = res_lasso_CL$r2))

# Elastic Net
set.seed(42)
enet_cv_CL   <- cv.glmnet(X_train, y_train_CL, alpha = 0.5, nfolds = 10)
pred_enet_CL <- predict(enet_cv_CL, newx = X_test, s = "lambda.1se")
res_enet_CL  <- eval_reg(y_test_CL, pred_enet_CL, "Elastic Net (alpha=0.5)")
results_CL   <- rbind(results_CL,
  data.frame(Model = "Elastic Net",
             RMSE = res_enet_CL$rmse, MAE = res_enet_CL$mae, R2 = res_enet_CL$r2))
```

## Linear Model Comparison

```{r reg-comparison}
cat("=== HEATING LOAD — Model Comparison ===\n")
print(results_HL %>% arrange(RMSE))

cat("\n=== COOLING LOAD — Model Comparison ===\n")
print(results_CL %>% arrange(RMSE))

p_HL <- ggplot(results_HL %>% arrange(RMSE),
               aes(x = reorder(Model, RMSE), y = RMSE)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = round(RMSE, 3)), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Linear Models — RMSE: Heating Load",
       x = NULL, y = "RMSE (kWh/m²)") +
  theme_minimal()

p_CL <- ggplot(results_CL %>% arrange(RMSE),
               aes(x = reorder(Model, RMSE), y = RMSE)) +
  geom_col(fill = "darkorange", alpha = 0.8) +
  geom_text(aes(label = round(RMSE, 3)), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Linear Models — RMSE: Cooling Load",
       x = NULL, y = "RMSE (kWh/m²)") +
  theme_minimal()

grid.arrange(p_HL, p_CL, ncol = 2)
```

The RMSE reduction from Baseline OLS to the best model is 0.412 kWh/m² for
heating load (14.5% improvement) and 0.273 kWh/m² for cooling load (8.6%
improvement). The smaller gain for cooling load reflects a flatter response
surface that leaves less room for feature engineering to help. **The linear
baselines to carry forward into Part 3 are Elastic Net for heating load
(RMSE=2.428, R²=0.941) and Ridge for cooling load (RMSE=2.888, R²=0.903).**

---

# Part 3 — Domain-Appropriate Regression Models

## Why These Models?

The models in a conventional nonlinear analysis — Random Forest, XGBoost, SVM,
Neural Network — are general-purpose algorithms. They work on almost any tabular
dataset without needing to know what the data represents. That is both their
strength and their limitation.

ENB2012 was generated by EnergyPlus, a thermodynamic simulator, using the
physics of heat transfer through building envelopes. The relationship between
building geometry and energy load is not arbitrary — it follows physical laws.
This creates two opportunities: (1) models suited to small, structured,
simulation-generated datasets can extract more information from 768 rows than
a general-purpose learner, and (2) models can be validated not just by test
error but by checking whether predictions respect physical logic.

**The linear baseline established in Part 2 — Elastic Net (HL RMSE=2.428) and
Ridge (CL RMSE=2.888) — is the target these models are measured against.**

Three models are introduced:

- **GPR with an RBF kernel** — a Bayesian model that treats the dataset as a
  probability distribution over functions; naturally suited to surrogate
  modelling of simulation outputs.
- **GPR with a Laplace (Matérn ½) kernel** — a variant assuming the
  geometry-to-load relationship is less smooth than the RBF kernel assumes,
  allowing systematic comparison of kernel choices. Note: `kernlab` does not
  implement Matérn 5/2 natively; `laplacedot` (Ornstein-Uhlenbeck / Matérn ½)
  is the closest available kernel and is used here with that label.
- **LightGBM** — a modern leaf-wise gradient boosting algorithm that handles
  the discrete, categorical-like structure of ENB2012 more efficiently than
  depth-wise methods.

## Data Setup

```{r dom-reg-setup}
# Reload the post-EDA dataset (features already engineered above;
# reproduced here for standalone section clarity)
df_gpr <- read_excel("ENB2012_post_EDA.xlsx")

df_gpr$Orientation_f     <- factor(df_gpr$Orientation, levels = 2:5,
                                    labels = c("N","E","S","W"))
df_gpr$OverallHeight_f   <- factor(df_gpr$OverallHeight, levels = c(3.5,7.0),
                                    labels = c("Low_3.5m","High_7m"))
df_gpr$GlazingArea_f     <- factor(df_gpr$GlazingArea, levels = c(0,0.1,0.25,0.4),
                                    labels = c("None","Low_10pct","Med_25pct","High_40pct"))
df_gpr$GlazingAreaDist_f <- factor(df_gpr$GlazingAreaDist, levels = 0:5,
                                    labels = c("None","Uniform","N_heavy",
                                               "E_heavy","S_heavy","W_heavy"))

df_gpr <- df_gpr %>%
  mutate(
    CompactnessXHeight  = RelCompactness * OverallHeight,
    WallToRoofRatio     = WallArea / RoofArea,
    GlazingXCompactness = GlazingArea * RelCompactness,
    SurfToVol           = SurfaceArea / (WallArea + RoofArea),
    GlazingXHeight      = GlazingArea * OverallHeight,
    WallXGlazing        = WallArea * GlazingArea,
    CompactnessSquared  = RelCompactness^2,
    HeightSquared       = OverallHeight^2,
    GlazingToWall       = GlazingArea / (WallArea + 1),
    EnvelopeArea        = WallArea + RoofArea,
    WallPerHeight       = WallArea / (OverallHeight + 1),
    group_key           = paste(RelCompactness, WallArea, RoofArea,
                                OverallHeight, GlazingArea, GlazingAreaDist, sep = "_")
  )

set.seed(42)
unique_groups_gpr  <- unique(df_gpr$group_key)
n_train_gpr        <- round(0.80 * length(unique_groups_gpr))
train_groups_gpr   <- sample(unique_groups_gpr, n_train_gpr)

df_train_gpr <- df_gpr %>% filter( group_key %in% train_groups_gpr)
df_test_gpr  <- df_gpr %>% filter(!group_key %in% train_groups_gpr)

cat("Training rows:", nrow(df_train_gpr), "\n")
cat("Test rows    :", nrow(df_test_gpr),  "\n")
cat("Group overlap:", length(intersect(df_train_gpr$group_key,
                                       df_test_gpr$group_key)), "<- zero leakage\n")
```

## Feature Set and Scaling

`SurfaceArea` and `Orientation` are excluded. `Orientation` contributes less
than 7% of within-group variance in the target — confirmed by the EDA and
consistent with near-zero coefficients across all linear models.

```{r dom-reg-features}
features <- c(
  "RelCompactness", "WallArea", "RoofArea",
  "OverallHeight", "GlazingArea", "GlazingAreaDist",
  "CompactnessXHeight", "WallToRoofRatio", "GlazingXCompactness",
  "SurfToVol", "GlazingXHeight", "WallXGlazing",
  "CompactnessSquared", "GlazingToWall", "EnvelopeArea", "WallPerHeight"
)

preproc_gpr <- caret::preProcess(df_train_gpr[, features],
                                  method = c("center", "scale"))
df_train_gpr[, features] <- predict(preproc_gpr, df_train_gpr[, features])
df_test_gpr[, features]  <- predict(preproc_gpr, df_test_gpr[, features])

cat("Scaling complete. Fitted on training set only.\n")
```

## Evaluation Helper

```{r dom-reg-eval-helper}
eval_reg_dom <- function(actual, predicted, model_name) {
  rmse <- sqrt(mean((actual - predicted)^2))
  mae  <- mean(abs(actual - predicted))
  r2   <- 1 - sum((actual - predicted)^2) / sum((actual - mean(actual))^2)
  cat(sprintf("%-35s  RMSE=%.4f  MAE=%.4f  R²=%.4f\n", model_name, rmse, mae, r2))
  data.frame(Model = model_name, RMSE = rmse, MAE = mae, R2 = r2)
}

# Seed results tables with the linear baselines from Part 2 for direct comparison
results_y1 <- data.frame(Model = "Linear Baseline (Elastic Net)",
                          RMSE  = 2.428, MAE = 1.760, R2 = 0.9411)
results_y2 <- data.frame(Model = "Linear Baseline (Ridge)",
                          RMSE  = 2.888, MAE = 2.060, R2 = 0.9026)
```

## Model A — GPR with RBF Kernel

The RBF kernel assumes the function being learned is **infinitely smooth**:
buildings that differ slightly in geometry will differ only slightly in energy
load, with no abrupt jumps anywhere. In mathematical terms:

k(x, x') = σ² × exp(−||x − x'||² / 2ℓ²)

Both σ² (signal variance) and ℓ (length-scale) are estimated from the
training data by maximising the marginal likelihood. Building heat transfer
is governed by continuous physical equations, making the smoothness assumption
physically reasonable for ENB2012.

```{r dom-gpr-rbf-y1}
set.seed(42)
gpr_rbf_y1 <- gausspr(
  x      = as.matrix(df_train_gpr[, features]),
  y      = df_train_gpr$HeatingLoad,
  kernel = "rbfdot",
  kpar   = "automatic",
  var    = 0.01
)
cat("=== GPR (RBF) — Heating Load ===\n")
print(gpr_rbf_y1)

pred_gpr_rbf_y1 <- predict(gpr_rbf_y1,
                             newdata = as.matrix(df_test_gpr[, features]))
results_y1 <- rbind(results_y1,
  eval_reg_dom(df_test_gpr$HeatingLoad, pred_gpr_rbf_y1, "GPR (RBF)"))

gpr_rbf_y1_df <- data.frame(
  Actual    = df_test_gpr$HeatingLoad,
  Predicted = as.vector(pred_gpr_rbf_y1),
  Residual  = df_test_gpr$HeatingLoad - as.vector(pred_gpr_rbf_y1)
)
p_gpr_rbf_y1 <- ggplot(gpr_rbf_y1_df, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.6, colour = "steelblue", size = 2) +
  geom_abline(slope = 1, intercept = 0, colour = "black",
              linetype = "dashed", linewidth = 0.8) +
  labs(title    = "GPR (RBF) — Actual vs Predicted: Heating Load",
       subtitle = "Points on the dashed line are perfect predictions",
       x = "Actual HeatingLoad (kWh/m²)", y = "Predicted HeatingLoad (kWh/m²)") +
  theme_minimal()
print(p_gpr_rbf_y1)
```

```{r dom-gpr-rbf-y2}
set.seed(42)
gpr_rbf_y2 <- gausspr(
  x      = as.matrix(df_train_gpr[, features]),
  y      = df_train_gpr$CoolingLoad,
  kernel = "rbfdot",
  kpar   = "automatic",
  var    = 0.01
)
pred_gpr_rbf_y2 <- predict(gpr_rbf_y2,
                             newdata = as.matrix(df_test_gpr[, features]))
results_y2 <- rbind(results_y2,
  eval_reg_dom(df_test_gpr$CoolingLoad, pred_gpr_rbf_y2, "GPR (RBF)"))
```

## Model B — GPR with Laplace (Matérn ½) Kernel

The Laplace kernel (Ornstein-Uhlenbeck / Matérn ν=1/2) assumes the underlying
function is continuous but not differentiable — allowing for sharper transitions
in the geometry-to-load relationship than the RBF kernel. ENB2012's features
include discrete variables (`GlazingArea` takes only four values; `OverallHeight`
only two), and the EnergyPlus outputs jump discontinuously between these levels
rather than varying smoothly. The less-smooth Laplace kernel may handle these
step-like patterns better than the infinitely smooth RBF kernel.

Comparing the two models' RMSE and residual patterns directly tells us which
smoothness assumption better matches the actual structure of the data.

```{r dom-gpr-laplace-y1}
set.seed(42)
gpr_mat_y1 <- gausspr(
  x      = as.matrix(df_train_gpr[, features]),
  y      = df_train_gpr$HeatingLoad,
  kernel = "laplacedot",
  kpar   = "automatic",
  var    = 0.01
)
cat("=== GPR (Laplace / Matern 1/2) — Heating Load ===\n")
print(gpr_mat_y1)

pred_gpr_mat_y1 <- predict(gpr_mat_y1,
                             newdata = as.matrix(df_test_gpr[, features]))
results_y1 <- rbind(results_y1,
  eval_reg_dom(df_test_gpr$HeatingLoad, pred_gpr_mat_y1, "GPR (Laplace)"))

gpr_mat_y1_df <- data.frame(
  Actual    = df_test_gpr$HeatingLoad,
  Predicted = as.vector(pred_gpr_mat_y1),
  Residual  = df_test_gpr$HeatingLoad - as.vector(pred_gpr_mat_y1)
)
p_gpr_mat_y1 <- ggplot(gpr_mat_y1_df, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.6, colour = "steelblue", size = 2) +
  geom_abline(slope = 1, intercept = 0, colour = "black",
              linetype = "dashed", linewidth = 0.8) +
  labs(title    = "GPR (Laplace) — Actual vs Predicted: Heating Load",
       subtitle = "Points on the dashed line are perfect predictions",
       x = "Actual HeatingLoad (kWh/m²)", y = "Predicted HeatingLoad (kWh/m²)") +
  theme_minimal()
print(p_gpr_mat_y1)
```

```{r dom-gpr-laplace-y2}
set.seed(42)
gpr_mat_y2 <- gausspr(
  x      = as.matrix(df_train_gpr[, features]),
  y      = df_train_gpr$CoolingLoad,
  kernel = "laplacedot",
  kpar   = "automatic",
  var    = 0.01
)
pred_gpr_mat_y2 <- predict(gpr_mat_y2,
                             newdata = as.matrix(df_test_gpr[, features]))
results_y2 <- rbind(results_y2,
  eval_reg_dom(df_test_gpr$CoolingLoad, pred_gpr_mat_y2, "GPR (Laplace)"))
```

## GPR Kernel Comparison

```{r dom-gpr-kernel-compare}
kernel_compare_y1 <- bind_rows(
  data.frame(Actual    = df_test_gpr$HeatingLoad,
             Predicted = as.vector(pred_gpr_rbf_y1), Kernel = "RBF"),
  data.frame(Actual    = df_test_gpr$HeatingLoad,
             Predicted = as.vector(pred_gpr_mat_y1), Kernel = "Laplace")
)
p_kernel_y1 <- ggplot(kernel_compare_y1, aes(x = Actual, y = Predicted,
                                               colour = Kernel)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_abline(slope = 1, intercept = 0, colour = "black",
              linetype = "dashed", linewidth = 0.8) +
  scale_colour_manual(values = c("RBF" = "steelblue", "Laplace" = "tomato")) +
  labs(title    = "GPR Kernel Comparison — Heating Load",
       subtitle = "Divergence shows where smoothness assumptions differ",
       x = "Actual HeatingLoad (kWh/m²)", y = "Predicted HeatingLoad (kWh/m²)") +
  theme_minimal()
print(p_kernel_y1)
```

## Model C — LightGBM

LightGBM grows trees **leaf-wise**: at each step it finds the single leaf whose
split would most reduce the loss and expands only that leaf. This produces
deeper, more asymmetric trees that fit complex patterns more efficiently than
depth-wise methods. For ENB2012 specifically, LightGBM's histogram-based
split-finding algorithm naturally groups the few discrete feature values into
bins before finding splits, aligning well with the dataset's structure.

```{r dom-lgbm-y1}
X_train_lgbm <- as.matrix(df_train_gpr[, features])
X_test_lgbm  <- as.matrix(df_test_gpr[,  features])

dtrain_y1 <- lgb.Dataset(data  = X_train_lgbm,
                           label = df_train_gpr$HeatingLoad)

lgbm_grid <- expand.grid(
  num_leaves    = c(15, 31),
  learning_rate = c(0.05, 0.10)
)

lgbm_tune_y1 <- purrr::pmap_dfr(lgbm_grid, function(num_leaves, learning_rate) {
  params <- list(
    objective = "regression", metric = "rmse",
    num_leaves = num_leaves, learning_rate = learning_rate,
    min_data_in_leaf = 5, feature_fraction = 0.8,
    bagging_fraction = 0.8, bagging_freq = 5, verbose = -1
  )
  set.seed(42)
  cv_result <- lgb.cv(params = params, data = dtrain_y1, nfold = 5,
                       nrounds = 500, early_stopping_rounds = 30, verbose = -1)
  best_rmse <- min(unlist(cv_result$record_evals$valid$rmse$eval))
  best_iter <- which.min(unlist(cv_result$record_evals$valid$rmse$eval))
  data.frame(num_leaves = num_leaves, learning_rate = learning_rate,
             best_rmse_cv = round(best_rmse, 5), best_nrounds = best_iter)
})

cat("=== LightGBM Tuning Grid — Heating Load ===\n")
print(lgbm_tune_y1)

best_row_y1 <- lgbm_tune_y1[which.min(lgbm_tune_y1$best_rmse_cv), ]
cat("\nSelected configuration:\n")
print(best_row_y1)

best_params_y1 <- list(
  objective = "regression", metric = "rmse",
  num_leaves = best_row_y1$num_leaves, learning_rate = best_row_y1$learning_rate,
  min_data_in_leaf = 5, feature_fraction = 0.8,
  bagging_fraction = 0.8, bagging_freq = 5, verbose = -1
)

set.seed(42)
lgbm_y1 <- lgb.train(params  = best_params_y1,
                       data    = dtrain_y1,
                       nrounds = best_row_y1$best_nrounds)

pred_lgbm_y1 <- predict(lgbm_y1, X_test_lgbm)
results_y1 <- rbind(results_y1,
  eval_reg_dom(df_test_gpr$HeatingLoad, pred_lgbm_y1, "LightGBM"))

lgbm_imp_y1 <- lgb.importance(lgbm_y1, percentage = TRUE)
p_lgbm_imp_y1 <- lgbm_imp_y1 %>% head(12) %>%
  ggplot(aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = paste0(round(Gain * 100, 1), "%")),
            hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title    = "LightGBM Feature Importance (Gain) — Heating Load",
       subtitle = "Gain = total loss reduction attributed to each feature's splits",
       x = "Feature", y = "Gain (proportion of total)") +
  theme_minimal()
print(p_lgbm_imp_y1)
```

```{r dom-lgbm-y2}
dtrain_y2 <- lgb.Dataset(data  = X_train_lgbm,
                           label = df_train_gpr$CoolingLoad)

lgbm_tune_y2 <- purrr::pmap_dfr(lgbm_grid, function(num_leaves, learning_rate) {
  params <- list(
    objective = "regression", metric = "rmse",
    num_leaves = num_leaves, learning_rate = learning_rate,
    min_data_in_leaf = 5, feature_fraction = 0.8,
    bagging_fraction = 0.8, bagging_freq = 5, verbose = -1
  )
  set.seed(42)
  cv_result <- lgb.cv(params = params, data = dtrain_y2, nfold = 5,
                       nrounds = 500, early_stopping_rounds = 30, verbose = -1)
  best_rmse <- min(unlist(cv_result$record_evals$valid$rmse$eval))
  best_iter <- which.min(unlist(cv_result$record_evals$valid$rmse$eval))
  data.frame(num_leaves = num_leaves, learning_rate = learning_rate,
             best_rmse_cv = round(best_rmse, 5), best_nrounds = best_iter)
})

best_row_y2 <- lgbm_tune_y2[which.min(lgbm_tune_y2$best_rmse_cv), ]
best_params_y2 <- list(
  objective = "regression", metric = "rmse",
  num_leaves = best_row_y2$num_leaves, learning_rate = best_row_y2$learning_rate,
  min_data_in_leaf = 5, feature_fraction = 0.8,
  bagging_fraction = 0.8, bagging_freq = 5, verbose = -1
)
set.seed(42)
lgbm_y2 <- lgb.train(params = best_params_y2, data = dtrain_y2,
                       nrounds = best_row_y2$best_nrounds)
pred_lgbm_y2 <- predict(lgbm_y2, X_test_lgbm)
results_y2 <- rbind(results_y2,
  eval_reg_dom(df_test_gpr$CoolingLoad, pred_lgbm_y2, "LightGBM"))
```

## Physics-Consistency Check

A model can have low RMSE and still be physically wrong. This check evaluates
**monotonic ordering**: does each model preserve the direction of three physical
relationships that the EnergyPlus simulator encodes?

1. Higher glazing → higher heating load (more surface through which heat escapes)
2. Higher compactness → lower heating load (less envelope area per unit volume)
3. Greater height → higher heating load (more external surface area)

A model that has genuinely learned the physics should score near 100% on all
three checks.

```{r dom-physics}
check_physics <- function(model_obj, model_name, model_type = "gausspr") {
  predict_fn <- function(newdata_mat) {
    if (model_type == "gausspr") as.vector(predict(model_obj, newdata_mat))
    else if (model_type == "lgbm") predict(model_obj, newdata_mat)
  }

  base_grid <- expand.grid(
    RelCompactness  = c(0.69, 0.82),
    WallArea        = c(269.5, 343),
    RoofArea        = c(122.5, 147),
    OverallHeight   = c(3.5, 7.0),
    GlazingArea     = c(0.10, 0.25, 0.40),
    GlazingAreaDist = c(1, 3)
  ) %>% mutate(
    CompactnessXHeight  = RelCompactness * OverallHeight,
    WallToRoofRatio     = WallArea / RoofArea,
    GlazingXCompactness = GlazingArea * RelCompactness,
    SurfToVol           = (WallArea + 2 * RoofArea) / (WallArea + RoofArea),
    GlazingXHeight      = GlazingArea * OverallHeight,
    WallXGlazing        = WallArea * GlazingArea,
    CompactnessSquared  = RelCompactness^2,
    HeightSquared       = OverallHeight^2,
    GlazingToWall       = GlazingArea / (WallArea + 1),
    EnvelopeArea        = WallArea + RoofArea,
    WallPerHeight       = WallArea / (OverallHeight + 1)
  )
  base_grid[, features] <- predict(preproc_gpr, base_grid[, features])
  base_grid$pred_HL     <- predict_fn(as.matrix(base_grid[, features]))

  glaz_score <- mean(base_grid %>%
    group_by(RelCompactness, WallArea, RoofArea, OverallHeight, GlazingAreaDist) %>%
    arrange(GlazingArea) %>%
    summarise(correct = all(diff(pred_HL) > 0), .groups = "drop") %>%
    pull(correct)) * 100

  comp_score <- mean(base_grid %>%
    group_by(WallArea, RoofArea, OverallHeight, GlazingArea, GlazingAreaDist) %>%
    arrange(RelCompactness) %>%
    summarise(correct = all(diff(pred_HL) < 0), .groups = "drop") %>%
    pull(correct)) * 100

  ht_score <- mean(base_grid %>%
    group_by(RelCompactness, WallArea, RoofArea, GlazingArea, GlazingAreaDist) %>%
    arrange(OverallHeight) %>%
    summarise(correct = all(diff(pred_HL) > 0), .groups = "drop") %>%
    pull(correct)) * 100

  cat(sprintf("%-25s  Glazing: %5.1f%%  Compactness: %5.1f%%  Height: %5.1f%%\n",
              model_name, glaz_score, comp_score, ht_score))
  data.frame(Model = model_name,
             Glazing_pct = glaz_score, Compact_pct = comp_score,
             Height_pct  = ht_score)
}

physics_results <- bind_rows(
  check_physics(gpr_rbf_y1, "GPR (RBF)",      "gausspr"),
  check_physics(gpr_mat_y1, "GPR (Laplace)",   "gausspr"),
  check_physics(lgbm_y1,    "LightGBM",        "lgbm")
)

p_physics <- physics_results %>%
  pivot_longer(ends_with("_pct"), names_to = "Check", values_to = "Score") %>%
  mutate(Check = recode(Check,
    "Glazing_pct" = "Glazing → HL",
    "Compact_pct" = "Compactness → HL",
    "Height_pct"  = "Height → HL")) %>%
  ggplot(aes(x = Model, y = Score, fill = Check)) +
  geom_col(position = "dodge", alpha = 0.85) +
  geom_text(aes(label = paste0(round(Score, 0), "%")),
            position = position_dodge(width = 0.9), vjust = -0.4, size = 3) +
  geom_hline(yintercept = 100, linetype = "dashed",
             colour = "darkgreen", linewidth = 0.7) +
  scale_fill_manual(values = c("steelblue","tomato","goldenrod")) +
  scale_y_continuous(limits = c(0, 115)) +
  labs(title    = "Physics-Consistency Check — Heating Load",
       subtitle = "% of building pairs where model predicts the physically correct direction",
       x = "Model", y = "Correct ordering (%)", fill = "Physical rule") +
  theme_minimal()
print(p_physics)
```

## Domain Regression Model Comparison

```{r dom-reg-comparison}
results_y1 <- results_y1 %>% arrange(RMSE)
results_y2 <- results_y2 %>% arrange(RMSE)

cat("=== HEATING LOAD — Domain Regression Comparison ===\n")
print(results_y1, row.names = FALSE)

cat("\n=== COOLING LOAD — Domain Regression Comparison ===\n")
print(results_y2, row.names = FALSE)

p_dom_y1 <- ggplot(results_y1, aes(x = reorder(Model, RMSE), y = RMSE,
                                     fill = grepl("Baseline", Model))) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = round(RMSE, 3)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "grey60", "FALSE" = "steelblue"),
                    guide = "none") +
  labs(title    = "Domain Regression — RMSE: Heating Load",
       subtitle = "Grey bar = linear baseline from Part 2",
       x = "Model", y = "RMSE (kWh/m²)") +
  theme_minimal()

p_dom_y2 <- ggplot(results_y2, aes(x = reorder(Model, RMSE), y = RMSE,
                                     fill = grepl("Baseline", Model))) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = round(RMSE, 3)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "grey60", "FALSE" = "darkorange"),
                    guide = "none") +
  labs(title    = "Domain Regression — RMSE: Cooling Load",
       subtitle = "Grey bar = linear baseline from Part 2",
       x = "Model", y = "RMSE (kWh/m²)") +
  theme_minimal()

grid.arrange(p_dom_y1, p_dom_y2, ncol = 2)
```

---

# Part 4 — Domain-Appropriate Classification

## Why These Models?

The general-purpose classifiers from a conventional analysis treat all
misclassifications as equally bad. In reality, predicting "High" when the true
class is "Low" is far more serious than predicting "Medium" for a true-Low
building — a building engineer would make very different retrofit decisions in
each case. The three models here are chosen specifically because they engage
with this ordinal structure:

- **The Standards-Based Classifier** applies EU Energy Performance of Buildings
  Directive thresholds without any training. It gives us a domain baseline: how
  well does the rule a practitioner would apply compare to a learned model?
- **The Ordinal Random Forest** (implemented via `ranger` probability forest)
  is trained to minimise the Ranked Probability Score, penalising large-order
  errors more than small ones.
- **Gaussian Naive Bayes with Physical Priors** partially informs feature
  distributions from heat transfer principles rather than estimating them
  purely from data.

## Constructing the Classification Target

```{r clf-targets}
# Reload and re-engineer (classification shares the same feature pipeline)
df_clf <- read_excel("ENB2012_post_EDA.xlsx")

df_clf$Orientation_f     <- factor(df_clf$Orientation, levels = 2:5,
                                    labels = c("N","E","S","W"))
df_clf$OverallHeight_f   <- factor(df_clf$OverallHeight, levels = c(3.5,7.0),
                                    labels = c("Low_3.5m","High_7m"))
df_clf$GlazingArea_f     <- factor(df_clf$GlazingArea, levels = c(0,0.1,0.25,0.4),
                                    labels = c("None","Low_10pct","Med_25pct","High_40pct"))
df_clf$GlazingAreaDist_f <- factor(df_clf$GlazingAreaDist, levels = 0:5,
                                    labels = c("None","Uniform","N_heavy",
                                               "E_heavy","S_heavy","W_heavy"))

df_clf <- df_clf %>%
  mutate(
    CompactnessXHeight  = RelCompactness * OverallHeight,
    WallToRoofRatio     = WallArea / RoofArea,
    GlazingXCompactness = GlazingArea * RelCompactness,
    SurfToVol           = SurfaceArea / (WallArea + RoofArea),
    GlazingXHeight      = GlazingArea * OverallHeight,
    WallXGlazing        = WallArea * GlazingArea,
    CompactnessSquared  = RelCompactness^2,
    HeightSquared       = OverallHeight^2,
    GlazingToWall       = GlazingArea / (WallArea + 1),
    EnvelopeArea        = WallArea + RoofArea,
    WallPerHeight       = WallArea / (OverallHeight + 1),
    group_key           = paste(RelCompactness, WallArea, RoofArea,
                                OverallHeight, GlazingArea, GlazingAreaDist, sep = "_")
  )

# Tertile boundaries computed from the full dataset before splitting
hl_breaks <- quantile(df_clf$HeatingLoad,  probs = c(0, 1/3, 2/3, 1))
cl_breaks <- quantile(df_clf$CoolingLoad,  probs = c(0, 1/3, 2/3, 1))

df_clf <- df_clf %>%
  mutate(
    HL_class     = cut(HeatingLoad, breaks = hl_breaks,
                       labels = c("Low","Medium","High"), include.lowest = TRUE),
    CL_class     = cut(CoolingLoad, breaks = cl_breaks,
                       labels = c("Low","Medium","High"), include.lowest = TRUE),
    HL_class_ord = factor(HL_class, levels = c("Low","Medium","High"), ordered = TRUE),
    CL_class_ord = factor(CL_class, levels = c("Low","Medium","High"), ordered = TRUE)
  )

cat("HeatingLoad tertile boundaries:\n")
cat(sprintf("  Low / Medium boundary : %.2f kWh/m²\n", hl_breaks[2]))
cat(sprintf("  Medium / High boundary: %.2f kWh/m²\n", hl_breaks[3]))
cat("  (EU A/B threshold: 15 kWh/m²,  EU B/C threshold: 30 kWh/m²)\n\n")

cat("Class distribution — Heating Load:\n")
print(table(df_clf$HL_class))
cat("\nClass distribution — Cooling Load:\n")
print(table(df_clf$CL_class))
```

The tertile boundaries land very close to the EU Energy Performance of
Buildings Directive thresholds (Low/Medium ≈ 14.5 vs EU 15 kWh/m²;
Medium/High ≈ 28.6 vs EU 30 kWh/m²). This near-alignment is not coincidental —
EnergyPlus simulates realistic performance ranges that naturally span the same
territory energy standards were written to cover. It makes the standards-based
classifier a meaningful comparison rather than a trivial one.

## Group-Aware Split

```{r clf-split}
set.seed(42)
unique_groups_clf <- unique(df_clf$group_key)
n_train_clf       <- round(0.80 * length(unique_groups_clf))
train_groups_clf  <- sample(unique_groups_clf, n_train_clf)

df_train_clf <- df_clf %>% filter( group_key %in% train_groups_clf)
df_test_clf  <- df_clf %>% filter(!group_key %in% train_groups_clf)

cat("Training rows:", nrow(df_train_clf), "\n")
cat("Test rows    :", nrow(df_test_clf),  "\n")
cat("Group overlap:", length(intersect(df_train_clf$group_key,
                                       df_test_clf$group_key)), "\n")
```

## Feature Set and Scaling

`SurfaceArea` and `Orientation` are excluded. Orientation contributes less
than 7% of within-group variance in the target — confirmed by the EDA and
consistent with its near-zero coefficients across all linear models. Scaling
is fitted on training data only.

```{r clf-scale}
preproc_clf <- caret::preProcess(df_train_clf[, features],
                                  method = c("center", "scale"))
df_train_clf[, features] <- predict(preproc_clf, df_train_clf[, features])
df_test_clf[, features]  <- predict(preproc_clf, df_test_clf[, features])

# Assemble model datasets
train_HL <- df_train_clf[, c(features, "HL_class")]
test_HL  <- df_test_clf[,  c(features, "HL_class")]
train_CL <- df_train_clf[, c(features, "CL_class")]
test_CL  <- df_test_clf[,  c(features, "CL_class")]

# Ordered versions for ranger and RPS
train_HL_ord <- df_train_clf[, c(features, "HL_class_ord")]
test_HL_ord  <- df_test_clf[,  c(features, "HL_class_ord")]
train_CL_ord <- df_train_clf[, c(features, "CL_class_ord")]
test_CL_ord  <- df_test_clf[,  c(features, "CL_class_ord")]

cat("Class distribution in training set — HL:\n")
print(table(train_HL$HL_class))
```

## Evaluation Helpers

```{r clf-eval-helpers}
eval_clf <- function(actual, predicted, model_name) {
  acc <- mean(actual == predicted)
  cm  <- table(Actual = actual, Predicted = predicted)
  cat(sprintf("%-35s  Accuracy = %.4f\n", model_name, acc))
  list(model_name = model_name, accuracy = acc, confusion = cm)
}

ranked_prob_score <- function(prob_matrix, actual_ordered) {
  classes <- levels(actual_ordered)
  k <- length(classes)
  n <- length(actual_ordered)
  rps_vals <- sapply(seq_len(n), function(i) {
    true_class <- which(classes == as.character(actual_ordered[i]))
    obs_cdf    <- cumsum(c(rep(0, true_class - 1), 1, rep(0, k - true_class)))
    obs_cdf    <- obs_cdf[-k]
    pred_cdf   <- cumsum(prob_matrix[i, ])[-k]
    mean((pred_cdf - obs_cdf)^2)
  })
  mean(rps_vals)
}

results_HL_clf <- data.frame()
results_CL_clf <- data.frame()
```

## Model 1 — Standards-Based Classifier (Domain Baseline)

Before training any model, a building energy consultant classifying buildings by
heating load would apply the thresholds from the energy performance standard that
governs their work. The gap between this classifier's accuracy and the accuracy
of the learned models is the empirical value that machine learning adds over
domain expertise alone.

```{r clf-standards-hl}
standards_pred_HL <- case_when(
  df_test_clf$HeatingLoad < 15 ~ "Low",
  df_test_clf$HeatingLoad < 30 ~ "Medium",
  TRUE                         ~ "High"
) %>% factor(levels = c("Low", "Medium", "High"))

cat("=== STANDARDS-BASED CLASSIFIER — Heating Load ===\n")
cat("EU thresholds: Low < 15, 15 ≤ Medium < 30, High ≥ 30 kWh/m²\n\n")

std_HL_result <- eval_clf(test_HL$HL_class, standards_pred_HL,
                           "Standards-Based (EU)")
cat("\nConfusion matrix:\n")
print(std_HL_result$confusion)

results_HL_clf <- rbind(results_HL_clf,
  data.frame(Model = "Standards-Based (EU)", Accuracy = std_HL_result$accuracy))

p_std_HL <- data.frame(
  HeatingLoad     = df_test_clf$HeatingLoad,
  True_Class      = test_HL$HL_class,
  Predicted_Class = standards_pred_HL,
  Correct         = (test_HL$HL_class == standards_pred_HL)
) %>%
  ggplot(aes(x = HeatingLoad, y = True_Class, colour = Correct, shape = Correct)) +
  geom_jitter(height = 0.15, alpha = 0.7, size = 2.5) +
  geom_vline(xintercept = 15, linetype = "dashed", colour = "steelblue", linewidth = 0.8) +
  geom_vline(xintercept = 30, linetype = "dashed", colour = "tomato",   linewidth = 0.8) +
  scale_colour_manual(values = c("TRUE" = "darkgreen", "FALSE" = "red"),
                      name = "Correctly\nclassified") +
  scale_shape_manual(values  = c("TRUE" = 16, "FALSE" = 4),
                     name = "Correctly\nclassified") +
  annotate("text", x = 13, y = 3.4, label = "EU A/B\n(15)", size = 3, colour = "steelblue") +
  annotate("text", x = 28, y = 3.4, label = "EU B/C\n(30)", size = 3, colour = "tomato") +
  labs(title    = "Standards-Based Classifier — Heating Load",
       subtitle = "Dashed lines show EU energy class boundaries",
       x = "Actual HeatingLoad (kWh/m²)", y = "True Class") +
  theme_minimal()
print(p_std_HL)
```

```{r clf-standards-cl}
standards_pred_CL <- case_when(
  df_test_clf$CoolingLoad < 15 ~ "Low",
  df_test_clf$CoolingLoad < 30 ~ "Medium",
  TRUE                         ~ "High"
) %>% factor(levels = c("Low", "Medium", "High"))

cat("=== STANDARDS-BASED CLASSIFIER — Cooling Load ===\n")
std_CL_result <- eval_clf(test_CL$CL_class, standards_pred_CL, "Standards-Based (EU)")
cat("\nConfusion matrix:\n")
print(std_CL_result$confusion)

results_CL_clf <- rbind(results_CL_clf,
  data.frame(Model = "Standards-Based (EU)", Accuracy = std_CL_result$accuracy))
```

## Model 2 — Ordinal Random Forest

Standard classification models treat the three energy classes as unrelated
categories: a two-step error (Low → High) costs the same as a one-step error
(Low → Medium). A building engineer acting on these predictions would make very
different decisions depending on error direction, so this symmetry is a problem.

The `ranger` probability forest below is trained to return class probabilities,
from which the Ranked Probability Score (RPS) is computed. RPS penalises the
cumulative distance between the predicted probability distribution and the true
class, so two-step errors carry a higher penalty than one-step errors.

```{r clf-orf-hl}
cat("=== ORDINAL RANDOM FOREST — Heating Load ===\n")
cat("Training model...\n")

train_HL_ranger <- train_HL_ord
train_HL_ranger$HL_class_ord <- factor(
  as.character(train_HL_ord$HL_class_ord),
  levels = c("Low","Medium","High"), ordered = FALSE
)
test_HL_ranger <- test_HL_ord
test_HL_ranger$HL_class_ord <- factor(
  as.character(test_HL_ord$HL_class_ord),
  levels = c("Low","Medium","High"), ordered = FALSE
)

set.seed(42)
orf_y1 <- ranger(
  dependent.variable.name = "HL_class_ord",
  data                    = train_HL_ranger,
  num.trees               = 500,
  mtry                    = floor(sqrt(length(features))),
  min.node.size           = 5,
  probability             = TRUE,
  importance              = "impurity",
  num.threads             = parallel::detectCores() - 1
)

pred_prob_orf_y1  <- predict(orf_y1, data = test_HL_ranger)$predictions
pred_prob_orf_y1  <- pred_prob_orf_y1[, c("Low","Medium","High")]
pred_class_orf_y1 <- factor(
  colnames(pred_prob_orf_y1)[apply(pred_prob_orf_y1, 1, which.max)],
  levels = c("Low","Medium","High")
)

orf_HL_result <- eval_clf(test_HL$HL_class, pred_class_orf_y1, "Ordinal Random Forest")
cat("\nConfusion matrix:\n")
print(orf_HL_result$confusion)

rps_orf_HL <- ranked_prob_score(
  pred_prob_orf_y1,
  factor(test_HL$HL_class, levels = c("Low","Medium","High"), ordered = TRUE)
)
cat(sprintf("\nRanked Probability Score: %.5f\n", rps_orf_HL))
cat("(Lower is better; RPS = 0 means every building classified perfectly)\n")

results_HL_clf <- rbind(results_HL_clf,
  data.frame(Model = "Ordinal Random Forest", Accuracy = orf_HL_result$accuracy))

# Variable importance
orf_imp_y1 <- data.frame(
  Feature    = names(orf_y1$variable.importance),
  Importance = orf_y1$variable.importance
) %>% arrange(desc(Importance)) %>% head(12)

ggplot(orf_imp_y1, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = round(Importance, 1)), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title    = "Ordinal Random Forest — Variable Importance: Heating Load",
       subtitle = "Mean decrease in node impurity across all 500 trees",
       x = "Feature", y = "Importance") +
  theme_minimal()

# Confusion matrix heatmap
as.data.frame(orf_HL_result$confusion) %>% rename(Count = Freq) %>%
  ggplot(aes(x = Predicted, y = Actual, fill = Count)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Count), size = 5, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Ordinal Random Forest — Confusion Matrix: Heating Load",
       x = "Predicted Class", y = "Actual Class") +
  theme_minimal()
```

```{r clf-orf-cl}
train_CL_ranger <- train_CL_ord
train_CL_ranger$CL_class_ord <- factor(
  as.character(train_CL_ord$CL_class_ord),
  levels = c("Low","Medium","High"), ordered = FALSE
)
test_CL_ranger <- test_CL_ord
test_CL_ranger$CL_class_ord <- factor(
  as.character(test_CL_ord$CL_class_ord),
  levels = c("Low","Medium","High"), ordered = FALSE
)

set.seed(42)
orf_y2 <- ranger(
  dependent.variable.name = "CL_class_ord",
  data                    = train_CL_ranger,
  num.trees               = 500,
  mtry                    = floor(sqrt(length(features))),
  min.node.size           = 5,
  probability             = TRUE,
  importance              = "impurity",
  num.threads             = parallel::detectCores() - 1
)

pred_prob_orf_y2  <- predict(orf_y2, data = test_CL_ranger)$predictions
pred_prob_orf_y2  <- pred_prob_orf_y2[, c("Low","Medium","High")]
pred_class_orf_y2 <- factor(
  colnames(pred_prob_orf_y2)[apply(pred_prob_orf_y2, 1, which.max)],
  levels = c("Low","Medium","High")
)

orf_CL_result <- eval_clf(test_CL$CL_class, pred_class_orf_y2, "Ordinal Random Forest")
cat("\nConfusion matrix:\n")
print(orf_CL_result$confusion)

rps_orf_CL <- ranked_prob_score(
  pred_prob_orf_y2,
  factor(test_CL$CL_class, levels = c("Low","Medium","High"), ordered = TRUE)
)
cat(sprintf("\nRanked Probability Score: %.5f\n", rps_orf_CL))

results_CL_clf <- rbind(results_CL_clf,
  data.frame(Model = "Ordinal Random Forest", Accuracy = orf_CL_result$accuracy))
```

## Model 3 — Gaussian Naive Bayes with Physical Priors

Naive Bayes estimates a mean and standard deviation for each feature within each
energy class, then classifies by maximising the class posterior probability. The
standard version estimates all parameters from data. The prior-blended version
adjusts the class-conditional means for key features using physical reasoning:
buildings in the Low energy class should tend toward high compactness (less
envelope per unit volume), low glazing area (less conductive heat loss), and
smaller height. The blend weight (30% prior, 70% data) keeps the model
data-driven while preventing physically implausible patterns when data is limited.

```{r clf-nb-hl}
# ── Step 1: fit standard Gaussian Naive Bayes ──────────────────────────────────
set.seed(42)
gnb_y1 <- gaussian_naive_bayes(
  x = as.matrix(train_HL[, features]),
  y = train_HL$HL_class
)

# ── Step 2: define physical prior means (in standardised feature space) ────────
# Low HL  → high compactness (+), low glazing (−), low height (−)
# High HL → low compactness (−), high glazing (+), high height (+)
physical_prior <- list(
  Low    = list(RelCompactness=0.5,  GlazingArea=-0.5, OverallHeight=-0.5,
                CompactnessXHeight=0.3, WallXGlazing=-0.4),
  Medium = list(RelCompactness=0.0,  GlazingArea=0.0,  OverallHeight=0.0,
                CompactnessXHeight=0.0, WallXGlazing=0.0),
  High   = list(RelCompactness=-0.5, GlazingArea=0.5,  OverallHeight=0.5,
                CompactnessXHeight=-0.3, WallXGlazing=0.4)
)
prior_weight <- 0.3
blend_means  <- function(data_mean, prior_mean, w = prior_weight)
  (1 - w) * data_mean + w * prior_mean

# ── Step 3: floor SDs to prevent near-zero instability ────────────────────────
gnb_means_y1 <- gnb_y1$params$mu
gnb_sds_y1   <- gnb_y1$params$sd

for (feat in colnames(gnb_sds_y1)) {
  fallback <- max(sd(train_HL[[feat]], na.rm = TRUE) * 0.5, 0.1)
  for (cls in rownames(gnb_sds_y1)) {
    val <- gnb_sds_y1[cls, feat]
    if (is.nan(val) || is.na(val) || val < 0.1) gnb_sds_y1[cls, feat] <- fallback
  }
}

# ── Step 4: blend data means with physical priors for key features ─────────────
prior_feats       <- c("RelCompactness","GlazingArea","OverallHeight",
                        "CompactnessXHeight","WallXGlazing")
gnb_means_std_y1  <- gnb_means_y1   # unblended copy for standard NB

for (feat in prior_feats) {
  if (!feat %in% colnames(gnb_means_y1)) next
  for (cls in c("Low","Medium","High")) {
    pv <- physical_prior[[cls]][[feat]]
    if (!is.null(pv))
      gnb_means_y1[cls, feat] <- blend_means(gnb_means_y1[cls, feat], pv)
  }
}

# ── Step 5: manual log-likelihood prediction with blended parameters ───────────
predict_gnb_blended <- function(x_mat, means_mat, sds_mat, classes) {
  log_liks <- sapply(classes, function(cls) {
    mu  <- means_mat[cls, ]
    sig <- sds_mat[cls, ]
    rowSums(mapply(function(x_col, m, s) dnorm(x_col, mean = m, sd = s, log = TRUE),
                   as.data.frame(x_mat), mu, sig))
  })
  classes[apply(log_liks, 1, which.max)]
}

x_test_mat  <- as.matrix(test_HL[, features])
classes_vec <- c("Low","Medium","High")

pred_gnb_std_y1 <- predict_gnb_blended(
  x_test_mat, gnb_means_std_y1, gnb_sds_y1, classes_vec
) %>% factor(levels = classes_vec)

pred_gnb_prior_y1 <- predict_gnb_blended(
  x_test_mat, gnb_means_y1, gnb_sds_y1, classes_vec
) %>% factor(levels = classes_vec)

cat("--- Standard Naive Bayes (no prior) ---\n")
gnb_std_result <- eval_clf(test_HL$HL_class, pred_gnb_std_y1,
                            "Naive Bayes (standard)")
cat("Confusion matrix:\n")
print(gnb_std_result$confusion)

cat("\n--- Naive Bayes with Physical Priors ---\n")
gnb_prior_result <- eval_clf(test_HL$HL_class, pred_gnb_prior_y1,
                              "Naive Bayes (physical prior)")
cat("Confusion matrix:\n")
print(gnb_prior_result$confusion)

results_HL_clf <- rbind(results_HL_clf,
  data.frame(Model = "Naive Bayes (standard)",       Accuracy = gnb_std_result$accuracy),
  data.frame(Model = "Naive Bayes (physical prior)",  Accuracy = gnb_prior_result$accuracy)
)

# Prior effect plot
data.frame(
  Feature    = prior_feats,
  Low_data   = as.numeric(gnb_y1$params$mu["Low",  prior_feats]),
  Low_prior  = as.numeric(gnb_means_y1["Low",  prior_feats]),
  High_data  = as.numeric(gnb_y1$params$mu["High", prior_feats]),
  High_prior = as.numeric(gnb_means_y1["High", prior_feats])
) %>%
  pivot_longer(-Feature, names_to = "Type", values_to = "Mean") %>%
  mutate(Class  = ifelse(grepl("Low",  Type), "Low",  "High"),
         Source = ifelse(grepl("data", Type), "Data only", "With physical prior")) %>%
  ggplot(aes(x = Feature, y = Mean, fill = Source, alpha = Class)) +
  geom_col(position = "dodge") +
  scale_alpha_manual(values = c("Low" = 0.6, "High" = 1.0)) +
  scale_fill_manual(values = c("Data only" = "steelblue",
                                "With physical prior" = "tomato")) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  labs(title    = "Physical Prior Effect on Class-Conditional Means — Heating Load",
       subtitle = "Bars show data-only vs prior-blended means (scaled feature space)",
       x = "Feature", y = "Class-conditional mean (scaled)") +
  theme_minimal()
```

```{r clf-nb-cl}
# ── Cooling Load NB — same pipeline; CL prior reversal: glazing increases CL ──
physical_prior_CL <- list(
  Low    = list(RelCompactness=0.5,  GlazingArea=-0.5, OverallHeight=-0.4,
                CompactnessXHeight=0.3, WallXGlazing=-0.3),
  Medium = list(RelCompactness=0.0,  GlazingArea=0.0,  OverallHeight=0.0,
                CompactnessXHeight=0.0, WallXGlazing=0.0),
  High   = list(RelCompactness=-0.5, GlazingArea=0.5,  OverallHeight=0.4,
                CompactnessXHeight=-0.3, WallXGlazing=0.3)
)

set.seed(42)
gnb_y2 <- gaussian_naive_bayes(
  x = as.matrix(train_CL[, features]),
  y = train_CL$CL_class
)

gnb_means_y2     <- gnb_y2$params$mu
gnb_means_std_y2 <- gnb_y2$params$mu
gnb_sds_y2       <- gnb_y2$params$sd

for (feat in colnames(gnb_sds_y2)) {
  fallback <- max(sd(train_CL[[feat]], na.rm = TRUE) * 0.5, 0.1)
  for (cls in rownames(gnb_sds_y2)) {
    val <- gnb_sds_y2[cls, feat]
    if (is.nan(val) || is.na(val) || val < 0.1) gnb_sds_y2[cls, feat] <- fallback
  }
}

for (feat in prior_feats) {
  if (!feat %in% colnames(gnb_means_y2)) next
  for (cls in c("Low","Medium","High")) {
    pv <- physical_prior_CL[[cls]][[feat]]
    if (!is.null(pv))
      gnb_means_y2[cls, feat] <- blend_means(gnb_means_y2[cls, feat], pv)
  }
}

x_test_CL <- as.matrix(test_CL[, features])

pred_gnb_std_y2 <- predict_gnb_blended(
  x_test_CL, gnb_means_std_y2, gnb_sds_y2, classes_vec
) %>% factor(levels = classes_vec)

pred_gnb_prior_y2 <- predict_gnb_blended(
  x_test_CL, gnb_means_y2, gnb_sds_y2, classes_vec
) %>% factor(levels = classes_vec)

cat("--- Standard Naive Bayes (no prior) — CL ---\n")
gnb_std_CL  <- eval_clf(test_CL$CL_class, pred_gnb_std_y2, "Naive Bayes (standard)")
cat("--- Naive Bayes with Physical Priors — CL ---\n")
gnb_prior_CL <- eval_clf(test_CL$CL_class, pred_gnb_prior_y2,
                          "Naive Bayes (physical prior)")

results_CL_clf <- rbind(results_CL_clf,
  data.frame(Model = "Naive Bayes (standard)",       Accuracy = gnb_std_CL$accuracy),
  data.frame(Model = "Naive Bayes (physical prior)",  Accuracy = gnb_prior_CL$accuracy)
)
```

## Ordinal Error Analysis

One advantage of ordinal classification is the ability to break errors down by
severity. This section computes the error distance distribution — how many of
each model's mistakes were one class off versus two classes off.

```{r clf-ordinal-errors}
ordinal_distance <- function(actual, predicted) {
  class_order <- c("Low" = 1, "Medium" = 2, "High" = 3)
  abs(class_order[as.character(actual)] - class_order[as.character(predicted)])
}

error_dist_HL <- data.frame(
  Model = c(rep("Standards-Based (EU)", nrow(test_HL)),
            rep("Ordinal Random Forest", nrow(test_HL)),
            rep("NB (standard)",         nrow(test_HL)),
            rep("NB (physical prior)",   nrow(test_HL))),
  Distance = c(
    ordinal_distance(test_HL$HL_class, standards_pred_HL),
    ordinal_distance(test_HL$HL_class, pred_class_orf_y1),
    ordinal_distance(test_HL$HL_class, pred_gnb_std_y1),
    ordinal_distance(test_HL$HL_class, pred_gnb_prior_y1)
  )
)

error_dist_HL %>%
  group_by(Model, Distance) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Model) %>%
  mutate(Proportion = Count / sum(Count)) %>%
  ggplot(aes(x = factor(Distance), y = Proportion,
             fill = factor(Distance))) +
  geom_col() +
  facet_wrap(~Model, ncol = 2) +
  scale_fill_manual(
    values = c("0" = "darkgreen", "1" = "goldenrod", "2" = "tomato"),
    labels = c("0" = "Correct", "1" = "1 class off", "2" = "2 classes off"),
    name   = "Error severity"
  ) +
  scale_y_continuous(labels = scales::percent) +
  labs(title    = "Ordinal Error Distribution — Heating Load",
       subtitle = "Green = correct; red = worst errors",
       x = "Error distance (classes)", y = "Proportion of test buildings") +
  theme_minimal()
```

## Classification Model Comparison

```{r clf-comparison}
results_HL_clf <- results_HL_clf %>% arrange(desc(Accuracy))
results_CL_clf <- results_CL_clf %>% arrange(desc(Accuracy))

cat("=== HEATING LOAD — Classification Comparison ===\n")
print(results_HL_clf, row.names = FALSE)

cat("\n=== COOLING LOAD — Classification Comparison ===\n")
print(results_CL_clf, row.names = FALSE)

p_clf_HL <- ggplot(results_HL_clf,
                    aes(x = reorder(Model, Accuracy), y = Accuracy)) +
  geom_col(fill = "steelblue", alpha = 0.85) +
  geom_text(aes(label = paste0(round(Accuracy * 100, 1), "%")),
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 1.1), labels = scales::percent) +
  labs(title    = "Classification Models — Accuracy: Heating Load",
       subtitle = "Standards-Based is the no-training domain baseline",
       x = "Model", y = "Accuracy") +
  theme_minimal()

p_clf_CL <- ggplot(results_CL_clf,
                    aes(x = reorder(Model, Accuracy), y = Accuracy)) +
  geom_col(fill = "darkorange", alpha = 0.85) +
  geom_text(aes(label = paste0(round(Accuracy * 100, 1), "%")),
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 1.1), labels = scales::percent) +
  labs(title    = "Classification Models — Accuracy: Cooling Load",
       subtitle = "Standards-Based is the no-training domain baseline",
       x = "Model", y = "Accuracy") +
  theme_minimal()

grid.arrange(p_clf_HL, p_clf_CL, ncol = 2)
```

---

# Part 5 — Cross-Model Comparison and Conclusions

## Regression: All Models Side by Side

This section brings together the linear baselines from Part 2 and the
domain-appropriate regression results from Part 3 into a single comparison.
All metrics are on the same held-out test set (the 20% group-aware holdout,
set.seed=42).

```{r final-reg-comparison}
# Heating load — full summary including linear models
reg_summary_HL <- bind_rows(
  results_HL %>%
    mutate(Family = "Linear (Part 2)") %>%
    rename(RMSE_val = RMSE, MAE_val = MAE, R2_val = R2),
  results_y1 %>%
    mutate(Family = ifelse(grepl("Baseline", Model), "Linear (Part 2)",
                           "Domain-Appropriate (Part 3)")) %>%
    rename(RMSE_val = RMSE, MAE_val = MAE, R2_val = R2)
) %>% distinct(Model, .keep_all = TRUE)

cat("=== FULL REGRESSION COMPARISON — Heating Load ===\n")
print(reg_summary_HL %>% arrange(RMSE_val) %>%
        select(Model, Family, RMSE_val, MAE_val, R2_val), row.names = FALSE)

# Cooling load
reg_summary_CL <- bind_rows(
  results_CL %>%
    mutate(Family = "Linear (Part 2)") %>%
    rename(RMSE_val = RMSE, MAE_val = MAE, R2_val = R2),
  results_y2 %>%
    mutate(Family = ifelse(grepl("Baseline", Model), "Linear (Part 2)",
                           "Domain-Appropriate (Part 3)")) %>%
    rename(RMSE_val = RMSE, MAE_val = MAE, R2_val = R2)
) %>% distinct(Model, .keep_all = TRUE)

cat("\n=== FULL REGRESSION COMPARISON — Cooling Load ===\n")
print(reg_summary_CL %>% arrange(RMSE_val) %>%
        select(Model, Family, RMSE_val, MAE_val, R2_val), row.names = FALSE)

# Visual
p_all_HL <- ggplot(reg_summary_HL %>% arrange(RMSE_val),
                    aes(x = reorder(Model, RMSE_val), y = RMSE_val, fill = Family)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = round(RMSE_val, 3)), hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_manual(values = c("Linear (Part 2)" = "steelblue",
                                "Domain-Appropriate (Part 3)" = "tomato")) +
  labs(title    = "All Regression Models — RMSE: Heating Load",
       subtitle = "Blue = linear; red = domain-appropriate",
       x = "Model", y = "RMSE (kWh/m²)", fill = NULL) +
  theme_minimal()

p_all_CL <- ggplot(reg_summary_CL %>% arrange(RMSE_val),
                    aes(x = reorder(Model, RMSE_val), y = RMSE_val, fill = Family)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = round(RMSE_val, 3)), hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_manual(values = c("Linear (Part 2)" = "steelblue",
                                "Domain-Appropriate (Part 3)" = "darkorange")) +
  labs(title    = "All Regression Models — RMSE: Cooling Load",
       subtitle = "Blue = linear; orange = domain-appropriate",
       x = "Model", y = "RMSE (kWh/m²)", fill = NULL) +
  theme_minimal()

grid.arrange(p_all_HL, p_all_CL, ncol = 2)
```

## Conclusions

The analysis proceeds through four interconnected stages, each motivating the
next. The key findings and linkages are as follows.

**From the EDA:** The dataset's combinatorial grid structure means all modelling
must use group-aware splits to avoid orientation-sibling leakage. The bimodal
response distributions, driven entirely by `OverallHeight`, and the severe VIF
inflation (up to 212 for `RoofArea`) are the two facts that every modelling
decision flows from.

**From the linear models:** Regularisation delivers a measurable improvement
over baseline OLS — 14.5% RMSE reduction for heating load and 8.6% for cooling
load — confirming that the VIF-flagged collinearity was actively hurting raw OLS
performance. The best linear models are Elastic Net for heating load
(RMSE=2.428, R²=0.941) and Ridge for cooling load (RMSE=2.888, R²=0.903).
The Breusch-Pagan test (BP=268.9, p<2.2e-16) confirms severe heteroscedasticity
in the best linear model, a direct consequence of the two-height-cluster
structure that linear methods cannot partition.

**From the domain-appropriate regression models:** LightGBM substantially
outperforms all linear baselines, achieving RMSE=0.519 for heating load
(R²=0.997) and RMSE=1.640 for cooling load (R²=0.969). The GPR models offer
a useful secondary finding: GPR (RBF) outperforms GPR (Laplace) for both
targets, suggesting the geometry-to-load relationship is smoother than the
step-like character of the discrete feature values implies. The physics
consistency check confirms that all three domain models preserve the physically
expected monotonic orderings for glazing, compactness, and height — they have
not learned spurious patterns.

**From the classification models:** The Ordinal Random Forest achieves
substantially higher accuracy than the no-training Standards-Based classifier
on both targets, quantifying the value that learning from data adds over domain
rules alone. The Ranked Probability Score results show that the Ordinal Random
Forest produces very few two-step errors, which is the property that matters
most for practical building retrofit decisions. The prior-blended Naive Bayes
demonstrates that physically motivated priors can be incorporated without
large accuracy losses, offering a more interpretable model at a modest
performance cost.

**Overall:** For regression, LightGBM should be the model carried forward for
building energy load prediction on ENB2012-like datasets. For classification,
the Ordinal Random Forest is the preferred model, with the Standards-Based
classifier retained as a benchmark for communicating the marginal value of
machine learning over regulatory thresholds. In both cases, the domain-specific
model design — group-aware splits, physics-consistent feature engineering, and
ordinal-aware loss functions — contributed substantially more to performance
than model family selection alone.

---
