---
title: "Final Project — Nonlinear Model Analysis"
author: "Jakes Chege"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    number_sections: true
  pdf_document:
    toc: true
---

```{r packages}
pkgs <- c("tidyverse", "caret", "randomForest", "xgboost", "nnet",
          "mgcv", "splines", "corrplot", "ggplot2", "dplyr",
          "pdp", "vip", "readxl", "rpart", "rpart.plot", "earth")

lapply(pkgs, library, character.only = TRUE)
```

```{r setup-image-dir}
img_dir <- "/home/jakes/Documents/strathmore/Modules/Module 3/linear_models/classwork/final_project/markdowns/images"
dir.create(img_dir, showWarnings = FALSE, recursive = TRUE)
```

# Load Data from EDA Output

We load the post-EDA Excel file produced at the end of `EDA.Rmd`.
This file already has renamed columns, factor variants (`_f`), and
an `is_outlier` flag column.

```{r}
df <- read_excel("~/Documents/strathmore/Modules/Module 3/linear_models/classwork/final_project/markdowns/new_data/ENB2012_post_EDA.xlsx")
```

```{r}
cat("Dimensions:", nrow(df), "rows x", ncol(df), "columns\n")
cat("Columns:", paste(names(df), collapse = ", "), "\n\n")

glimpse(df)
```

# Restore Factor Variables

Excel does not preserve R factor types. We restore them from the raw
numeric columns that were carried over.

```{r restore-factors}
df$Orientation_f <- factor(
  df$Orientation,
  levels = 2:5,
  labels = c("N", "E", "S", "W")
)

df$OverallHeight_f <- factor(
  df$OverallHeight,
  levels = c(3.5, 7.0),
  labels = c("Low_3.5m", "High_7m")
)

df$GlazingArea_f <- factor(
  df$GlazingArea,
  levels = c(0, 0.1, 0.25, 0.4),
  labels = c("None", "Low_10pct", "Med_25pct", "High_40pct")
)

df$GlazingAreaDist_f <- factor(
  df$GlazingAreaDist,
  levels = 0:5,
  labels = c("None", "Uniform", "N_heavy", "E_heavy", "S_heavy", "W_heavy")
)

cat("Factor levels restored:\n")
cat("  Orientation_f    :", levels(df$Orientation_f), "\n")
cat("  OverallHeight_f  :", levels(df$OverallHeight_f), "\n")
cat("  GlazingArea_f    :", levels(df$GlazingArea_f), "\n")
cat("  GlazingAreaDist_f:", levels(df$GlazingAreaDist_f), "\n")
```

# Outlier Handling

The `is_outlier` flag was computed in EDA using the IQR method across
all numeric predictors and responses. We inspect the flagged rows and
decide on a strategy.

## Inspect flagged rows

```{r outlier-inspect}
cat("Total rows flagged as outliers:", sum(df$is_outlier, na.rm = TRUE), "\n")
cat("Percentage:", round(mean(df$is_outlier, na.rm = TRUE) * 100, 2), "%\n\n")

# Preview flagged rows
df %>%
  filter(is_outlier == TRUE) %>%
  dplyr::select(RelCompactness:CoolingLoad) %>%
  summary()
```

Since there are no outliers, we keep the data as it is.

# Response Variable Transformation

## Visual check: raw vs log

```{r log-check}
png(file.path(img_dir, "response_raw_vs_log.png"), width = 1200, height = 900, res = 150)
par(mfrow = c(2, 2))

hist(df$HeatingLoad,      breaks = 30, col = "lightblue",
     main = "Y1: Heating Load (raw)", xlab = "HeatingLoad")
hist(log(df$HeatingLoad), breaks = 30, col = "lightgreen",
     main = "log(Y1): Heating Load",  xlab = "log(HeatingLoad)")

hist(df$CoolingLoad,      breaks = 30, col = "lightblue",
     main = "Y2: Cooling Load (raw)", xlab = "CoolingLoad")
hist(log(df$CoolingLoad), breaks = 30, col = "lightgreen",
     main = "log(Y2): Cooling Load",  xlab = "log(CoolingLoad)")

par(mfrow = c(1, 1))
dev.off()

# Render inline as well
par(mfrow = c(2, 2))
hist(df$HeatingLoad,      breaks = 30, col = "lightblue",
     main = "Y1: Heating Load (raw)", xlab = "HeatingLoad")
hist(log(df$HeatingLoad), breaks = 30, col = "lightgreen",
     main = "log(Y1): Heating Load",  xlab = "log(HeatingLoad)")
hist(df$CoolingLoad,      breaks = 30, col = "lightblue",
     main = "Y2: Cooling Load (raw)", xlab = "CoolingLoad")
hist(log(df$CoolingLoad), breaks = 30, col = "lightgreen",
     main = "log(Y2): Cooling Load",  xlab = "log(CoolingLoad)")
par(mfrow = c(1, 1))
```

# Multicollinearity & VIF Review

SurfaceArea was found in EDA to be perfectly collinear with
WallArea + RoofArea. We drop it from the modelling set and confirm
VIF scores are acceptable for the remaining predictors.

```{r vif}
vif_model <- lm(
  HeatingLoad ~ RelCompactness + WallArea + RoofArea +
                OverallHeight + Orientation + GlazingArea + GlazingAreaDist,
  data = df
)

vif_vals <- car::vif(vif_model)

vif_table <- data.frame(
  Variable = names(vif_vals),
  VIF      = round(vif_vals, 3),
  Flag     = ifelse(vif_vals > 10, "HIGH — consider dropping",
             ifelse(vif_vals > 5,  "Moderate — monitor", "OK"))
)

print(vif_table)
```

We will keep them all, and accept the VIF. This is because Random Forest, XGBoost, and GAMs are not affected by multicollinearity at all. High VIF only damages OLS coefficient interpretation, not prediction accuracy.

# Polynomial Relationship Check (all predictors)

We extend the single-predictor check to cover all continuous predictors against
both outcomes, so we know which quadratic terms are worth including as engineered
features.

```{r poly-check, fig.height=10}
cont_preds <- c("RelCompactness", "WallArea", "RoofArea",
                "OverallHeight", "GlazingArea")

make_poly_plots <- function(response_var, point_col) {
  lapply(cont_preds, function(v) {
    ggplot(df, aes(x = .data[[v]], y = .data[[response_var]])) +
      geom_point(alpha = 0.25, size = 0.7, color = point_col) +
      geom_smooth(method = "lm",  formula = y ~ x,
                  color = "steelblue", se = FALSE, linewidth = 0.8) +
      geom_smooth(method = "lm",  formula = y ~ poly(x, 2),
                  color = "tomato", se = FALSE, linewidth = 0.8) +
      geom_smooth(method = "loess",
                  color = "forestgreen", se = FALSE,
                  linewidth = 0.8, linetype = "dashed") +
      labs(title = paste(v, "→", response_var),
           subtitle = "Blue=linear  Red=quadratic  Green=LOESS",
           x = v, y = response_var) +
      theme_minimal(base_size = 9)
  })
}

p_y1 <- make_poly_plots("HeatingLoad", "steelblue")
poly_y1_plot <- do.call(gridExtra::grid.arrange, c(p_y1, ncol = 3,
  list(top = "Polynomial check — Heating Load (Y1)")))
ggsave(file.path(img_dir, "poly_check_heating_load.png"),
       plot = poly_y1_plot, width = 14, height = 10, dpi = 150)

p_y2 <- make_poly_plots("CoolingLoad", "darkorange")
poly_y2_plot <- do.call(gridExtra::grid.arrange, c(p_y2, ncol = 3,
  list(top = "Polynomial check — Cooling Load (Y2)")))
ggsave(file.path(img_dir, "poly_check_cooling_load.png"),
       plot = poly_y2_plot, width = 14, height = 10, dpi = 150)
```

# Feature Engineering

## Interaction and ratio features

Physically motivated engineered features based on the scatter plots
and domain interpretation from EDA.

```{r feature-engineering}
df_clean <- df %>%
  mutate(
    # --- Original four from preprocessing.Rmd ---
    CompactnessXHeight  = RelCompactness * OverallHeight,
    WallToRoofRatio     = WallArea / RoofArea,
    GlazingXCompactness = GlazingArea * RelCompactness,
    SurfToVol           = SurfaceArea / (WallArea + RoofArea),

    # --- Additional physically motivated features ---
    GlazingXHeight      = GlazingArea * OverallHeight,
    # Tall buildings with more glazing lose/gain disproportionately more heat

    WallXGlazing        = WallArea * GlazingArea,
    # Interaction between exposed wall and window area

    CompactnessSquared  = RelCompactness^2,
    # Captures the nonlinear compactness-load relationship seen in poly plots

    HeightSquared       = OverallHeight^2,
    # OverallHeight is binary (3.5 / 7.0) so this is less critical,
    # but retained for completeness in polynomial model pipelines

    GlazingToWall       = GlazingArea / (WallArea + 1),
    # Facade transparency ratio: proportion of wall that is glass

    EnvelopeArea        = WallArea + RoofArea,
    # Total heat-exchange envelope (SurfaceArea minus floor, approx)

    WallPerHeight       = WallArea / (OverallHeight + 1)
    # Normalised wall exposure per unit height
  )

cat("New features added. Total columns now:", ncol(df_clean), "\n")
cat("Engineered feature names:\n")
cat(paste(" -",
  c("CompactnessXHeight","WallToRoofRatio","GlazingXCompactness",
    "SurfToVol","GlazingXHeight","WallXGlazing",
    "CompactnessSquared","HeightSquared","GlazingToWall",
    "EnvelopeArea","WallPerHeight")), sep = "\n")
```

```{r}
library(car)
candidate_features_clean3 <- c(
  "CompactnessXHeight",  # best single summary of shape x height
  "WallToRoofRatio",     # best single summary of envelope geometry
  "GlazingXHeight",      # glazing penalty amplified by height
  "CompactnessSquared"   # captures the nonlinear compactness curve
)

vif_model_y2d <- lm(
  CoolingLoad ~ .,
  data = df_clean[, c(candidate_features_clean3, "CoolingLoad")]
)

vif_values <- vif(vif_model_y2d)
print(sort(vif_values, decreasing = TRUE))
```

```{r}
candidate_features_final <- c(
  "CompactnessSquared",  # captures nonlinear compactness effect
  "WallToRoofRatio",     # envelope geometry
  "GlazingXHeight"       # glazing x height interaction
)

vif_model_y2e <- lm(
  CoolingLoad ~ .,
  data = df_clean[, c(candidate_features_final, "CoolingLoad")]
)

vif_values <- vif(vif_model_y2e)
print(sort(vif_values, decreasing = TRUE))
```

```{r}
library(glmnet)

# Final feature set
final_features <- c(
  "CompactnessSquared",
  "WallToRoofRatio",
  "GlazingXHeight"
)

# Prepare matrices
set.seed(42)
train_idx <- sample(nrow(df_clean), 0.8 * nrow(df_clean))

X_train <- as.matrix(df_clean[train_idx, final_features])
X_test  <- as.matrix(df_clean[-train_idx, final_features])
y_train <- df_clean$CoolingLoad[train_idx]
y_test  <- df_clean$CoolingLoad[-train_idx]

# Cross-validated Elastic Net
cv_fit_y2 <- cv.glmnet(X_train, y_train, alpha = 0.5, nfolds = 10)

png(file.path(img_dir, "elastic_net_cv_preliminary.png"), width = 900, height = 700, res = 150)
plot(cv_fit_y2)
dev.off()
plot(cv_fit_y2)

# Best lambda
cat("Best lambda (min):", cv_fit_y2$lambda.min, "\n")

# Coefficients
cat("\nModel coefficients:\n")
print(coef(cv_fit_y2, s = "lambda.min"))

# Predictions and evaluation
preds <- predict(cv_fit_y2, newx = X_test, s = "lambda.min")

rmse <- sqrt(mean((y_test - preds)^2))
mae  <- mean(abs(y_test - preds))
ss_res <- sum((y_test - preds)^2)
ss_tot <- sum((y_test - mean(y_train))^2)
r2   <- 1 - ss_res / ss_tot

cat("\n--- Test Set Performance ---\n")
cat("RMSE:", round(rmse, 3), "\n")
cat("MAE: ", round(mae,  3), "\n")
cat("R2:  ", round(r2,   3), "\n")
```

## Correlation of engineered features with outcomes

Quick screen to confirm the new features are informative before
they enter modelling.

```{r feature-corr}
eng_features <- c(
  "CompactnessXHeight","WallToRoofRatio","GlazingXCompactness",
  "SurfToVol","GlazingXHeight","WallXGlazing",
  "CompactnessSquared","GlazingToWall","EnvelopeArea","WallPerHeight"
)

cor_eng <- df_clean %>%
  dplyr::select(all_of(eng_features), HeatingLoad, CoolingLoad) %>%
  cor(use = "complete.obs")

# Show only correlations with the two outcomes
cor_with_outcomes <- data.frame(
  Feature         = eng_features,
  Cor_HeatingLoad = round(cor_eng[eng_features, "HeatingLoad"], 3),
  Cor_CoolingLoad = round(cor_eng[eng_features, "CoolingLoad"], 3)
) %>%
  arrange(desc(abs(Cor_HeatingLoad)))

print(cor_with_outcomes)
```

The top three features, SurfToVol, WallToRoofRatio, and CompactnessXHeight, act as the primary drivers, all demonstrating high-impact correlations exceeding 0.80 for both heating and cooling outcomes. The consistent magnitude across both targets implies that these predictors capture fundamental geometric properties that affect thermal efficiency regardless of whether the goal is heating or cooling.

The mid-tier group, led by WallPerHeight and GlazingXHeight, provides secondary stability with correlations in the 0.60–0.70 range, ensuring the model incorporates localized architectural details beyond raw proportions. Meanwhile, the lower-tier features like EnvelopeArea and GlazingToWall show notably weaker or negligible associations, suggesting they contribute minimal signal and might be candidates for dimensionality reduction or pruning. Overall, the model benefits from a "distributed load" of information rather than relying on a single, isolated variable, which typically improves generalization.

# Train / Test Split

The split is done **before** scaling. The scaler is fitted only on
training data and then applied to the test set — this prevents data
leakage. We stratify on `OverallHeight_f` because it is the most
dominant predictor identified in EDA.

```{r train-test-split}
set.seed(42)

train_idx <- caret::createDataPartition(
  df_clean$OverallHeight_f,
  p    = 0.80,
  list = FALSE
)

df_train <- df_clean[ train_idx, ]
df_test  <- df_clean[-train_idx, ]

cat("Training set:", nrow(df_train), "rows\n")
cat("Test set    :", nrow(df_test),  "rows\n")

# Confirm stratification held
cat("\nOverallHeight distribution — train:\n")
print(prop.table(table(df_train$OverallHeight_f)))

cat("\nOverallHeight distribution — test:\n")
print(prop.table(table(df_test$OverallHeight_f)))

eval_reg <- function(actual, predicted, model_name) {
  rmse <- sqrt(mean((actual - predicted)^2))
  mae  <- mean(abs(actual - predicted))
  r2   <- 1 - sum((actual - predicted)^2) / sum((actual - mean(actual))^2)
  cat(sprintf("%-30s  RMSE=%.4f  MAE=%.4f  R²=%.4f\n", model_name, rmse, mae, r2))
  return(data.frame(Model = model_name, RMSE = rmse, MAE = mae, R2 = r2))
}

results_y1 <- data.frame()
```

# Feature Scaling (Centre + Scale)

Scale-sensitive models (Ridge, Lasso, SVR, KNN, Neural Net) require
standardised predictors. Tree-based models (Random Forest, XGBoost)
do not require it but are not harmed by it, so we scale a single
unified numeric matrix.

Scaling is fitted on **training data only**.

```{r scaling}
scale_cols <- c(
  "RelCompactness", "WallArea", "RoofArea",
  "OverallHeight", "Orientation", "GlazingArea", "GlazingAreaDist",
  "CompactnessXHeight", "WallToRoofRatio", "GlazingXCompactness",
  "SurfToVol", "GlazingXHeight", "WallXGlazing",
  "CompactnessSquared", "GlazingToWall", "EnvelopeArea", "WallPerHeight"
)

preproc_params <- caret::preProcess(
  df_train[, scale_cols],
  method = c("center", "scale")
)

df_train[, scale_cols] <- predict(preproc_params, df_train[, scale_cols])
df_test[, scale_cols]  <- predict(preproc_params, df_test[, scale_cols])

cat("Scaling complete. Parameters fitted on training set only.\n\n")

train_stats <- df_train %>%
  dplyr::select(all_of(scale_cols)) %>%
  summarise(across(everything(),
    list(mean = ~round(mean(.), 4), sd = ~round(sd(.), 4))))

cat("Sample check (first 3 features) — should be mean≈0, sd≈1:\n")
print(train_stats[, 1:6])
```

# Semi-parametric and Nonlinear Models

Here, we cover the more flexible end of the modelling spectrum. Unlike the purely parametric models, the methods here either relax or abandon the assumption of a fixed functional form. Some, like the Generalised Additive Model, remain partially interpretable by estimating smooth functions of individual predictors. Others, like Random Forest and XGBoost, treat the data as a pattern recognition problem and make no assumptions about the shape of the relationships at all. The tradeoff is that flexibility typically comes at the cost of some interpretability, though variable importance measures go some way towards compensating for this. All models are evaluated on the same held-out test set used in the regression file so that the results are directly comparable.

## Y1 — Heating Load

The nonlinear models are applied first to heating load. Given the physical nature of building thermodynamics, there is good reason to expect some of the relationships between the building features and energy demand to be genuinely nonlinear, which makes this family of models a natural fit for the problem.

### Model 7 — GAM

The Generalised Additive Model sits at the boundary between parametric and nonlinear approaches. It replaces the fixed linear terms with smooth spline functions estimated from the data, allowing each predictor to follow its own flexible curve. Orientation and glazing area distribution are kept as linear terms given their discrete or near-discrete nature, while the remaining continuous predictors are given smooth terms. The REML fitting method is used to avoid overfitting the smooths, and the gam.check output is inspected to verify that the smooth complexity is appropriate.

```{r}
# ============================================================
# 8.7  MODEL 7 — GAM: Generalised Additive Model (Semi-parametric)
# ============================================================

# Temporary copies to avoid permanently converting columns to factors
gam_train <- df_train
gam_test  <- df_test

for (col in c("RoofArea", "GlazingArea", "GlazingAreaDist",
              "OverallHeight", "Orientation")) {
  gam_train[[col]] <- factor(gam_train[[col]])
  gam_test[[col]]  <- factor(gam_test[[col]],
                              levels = levels(gam_train[[col]]))
}

gam_y1 <- gam(
  HeatingLoad ~
    s(SurfToVol, k = 10) +
    s(CompactnessXHeight, k = 8) +
    s(GlazingXHeight, k = 6) +
    Orientation,
  data   = gam_train,
  method = "REML"
)

summary(gam_y1)

# Save GAM smooth effect plots
png(file.path(img_dir, "gam_y1_smooth_effects.png"), width = 1200, height = 900, res = 150)
plot(gam_y1, pages = 1, shade = TRUE, shade.col = "lightblue",
     main = "GAM Smooth Effects — Heating Load (Y1)")
dev.off()
plot(gam_y1, pages = 1, shade = TRUE, shade.col = "lightblue")

# Save gam.check diagnostics
png(file.path(img_dir, "gam_y1_diagnostics.png"), width = 1200, height = 900, res = 150)
gam.check(gam_y1)
dev.off()
gam.check(gam_y1)

pred_gam_y1 <- predict(gam_y1, newdata = gam_test)
results_y1  <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_gam_y1, "GAM")
)
```

### Model 8 — MARS

Multivariate Adaptive Regression Splines builds a piecewise linear model by automatically detecting breakpoints, called knots, in the predictor space and fitting separate linear segments on either side. It also considers interaction terms up to degree two, meaning it can capture cases where the effect of one predictor depends on the value of another. The result is a model that is more flexible than ordinary linear regression but still produces an explicit equation that can be inspected and interpreted. Variable importance scores from the evimp function show which predictors contribute most to the fit.

```{r}
# ============================================================
# 8.8  MODEL 8 — MARS: Multivariate Adaptive Regression Splines (Non-linear)
# ============================================================

# Full feature set — original plus engineered, excluding aliased columns
mars_features <- c(
  "RelCompactness", "WallArea", "RoofArea",
  "OverallHeight", "Orientation", "GlazingArea", "GlazingAreaDist",
  "CompactnessXHeight", "WallToRoofRatio", "GlazingXCompactness",
  "SurfToVol", "GlazingXHeight", "WallXGlazing",
  "CompactnessSquared", "GlazingToWall", "EnvelopeArea", "WallPerHeight"
)

mars_y1 <- earth(
  HeatingLoad ~ .,
  data   = df_train[, c(mars_features, "HeatingLoad")],
  degree = 2
)

summary(mars_y1)

# Save plotmo response curves
png(file.path(img_dir, "mars_y1_response_curves.png"), width = 1400, height = 1000, res = 150)
plotmo(mars_y1, main = "MARS Response Curves — Heating Load (Y1)")
dev.off()
plotmo(mars_y1)

# Save variable importance (evimp)
mars_y1_imp <- evimp(mars_y1)
print(mars_y1_imp)

# evimp() returns a matrix whose column count can vary by earth version.
# Use column index 1 (nsubsets) rather than names to stay version-safe.
mars_imp_df <- data.frame(
  Feature  = rownames(mars_y1_imp),
  nsubsets = mars_y1_imp[, 1],
  stringsAsFactors = FALSE
) %>%
  arrange(desc(nsubsets))

p_mars_y1_imp <- ggplot(mars_imp_df, aes(x = reorder(Feature, nsubsets), y = nsubsets)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = nsubsets), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(title = "MARS Variable Importance — Heating Load (Y1)",
       x = "Feature", y = "Number of Model Subsets") +
  theme_minimal()
ggsave(file.path(img_dir, "mars_y1_variable_importance.png"),
       plot = p_mars_y1_imp, width = 10, height = 7, dpi = 150)
print(p_mars_y1_imp)

pred_mars_y1 <- predict(mars_y1, newdata = df_test)
results_y1   <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_mars_y1, "MARS")
)
```

### Model 9 — Regression Tree

A single regression tree partitions the feature space into rectangular regions by repeatedly splitting on the predictor that most reduces prediction error. The result is a model that is visually intuitive and makes no assumptions about functional form, but which can be unstable since small changes in the training data can produce quite different tree structures. A complexity parameter of 0.01 is set as the minimum split improvement threshold, and the tree is then pruned back to the depth that minimises cross-validated error, trading some training accuracy for better generalisation.

```{r}
# ============================================================
# 8.9  MODEL 9 — Regression Tree (Non-parametric)
# ============================================================
features <- c(
  "RelCompactness", "WallArea", "RoofArea",
  "OverallHeight", "Orientation", "GlazingArea", "GlazingAreaDist",
  "CompactnessXHeight", "WallToRoofRatio", "GlazingXCompactness",
  "SurfToVol", "GlazingXHeight", "WallXGlazing",
  "CompactnessSquared", "GlazingToWall", "EnvelopeArea", "WallPerHeight"
)

tree_y1 <- rpart(HeatingLoad ~ .,
                 data    = df_train[, c(features, "HeatingLoad")],
                 control = rpart.control(cp = 0.01))

# Save full tree plot
png(file.path(img_dir, "tree_y1_full.png"), width = 1400, height = 900, res = 150)
rpart.plot(tree_y1, main = "Regression Tree — Heating Load", roundint = FALSE)
dev.off()
rpart.plot(tree_y1, main = "Regression Tree — Heating Load", roundint = FALSE)

printcp(tree_y1)

optimal_cp <- tree_y1$cptable[which.min(tree_y1$cptable[, "xerror"]), "CP"]
tree_y1_pruned <- prune(tree_y1, cp = optimal_cp)

# Save pruned tree plot
png(file.path(img_dir, "tree_y1_pruned.png"), width = 1400, height = 900, res = 150)
rpart.plot(tree_y1_pruned, main = "Pruned Regression Tree — Heating Load", roundint = FALSE)
dev.off()
rpart.plot(tree_y1_pruned, main = "Pruned Regression Tree — Heating Load", roundint = FALSE)

# Variable importance from rpart
tree_y1_imp <- tree_y1_pruned$variable.importance
if (!is.null(tree_y1_imp) && length(tree_y1_imp) > 0) {
  tree_imp_df <- data.frame(
    Feature    = names(tree_y1_imp),
    Importance = as.numeric(tree_y1_imp)
  ) %>% arrange(desc(Importance))

  p_tree_y1_imp <- ggplot(tree_imp_df, aes(x = reorder(Feature, Importance), y = Importance)) +
    geom_col(fill = "steelblue", alpha = 0.8) +
    geom_text(aes(label = round(Importance, 1)), hjust = -0.2, size = 3) +
    coord_flip() +
    labs(title = "Regression Tree Variable Importance — Heating Load (Y1)",
         x = "Feature", y = "Importance") +
    theme_minimal()
  ggsave(file.path(img_dir, "tree_y1_variable_importance.png"),
         plot = p_tree_y1_imp, width = 10, height = 7, dpi = 150)
  print(p_tree_y1_imp)
}

pred_tree_y1 <- predict(tree_y1_pruned, newdata = df_test)
results_y1 <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_tree_y1, "Regression Tree"))
```

### Model 10 — Random Forest

Random Forest addresses the instability of single trees by growing a large number of them, each trained on a bootstrap sample of the training data and using a random subset of predictors at each split. Predictions are averaged across all 500 trees, which dramatically reduces variance compared to a single tree. The mtry parameter is set to 3, meaning three randomly selected features are considered at each split. Variable importance plots are produced using both mean decrease in accuracy and mean decrease in node impurity to give a rounded picture of which predictors matter most.

```{r}
# ============================================================
# 8.10  MODEL 10 — Random Forest (Non-parametric, Ensemble)
# ============================================================
features <- c(
  "RelCompactness", "WallArea", "RoofArea",
  "OverallHeight", "Orientation", "GlazingArea", "GlazingAreaDist",
  "CompactnessXHeight", "WallToRoofRatio", "GlazingXCompactness",
  "SurfToVol", "GlazingXHeight", "WallXGlazing",
  "CompactnessSquared", "GlazingToWall", "EnvelopeArea", "WallPerHeight"
)

set.seed(42)
rf_y1 <- randomForest(HeatingLoad ~ .,
                      data      = df_train[, c(features, "HeatingLoad")],
                      ntree     = 500,
                      mtry      = 3,
                      importance = TRUE)
print(rf_y1)

# Save variable importance plot
png(file.path(img_dir, "rf_y1_variable_importance.png"), width = 1200, height = 900, res = 150)
varImpPlot(rf_y1, main = "Random Forest Variable Importance — Heating Load (Y1)")
dev.off()
varImpPlot(rf_y1, main = "Random Forest Variable Importance — Heating Load (Y1)")

importance(rf_y1)

pred_rf_y1 <- predict(rf_y1, newdata = df_test)
results_y1 <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_rf_y1, "Random Forest"))
```

### Model 11 — XGBoost

XGBoost is a gradient boosting algorithm that builds trees sequentially, with each new tree correcting the errors made by the ensemble so far. It tends to be among the strongest performers on tabular data because it combines boosting with regularisation terms and subsampling of both rows and columns, which helps prevent overfitting. Early stopping is used here to halt training once the test set error stops improving, avoiding unnecessary rounds. Feature importance scores are extracted after training to identify the most influential predictors.

```{r}
# ============================================================
# 8.11  MODEL 11 — XGBoost (Non-parametric, Gradient Boosting)
# ============================================================
# Build numeric matrices from the current split (df_train/df_test have factor
# columns after GAM conversion, so model.matrix handles dummy-encoding here).
features <- c(
  "RelCompactness", "WallArea", "RoofArea",
  "OverallHeight", "Orientation", "GlazingArea", "GlazingAreaDist",
  "CompactnessXHeight", "WallToRoofRatio", "GlazingXCompactness",
  "SurfToVol", "GlazingXHeight", "WallXGlazing",
  "CompactnessSquared", "GlazingToWall", "EnvelopeArea", "WallPerHeight"
)

x_train <- model.matrix(HeatingLoad ~ .,
              data = df_train[, c(features, "HeatingLoad")])[, -1]
y_train  <- df_train$HeatingLoad
x_test   <- model.matrix(HeatingLoad ~ .,
              data = df_test[, c(features, "HeatingLoad")])[, -1]

dtrain_y1 <- xgb.DMatrix(data = x_train, label = y_train)
dtest_y1  <- xgb.DMatrix(data = x_test,  label = df_test$HeatingLoad)

params_y1 <- list(
  objective        = "reg:squarederror",
  eta              = 0.1,
  max_depth        = 6,
  subsample        = 0.8,
  colsample_bytree = 0.8
)

xgb_y1 <- xgb.train(
  params               = params_y1,
  data                 = dtrain_y1,
  nrounds              = 200,
  watchlist            = list(train = dtrain_y1, test = dtest_y1),
  verbose              = 0,
  early_stopping_rounds = 20
)

xgb_imp_y1 <- xgb.importance(model = xgb_y1)

# Save XGBoost feature importance plot
png(file.path(img_dir, "xgb_y1_feature_importance.png"), width = 1200, height = 900, res = 150)
xgb.plot.importance(xgb_imp_y1, top_n = 8, main = "XGBoost Feature Importance — Heating Load (Y1)")
dev.off()
xgb.plot.importance(xgb_imp_y1, top_n = 8, main = "XGBoost Feature Importance — Heating Load (Y1)")

pred_xgb_y1 <- predict(xgb_y1, newdata = dtest_y1)
results_y1 <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_xgb_y1, "XGBoost"))
```

### Model 12 — SVM Regression

Support Vector Machine regression finds a function that fits as many training points as possible within a tolerance band of width epsilon around the predicted values, while penalising points that fall outside that band. The radial basis function kernel is used here, which maps the data into a higher-dimensional space and allows the model to capture complex nonlinear patterns. The cost parameter of 10 controls how severely outlying points are penalised, and the epsilon value of 0.1 defines the width of the insensitive tube.

```{r}
# ============================================================
# 8.12  MODEL 12 — SVM Regression (Non-parametric, Kernel)
# ============================================================
library(e1071)
features <- c(
  "RelCompactness", "WallArea", "RoofArea",
  "OverallHeight", "Orientation", "GlazingArea", "GlazingAreaDist",
  "CompactnessXHeight", "WallToRoofRatio", "GlazingXCompactness",
  "SurfToVol", "GlazingXHeight", "WallXGlazing",
  "CompactnessSquared", "GlazingToWall", "EnvelopeArea", "WallPerHeight"
)

svm_y1 <- svm(HeatingLoad ~ .,
              data    = df_train[, c(features, "HeatingLoad")],
              type    = "eps-regression",
              kernel  = "radial",
              cost    = 10,
              epsilon = 0.1)
summary(svm_y1)

pred_svm_y1 <- predict(svm_y1, newdata = df_test)
results_y1 <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_svm_y1, "SVM (RBF)"))
```

### Model 13 — Neural Network

A single hidden layer neural network is fitted with ten neurons in the hidden layer and a linear output activation, which is appropriate for regression tasks. The features are standardised before training since neural networks are sensitive to the scale of their inputs. Weight decay regularisation is applied with a value of 0.01 to discourage the model from overfitting by penalising large weights. The network is trained for up to 500 iterations.

```{r}
# ============================================================
# 8.13  MODEL 13 — Neural Network Regression (Non-linear)
# ============================================================
features <- c(
  "RelCompactness", "WallArea", "RoofArea",
  "OverallHeight", "Orientation", "GlazingArea", "GlazingAreaDist",
  "CompactnessXHeight", "WallToRoofRatio", "GlazingXCompactness",
  "SurfToVol", "GlazingXHeight", "WallXGlazing",
  "CompactnessSquared", "GlazingToWall", "EnvelopeArea", "WallPerHeight"
)

preproc <- preProcess(df_train[, features], method = c("center", "scale"))
train_scaled <- predict(preproc, df_train[, features])
test_scaled  <- predict(preproc, df_test[, features])
train_scaled$HeatingLoad <- df_train$HeatingLoad

set.seed(42)
nn_y1 <- nnet(HeatingLoad ~ ., data = train_scaled,
              size   = 10,
              linout = TRUE,
              maxit  = 500,
              decay  = 0.01,
              trace  = FALSE)

pred_nn_y1 <- predict(nn_y1, newdata = test_scaled)
results_y1 <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_nn_y1, "Neural Network"))
```

### Model 14 — k-Nearest Neighbors

k-Nearest Neighbors regression predicts the heating load for a new observation by finding the k most similar buildings in the training set and averaging their observed heating loads. It makes no assumptions at all about the functional form of the relationship and adapts entirely to local patterns in the data. The value of k is selected through 10-fold cross-validation across a grid of candidate values ranging from 3 to 15. Smaller values of k produce more flexible but noisier predictions, while larger values produce smoother but potentially biased ones.

```{r}
# ============================================================
# 8.14  MODEL 14 — k-Nearest Neighbors Regression (Non-parametric)
# ============================================================
features <- c(
  "RelCompactness", "WallArea", "RoofArea",
  "OverallHeight", "Orientation", "GlazingArea", "GlazingAreaDist",
  "CompactnessXHeight", "WallToRoofRatio", "GlazingXCompactness",
  "SurfToVol", "GlazingXHeight", "WallXGlazing",
  "CompactnessSquared", "GlazingToWall", "EnvelopeArea", "WallPerHeight"
)

set.seed(42)
knn_ctrl <- trainControl(method = "cv", number = 10)
knn_y1 <- train(HeatingLoad ~ .,
                data      = df_train[, c(features, "HeatingLoad")],
                method    = "knn",
                preProcess = c("center", "scale"),
                trControl  = knn_ctrl,
                tuneGrid   = data.frame(k = c(3, 5, 7, 9, 11, 15)))
print(knn_y1)

# Save k-NN tuning curve
p_knn_y1 <- plot(knn_y1)
png(file.path(img_dir, "knn_y1_tuning_curve.png"), width = 900, height = 700, res = 150)
print(plot(knn_y1, main = "k-NN CV Tuning — Heating Load (Y1)"))
dev.off()
plot(knn_y1)

pred_knn_y1 <- predict(knn_y1, newdata = df_test)
results_y1 <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_knn_y1, "k-NN Regression"))
```

### Y1 — Model Comparison

All eight nonlinear and semi-parametric models are now ranked by test set RMSE. The spread of results across these models tends to be wider than for the parametric models, reflecting the greater variability in how well different flexible approaches suit a given dataset. The R squared comparison chart provides a complementary view of explanatory power.

```{r}
# ============================================================
# Y1 COMPARISON — Semi-parametric & Nonlinear Models Only
# ============================================================
results_y1 <- results_y1 %>% arrange(RMSE)
cat("\n========== Y1: HEATING LOAD — SEMI-PARAMETRIC & NONLINEAR MODEL COMPARISON ==========\n")
print(results_y1, row.names = FALSE)

p_rmse_y1 <- ggplot(results_y1, aes(x = reorder(Model, RMSE), y = RMSE)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = round(RMSE, 3)), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Semi-parametric & Nonlinear Model Comparison — Heating Load (Y1)",
       x = "Model", y = "RMSE (Test Set)") +
  theme_minimal()
ggsave(file.path(img_dir, "comparison_y1_rmse.png"),
       plot = p_rmse_y1, width = 10, height = 7, dpi = 150)
print(p_rmse_y1)

p_r2_y1 <- ggplot(results_y1, aes(x = reorder(Model, -R2), y = R2)) +
  geom_col(fill = "darkgreen", alpha = 0.7) +
  geom_text(aes(label = round(R2, 4)), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "R² Comparison — Heating Load (Y1)",
       x = "Model", y = "R²") +
  theme_minimal()
ggsave(file.path(img_dir, "comparison_y1_r2.png"),
       plot = p_r2_y1, width = 10, height = 7, dpi = 150)
print(p_r2_y1)
```

## Y2 — Cooling Load

The same eight models are now applied to cooling load. Cooling load is generally considered a slightly harder prediction target because it depends more strongly on external conditions like solar gain and ambient temperature, factors that interact with building geometry in complex ways. Flexible models are therefore well positioned to capture these dynamics.

```{r}
results_y2 <- data.frame()
```

### Elastic Net — Cooling Load (Y2)

```{r}
# ============================================================
# MODEL X — Elastic Net: CoolingLoad (Y2)
# ============================================================
library(car)
elastic_features_y2 <- c(
  "CompactnessSquared",
  "WallToRoofRatio",
  "GlazingXHeight",
  "Orientation"
)

X_train_en <- as.matrix(df_train[, elastic_features_y2])
X_test_en  <- as.matrix(df_test[,  elastic_features_y2])
y_train_y2 <- df_train$CoolingLoad
y_test_y2  <- df_test$CoolingLoad

set.seed(42)
cv_fit_y2 <- cv.glmnet(
  X_train_en, y_train_y2,
  alpha  = 0.5,
  nfolds = 10
)

png(file.path(img_dir, "elastic_net_y2_cv.png"), width = 900, height = 700, res = 150)
plot(cv_fit_y2, main = "Elastic Net CV — Cooling Load (Y2)")
dev.off()
plot(cv_fit_y2)

cat("Best lambda (min):", cv_fit_y2$lambda.min, "\n")
cat("\nModel coefficients:\n")
print(coef(cv_fit_y2, s = "lambda.min"))

preds_y2 <- predict(cv_fit_y2, newx = X_test_en, s = "lambda.min")

results_y2 <- rbind(results_y2,
  eval_reg(y_test_y2, as.vector(preds_y2), "Elastic Net"))
cat("\n--- Elastic Net: CoolingLoad Test Performance ---\n")
print(results_y2)
```

### Model 7 — GAM

The GAM for cooling load uses the same smooth term structure as the heating load model, with REML fitting to control smoothness. The partial effect plots produced by the model are particularly informative here because they show visually how each building characteristic influences cooling demand across its observed range.

```{r}
# --- 9.7  GAM ---
gam_y2 <- gam(
  CoolingLoad ~
    s(RelCompactness, k = 12) +
    s(WallArea, k = 7) +
    RoofArea + OverallHeight + Orientation +
    GlazingArea + GlazingAreaDist,
  data   = df_train,
  method = "REML"
)
summary(gam_y2)

# Save GAM smooth effects
png(file.path(img_dir, "gam_y2_smooth_effects.png"), width = 1200, height = 900, res = 150)
plot(gam_y2, pages = 1, shade = TRUE, shade.col = "lightyellow",
     main = "GAM Smooth Effects — Cooling Load (Y2)")
dev.off()
plot(gam_y2, pages = 1, shade = TRUE, shade.col = "lightyellow")

# Save GAM diagnostics
png(file.path(img_dir, "gam_y2_diagnostics.png"), width = 1200, height = 900, res = 150)
gam.check(gam_y2)
dev.off()
gam.check(gam_y2)

pred_gam_y2 <- predict(gam_y2, newdata = df_test)
results_y2 <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_gam_y2, "GAM"))
```

### Model 8 — MARS

MARS is applied to cooling load with the same degree-two setting to allow pairwise interactions. Comparing the knot locations and retained terms to those from the heating load model gives a sense of whether the same nonlinearities are present in both energy demands or whether the relationships differ in character.

```{r}
# --- 9.8  MARS ---
mars_y2 <- earth(CoolingLoad ~ .,
                 data   = df_train[, c(features, "CoolingLoad")],
                 degree = 2)
summary(mars_y2)

# Save MARS response curves
png(file.path(img_dir, "mars_y2_response_curves.png"), width = 1400, height = 1000, res = 150)
plotmo(mars_y2, main = "MARS Response Curves — Cooling Load (Y2)")
dev.off()
plotmo(mars_y2)

# Save MARS variable importance
mars_y2_imp <- evimp(mars_y2)
print(mars_y2_imp)

# evimp() returns a matrix whose column count can vary by earth version.
# Use column index 1 (nsubsets) rather than names to stay version-safe.
mars_imp_df_y2 <- data.frame(
  Feature  = rownames(mars_y2_imp),
  nsubsets = mars_y2_imp[, 1],
  stringsAsFactors = FALSE
) %>%
  arrange(desc(nsubsets))

p_mars_y2_imp <- ggplot(mars_imp_df_y2, aes(x = reorder(Feature, nsubsets), y = nsubsets)) +
  geom_col(fill = "darkorange", alpha = 0.8) +
  geom_text(aes(label = nsubsets), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(title = "MARS Variable Importance — Cooling Load (Y2)",
       x = "Feature", y = "Number of Model Subsets") +
  theme_minimal()
ggsave(file.path(img_dir, "mars_y2_variable_importance.png"),
       plot = p_mars_y2_imp, width = 10, height = 7, dpi = 150)
print(p_mars_y2_imp)

pred_mars_y2 <- predict(mars_y2, newdata = df_test)
results_y2 <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_mars_y2, "MARS"))
```

### Model 9 — Regression Tree

The regression tree for cooling load is grown and pruned using the same procedure as for heating load. Because the tree partitions the feature space based purely on reducing prediction error, the structure of the cooling load tree may differ substantially from the heating load tree if the dominant predictors or their interaction patterns are different.

```{r}
# --- 9.9  Regression Tree ---
tree_y2 <- rpart(CoolingLoad ~ .,
                 data = df_train[, c(features, "CoolingLoad")])
optimal_cp_y2 <- tree_y2$cptable[which.min(tree_y2$cptable[, "xerror"]), "CP"]
tree_y2_pruned <- prune(tree_y2, cp = optimal_cp_y2)

# Save pruned tree
png(file.path(img_dir, "tree_y2_pruned.png"), width = 1400, height = 900, res = 150)
rpart.plot(tree_y2_pruned, main = "Pruned Regression Tree — Cooling Load (Y2)", roundint = FALSE)
dev.off()
rpart.plot(tree_y2_pruned, main = "Pruned Regression Tree — Cooling Load (Y2)")

# Variable importance from rpart
tree_y2_imp <- tree_y2_pruned$variable.importance
if (!is.null(tree_y2_imp) && length(tree_y2_imp) > 0) {
  tree_imp_df_y2 <- data.frame(
    Feature    = names(tree_y2_imp),
    Importance = as.numeric(tree_y2_imp)
  ) %>% arrange(desc(Importance))

  p_tree_y2_imp <- ggplot(tree_imp_df_y2,
                          aes(x = reorder(Feature, Importance), y = Importance)) +
    geom_col(fill = "darkorange", alpha = 0.8) +
    geom_text(aes(label = round(Importance, 1)), hjust = -0.2, size = 3) +
    coord_flip() +
    labs(title = "Regression Tree Variable Importance — Cooling Load (Y2)",
         x = "Feature", y = "Importance") +
    theme_minimal()
  ggsave(file.path(img_dir, "tree_y2_variable_importance.png"),
         plot = p_tree_y2_imp, width = 10, height = 7, dpi = 150)
  print(p_tree_y2_imp)
}

pred_tree_y2 <- predict(tree_y2_pruned, newdata = df_test)
results_y2 <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_tree_y2, "Regression Tree"))
```

### Model 10 — Random Forest

The Random Forest for cooling load uses the same hyperparameter settings as the heating load forest. Variable importance plots are produced and can be compared to those from the heating load forest to see whether the ranking of predictors is consistent across both targets or whether certain features become more or less relevant depending on which energy demand is being modelled.

```{r}
# --- 9.10  Random Forest ---
set.seed(42)
rf_y2 <- randomForest(CoolingLoad ~ .,
                      data      = df_train[, c(features, "CoolingLoad")],
                      ntree     = 500,
                      mtry      = 3,
                      importance = TRUE)

# Save RF variable importance plot
png(file.path(img_dir, "rf_y2_variable_importance.png"), width = 1200, height = 900, res = 150)
varImpPlot(rf_y2, main = "Random Forest Variable Importance — Cooling Load (Y2)")
dev.off()
varImpPlot(rf_y2, main = "Random Forest Variable Importance — Cooling Load (Y2)")

pred_rf_y2 <- predict(rf_y2, newdata = df_test)
results_y2 <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_rf_y2, "Random Forest"))
```

### Model 11 — XGBoost

XGBoost is applied to cooling load by reusing the parameter list defined for heating load, keeping the comparison clean. Early stopping monitors the test set error and halts training when no improvement is seen for 20 consecutive rounds.

```{r}
# --- 9.11  XGBoost ---
y_train_y2 <- df_train$CoolingLoad
dtrain_y2  <- xgb.DMatrix(data = x_train, label = y_train_y2)
dtest_y2   <- xgb.DMatrix(data = x_test,  label = df_test$CoolingLoad)

xgb_y2 <- xgb.train(
  params               = params_y1,
  data                 = dtrain_y2,
  nrounds              = 200,
  watchlist            = list(train = dtrain_y2, test = dtest_y2),
  verbose              = 0,
  early_stopping_rounds = 20
)

xgb_imp_y2 <- xgb.importance(model = xgb_y2)

# Save XGBoost feature importance plot
png(file.path(img_dir, "xgb_y2_feature_importance.png"), width = 1200, height = 900, res = 150)
xgb.plot.importance(xgb_imp_y2, top_n = 8, main = "XGBoost Feature Importance — Cooling Load (Y2)")
dev.off()
xgb.plot.importance(xgb_imp_y2, top_n = 8, main = "XGBoost Feature Importance — Cooling Load (Y2)")

pred_xgb_y2 <- predict(xgb_y2, newdata = dtest_y2)
results_y2 <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_xgb_y2, "XGBoost"))
```

### Model 12 — SVM Regression

The radial basis function SVM is fitted to cooling load with the same cost and epsilon values used for heating load. If the error surface for cooling load is smoother or more regular, the SVM may perform relatively better here than it did for heating load.

```{r}
# --- 9.12  SVM ---
svm_y2 <- svm(CoolingLoad ~ .,
              data    = df_train[, c(features, "CoolingLoad")],
              type    = "eps-regression",
              kernel  = "radial",
              cost    = 10,
              epsilon = 0.1)

pred_svm_y2 <- predict(svm_y2, newdata = df_test)
results_y2 <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_svm_y2, "SVM (RBF)"))
```

### Model 13 — Neural Network

The neural network for cooling load is trained on the same scaled features produced during the heating load analysis, with a new column for the cooling load outcome appended to the scaled training data. The architecture and regularisation settings are kept identical to allow a fair comparison of how well the network adapts to each target.

```{r}
# --- 9.13  Neural Network ---
# Build a clean training frame: start from the scaled features only, then
# append CoolingLoad. This prevents HeatingLoad (added to train_scaled in the
# Y1 NN section) from leaking into the Y2 model as a predictor.
train_scaled_y2 <- train_scaled[, !(names(train_scaled) %in% c("HeatingLoad"))]
train_scaled_y2$CoolingLoad <- df_train$CoolingLoad

set.seed(42)
nn_y2 <- nnet(CoolingLoad ~ ., data = train_scaled_y2,
              size   = 10,
              linout = TRUE,
              maxit  = 500,
              decay  = 0.01,
              trace  = FALSE)

pred_nn_y2 <- predict(nn_y2, newdata = test_scaled)
results_y2 <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_nn_y2, "Neural Network"))
```

### Model 14 — k-Nearest Neighbors

k-NN is applied to cooling load using the same cross-validated k selection grid. Because the feature space is identical to the one used for heating load, the optimal k may differ if the local structure of the cooling load surface is smoother or more irregular than that of heating load.

```{r}
# --- 9.14  k-NN ---
knn_y2 <- train(CoolingLoad ~ .,
                data      = df_train[, c(features, "CoolingLoad")],
                method    = "knn",
                preProcess = c("center", "scale"),
                trControl  = knn_ctrl,
                tuneGrid   = data.frame(k = c(3, 5, 7, 9, 11, 15)))

# Save k-NN tuning curve
png(file.path(img_dir, "knn_y2_tuning_curve.png"), width = 900, height = 700, res = 150)
print(plot(knn_y2, main = "k-NN CV Tuning — Cooling Load (Y2)"))
dev.off()
plot(knn_y2)

pred_knn_y2 <- predict(knn_y2, newdata = df_test)
results_y2 <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_knn_y2, "k-NN Regression"))
```

### Y2 — Model Comparison

The final comparison ranks all nonlinear and semi-parametric models for cooling load by test set RMSE. Taken together with the equivalent table for heating load, these results give a clear picture of which flexible approaches generalise best across both energy demands and which are more target-specific in their performance.

```{r}
# ============================================================
# Y2 COMPARISON — Semi-parametric & Nonlinear Models Only
# ============================================================
results_y2 <- results_y2 %>% arrange(RMSE)
cat("\n========== Y2: COOLING LOAD — SEMI-PARAMETRIC & NONLINEAR MODEL COMPARISON ==========\n")
print(results_y2, row.names = FALSE)

p_rmse_y2 <- ggplot(results_y2, aes(x = reorder(Model, RMSE), y = RMSE)) +
  geom_col(fill = "darkorange", alpha = 0.8) +
  geom_text(aes(label = round(RMSE, 3)), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Semi-parametric & Nonlinear Model Comparison — Cooling Load (Y2)",
       x = "Model", y = "RMSE (Test Set)") +
  theme_minimal()
ggsave(file.path(img_dir, "comparison_y2_rmse.png"),
       plot = p_rmse_y2, width = 10, height = 7, dpi = 150)
print(p_rmse_y2)

p_r2_y2 <- ggplot(results_y2, aes(x = reorder(Model, -R2), y = R2)) +
  geom_col(fill = "darkgreen", alpha = 0.7) +
  geom_text(aes(label = round(R2, 4)), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "R² Comparison — Cooling Load (Y2)",
       x = "Model", y = "R²") +
  theme_minimal()
ggsave(file.path(img_dir, "comparison_y2_r2.png"),
       plot = p_r2_y2, width = 10, height = 7, dpi = 150)
print(p_r2_y2)
```
