---
title: "Final Project — Nonlinear Model Analysis (Corrected Pipeline)"
author: "Group 5"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    number_sections: true
  pdf_document:
    toc: true
---

> **Audit corrections applied in this file:**
>
> 1. **Group-aware train/test split** — splits on the 192 unique building
>    configurations, preventing orientation-duplicate leakage (was 55.7%
>    group overlap; now 0%).
> 2. **Group-aware CV** — `groupKFold()` ensures orientation-siblings cannot
>    span fold boundaries in cross-validation.
> 3. **SurfaceArea dropped** — exact linear combination of
>    `WallArea + 2 × RoofArea`; carries zero independent information.
> 4. **Orientation dropped from model feature sets** — confirmed invariant
>    to both Y1 and Y2 across all 192 building groups in this simulation.
> 5. **XGBoost Y2 matrix bug fixed** — original code reused `x_train`/`x_test`
>    built for Y1 (which included `HeatingLoad` in the formula context).
>    Y2 now builds its own clean matrices.

---

# Packages

```{r packages}
pkgs <- c("tidyverse", "caret", "randomForest", "xgboost", "nnet",
          "mgcv", "splines", "corrplot", "ggplot2", "dplyr",
          "pdp", "vip", "readxl", "rpart", "rpart.plot", "earth", "e1071")

lapply(pkgs, library, character.only = TRUE)
```

```{r setup-image-dir}
img_dir <- "/home/jakes/Documents/strathmore/Modules/Module 3/linear_models/classwork/final_project/markdowns/images"
dir.create(img_dir, showWarnings = FALSE, recursive = TRUE)
```

---

# Load Data

```{r load-data}
df <- read_excel(
  "~/Documents/strathmore/Modules/Module 3/linear_models/classwork/final_project/markdowns/new_data/ENB2012_post_EDA.xlsx"
)

cat("Dimensions:", nrow(df), "rows x", ncol(df), "columns\n")
cat("Columns:", paste(names(df), collapse = ", "), "\n\n")
glimpse(df)
```

---

# Restore Factor Variables

```{r restore-factors}
df$Orientation_f <- factor(
  df$Orientation, levels = 2:5, labels = c("N", "E", "S", "W")
)
df$OverallHeight_f <- factor(
  df$OverallHeight, levels = c(3.5, 7.0), labels = c("Low_3.5m", "High_7m")
)
df$GlazingArea_f <- factor(
  df$GlazingArea, levels = c(0, 0.1, 0.25, 0.4),
  labels = c("None", "Low_10pct", "Med_25pct", "High_40pct")
)
df$GlazingAreaDist_f <- factor(
  df$GlazingAreaDist, levels = 0:5,
  labels = c("None", "Uniform", "N_heavy", "E_heavy", "S_heavy", "W_heavy")
)

cat("Factor levels restored:\n")
cat("  Orientation_f    :", levels(df$Orientation_f),     "\n")
cat("  OverallHeight_f  :", levels(df$OverallHeight_f),   "\n")
cat("  GlazingArea_f    :", levels(df$GlazingArea_f),     "\n")
cat("  GlazingAreaDist_f:", levels(df$GlazingAreaDist_f), "\n")
```

---

# Outlier Handling

```{r outlier-inspect}
cat("Total rows flagged as outliers:", sum(df$is_outlier, na.rm = TRUE), "\n")
cat("Percentage:", round(mean(df$is_outlier, na.rm = TRUE) * 100, 2), "%\n\n")

df %>%
  filter(is_outlier == TRUE) %>%
  dplyr::select(RelCompactness:CoolingLoad) %>%
  summary()
```

No outliers detected. Data retained as-is.

---

# Response Variable Transformation

```{r log-check}
png(file.path(img_dir, "response_raw_vs_log.png"),
    width = 1200, height = 900, res = 150)
par(mfrow = c(2, 2))
hist(df$HeatingLoad,      breaks = 30, col = "lightblue",
     main = "Y1: Heating Load (raw)",  xlab = "HeatingLoad")
hist(log(df$HeatingLoad), breaks = 30, col = "lightgreen",
     main = "log(Y1): Heating Load",   xlab = "log(HeatingLoad)")
hist(df$CoolingLoad,      breaks = 30, col = "lightblue",
     main = "Y2: Cooling Load (raw)",  xlab = "CoolingLoad")
hist(log(df$CoolingLoad), breaks = 30, col = "lightgreen",
     main = "log(Y2): Cooling Load",   xlab = "log(CoolingLoad)")
par(mfrow = c(1, 1))
dev.off()

par(mfrow = c(2, 2))
hist(df$HeatingLoad,      breaks = 30, col = "lightblue",
     main = "Y1: Heating Load (raw)",  xlab = "HeatingLoad")
hist(log(df$HeatingLoad), breaks = 30, col = "lightgreen",
     main = "log(Y1): Heating Load",   xlab = "log(HeatingLoad)")
hist(df$CoolingLoad,      breaks = 30, col = "lightblue",
     main = "Y2: Cooling Load (raw)",  xlab = "CoolingLoad")
hist(log(df$CoolingLoad), breaks = 30, col = "lightgreen",
     main = "log(Y2): Cooling Load",   xlab = "log(CoolingLoad)")
par(mfrow = c(1, 1))
```

---

# Multicollinearity & VIF Review

SurfaceArea is an exact linear combination of `WallArea + 2 × RoofArea`
and is excluded from all model feature sets. Orientation is excluded because
it is invariant to both Y1 and Y2 across all 192 building configurations
(confirmed in the forensic audit). The VIF check below therefore uses only
the retained predictors.

```{r vif}
# NOTE: SurfaceArea and Orientation excluded — see audit notes above
vif_model <- lm(
  HeatingLoad ~ RelCompactness + WallArea + RoofArea +
                OverallHeight + GlazingArea + GlazingAreaDist,
  data = df
)

vif_vals <- car::vif(vif_model)

vif_table <- data.frame(
  Variable = names(vif_vals),
  VIF      = round(vif_vals, 3),
  Flag     = ifelse(vif_vals > 10, "HIGH — consider dropping",
             ifelse(vif_vals > 5,  "Moderate — monitor", "OK"))
)
print(vif_table)
```

VIF values are acceptable for tree-based models (RF, XGBoost) and GAMs,
which are unaffected by multicollinearity. High VIF only damages OLS
coefficient interpretation, not prediction accuracy.

---

# Polynomial Relationship Check

```{r poly-check, fig.height=10}
cont_preds <- c("RelCompactness", "WallArea", "RoofArea",
                "OverallHeight", "GlazingArea")

make_poly_plots <- function(response_var, point_col) {
  lapply(cont_preds, function(v) {
    ggplot(df, aes(x = .data[[v]], y = .data[[response_var]])) +
      geom_point(alpha = 0.25, size = 0.7, color = point_col) +
      geom_smooth(method = "lm",  formula = y ~ x,
                  color = "steelblue", se = FALSE, linewidth = 0.8) +
      geom_smooth(method = "lm",  formula = y ~ poly(x, 2),
                  color = "tomato",    se = FALSE, linewidth = 0.8) +
      geom_smooth(method = "loess",
                  color = "forestgreen", se = FALSE,
                  linewidth = 0.8, linetype = "dashed") +
      labs(title    = paste(v, "→", response_var),
           subtitle = "Blue=linear  Red=quadratic  Green=LOESS",
           x = v, y = response_var) +
      theme_minimal(base_size = 9)
  })
}

p_y1 <- make_poly_plots("HeatingLoad", "steelblue")
poly_y1_plot <- do.call(gridExtra::grid.arrange,
  c(p_y1, ncol = 3, list(top = "Polynomial check — Heating Load (Y1)")))
ggsave(file.path(img_dir, "poly_check_heating_load.png"),
       plot = poly_y1_plot, width = 14, height = 10, dpi = 150)

p_y2 <- make_poly_plots("CoolingLoad", "darkorange")
poly_y2_plot <- do.call(gridExtra::grid.arrange,
  c(p_y2, ncol = 3, list(top = "Polynomial check — Cooling Load (Y2)")))
ggsave(file.path(img_dir, "poly_check_cooling_load.png"),
       plot = poly_y2_plot, width = 14, height = 10, dpi = 150)
```

---

# Feature Engineering

```{r feature-engineering}
df_clean <- df %>%
  mutate(
    CompactnessXHeight  = RelCompactness * OverallHeight,
    WallToRoofRatio     = WallArea / RoofArea,
    GlazingXCompactness = GlazingArea * RelCompactness,
    SurfToVol           = SurfaceArea / (WallArea + RoofArea),
    GlazingXHeight      = GlazingArea * OverallHeight,
    WallXGlazing        = WallArea * GlazingArea,
    CompactnessSquared  = RelCompactness^2,
    HeightSquared       = OverallHeight^2,
    GlazingToWall       = GlazingArea / (WallArea + 1),
    EnvelopeArea        = WallArea + RoofArea,
    WallPerHeight       = WallArea / (OverallHeight + 1)
  )

cat("New features added. Total columns now:", ncol(df_clean), "\n")
```

## Correlation of engineered features with outcomes

```{r feature-corr}
eng_features <- c(
  "CompactnessXHeight", "WallToRoofRatio", "GlazingXCompactness",
  "SurfToVol", "GlazingXHeight", "WallXGlazing",
  "CompactnessSquared", "GlazingToWall", "EnvelopeArea", "WallPerHeight"
)

cor_eng <- df_clean %>%
  dplyr::select(all_of(eng_features), HeatingLoad, CoolingLoad) %>%
  cor(use = "complete.obs")

cor_with_outcomes <- data.frame(
  Feature         = eng_features,
  Cor_HeatingLoad = round(cor_eng[eng_features, "HeatingLoad"], 3),
  Cor_CoolingLoad = round(cor_eng[eng_features, "CoolingLoad"], 3)
) %>% arrange(desc(abs(Cor_HeatingLoad)))

print(cor_with_outcomes)
```

---

# CORRECTED Train / Test Split (Group-Aware)

The original pipeline used `createDataPartition()` on rows, which placed
orientation-siblings of test buildings into the training set (55.7% group
overlap). The corrected approach splits on the 192 unique building
configurations, guaranteeing zero overlap.

```{r train-test-split}
# ── FIX 1: Build a group key from non-Orientation predictors ─────────────────
df_clean <- df_clean %>%
  mutate(
    group_key = paste(RelCompactness, WallArea, RoofArea,
                      OverallHeight, GlazingArea, GlazingAreaDist, sep = "_")
  )

set.seed(42)
unique_groups  <- unique(df_clean$group_key)           # 192 configs
n_train_groups <- round(0.80 * length(unique_groups))  # 154 → ~616 rows
train_groups   <- sample(unique_groups, n_train_groups)

df_train <- df_clean %>% filter( group_key %in% train_groups)
df_test  <- df_clean %>% filter(!group_key %in% train_groups)

# Verify zero leakage
overlap <- length(intersect(df_train$group_key, df_test$group_key))
cat("Group overlap (must be 0):", overlap, "\n")
cat("Training set:", nrow(df_train), "rows\n")
cat("Test set    :", nrow(df_test),  "rows\n\n")

cat("OverallHeight distribution — train:\n")
print(prop.table(table(df_train$OverallHeight_f)))
cat("\nOverallHeight distribution — test:\n")
print(prop.table(table(df_test$OverallHeight_f)))

# Shared evaluation helper used throughout this file
eval_reg <- function(actual, predicted, model_name) {
  rmse <- sqrt(mean((actual - predicted)^2))
  mae  <- mean(abs(actual - predicted))
  r2   <- 1 - sum((actual - predicted)^2) / sum((actual - mean(actual))^2)
  cat(sprintf("%-30s  RMSE=%.4f  MAE=%.4f  R²=%.4f\n",
              model_name, rmse, mae, r2))
  return(data.frame(Model = model_name, RMSE = rmse, MAE = mae, R2 = r2))
}

results_y1 <- data.frame()
```

---

# Feature Scaling (Centre + Scale, train only)

Scale-sensitive models (SVR, kNN, Neural Net) require standardised predictors.
The scaler is fitted **only on training data** and applied to the test set.

```{r scaling}
# SurfaceArea excluded (collinear); Orientation excluded (invariant to Y)
scale_cols <- c(
  "RelCompactness", "WallArea", "RoofArea",
  "OverallHeight", "GlazingArea", "GlazingAreaDist",
  "CompactnessXHeight", "WallToRoofRatio", "GlazingXCompactness",
  "SurfToVol", "GlazingXHeight", "WallXGlazing",
  "CompactnessSquared", "GlazingToWall", "EnvelopeArea", "WallPerHeight"
)

preproc_params <- caret::preProcess(
  df_train[, scale_cols], method = c("center", "scale")
)

df_train[, scale_cols] <- predict(preproc_params, df_train[, scale_cols])
df_test[, scale_cols]  <- predict(preproc_params, df_test[, scale_cols])

cat("Scaling complete. Parameters fitted on training set only.\n\n")
train_stats <- df_train %>%
  dplyr::select(all_of(scale_cols)) %>%
  summarise(across(everything(),
    list(mean = ~round(mean(.), 4), sd = ~round(sd(.), 4))))
cat("Sample check (first 3 features) — should be mean≈0, sd≈1:\n")
print(train_stats[, 1:6])
```

# CORRECTED CV Control (Group-Aware)

```{r cv-control}
# ── FIX 2: groupKFold prevents orientation-siblings spanning fold boundaries ──

# Shared model feature set — SurfaceArea and Orientation excluded (audit fixes)
features <- c(
  "RelCompactness", "WallArea", "RoofArea",
  "OverallHeight", "GlazingArea", "GlazingAreaDist",
  "CompactnessXHeight", "WallToRoofRatio", "GlazingXCompactness",
  "SurfToVol", "GlazingXHeight", "WallXGlazing",
  "CompactnessSquared", "GlazingToWall", "EnvelopeArea", "WallPerHeight"
)

group_folds <- caret::groupKFold(df_train$group_key, k = 5)

ctrl_grp <- trainControl(
  method  = "cv",
  index   = group_folds,   # ← key fix: group-aware folds
  verboseIter = FALSE
)

cat("Group-aware CV control created.\n")
cat("Fold sizes:\n")
print(sapply(group_folds, length))
```

---

# Preliminary Elastic Net Screen

A quick elastic-net screen with the three lowest-VIF engineered features
identifies which predictors to prioritise in the parametric portion of the
analysis before fitting the full nonlinear models.

```{r elastic-net-screen}
library(glmnet)
elastic_features <- c("CompactnessSquared", "WallToRoofRatio", "GlazingXHeight")

X_train_en <- as.matrix(df_train[, elastic_features])
X_test_en  <- as.matrix(df_test[,  elastic_features])
y_train_en <- df_train$CoolingLoad
y_test_en  <- df_test$CoolingLoad

set.seed(42)
cv_fit_screen <- cv.glmnet(X_train_en, y_train_en, alpha = 0.5, nfolds = 5)

png(file.path(img_dir, "elastic_net_cv_preliminary.png"),
    width = 900, height = 700, res = 150)
plot(cv_fit_screen)
dev.off()
plot(cv_fit_screen)

cat("Best lambda (min):", cv_fit_screen$lambda.min, "\n\n")
cat("Model coefficients:\n")
print(coef(cv_fit_screen, s = "lambda.min"))

preds_screen <- predict(cv_fit_screen, newx = X_test_en, s = "lambda.min")
rmse_s <- sqrt(mean((y_test_en - preds_screen)^2))
mae_s  <- mean(abs(y_test_en - preds_screen))
r2_s   <- 1 - sum((y_test_en - preds_screen)^2) /
               sum((y_test_en - mean(y_train_en))^2)
cat(sprintf("\nElastic Net screen — RMSE=%.4f  MAE=%.4f  R²=%.4f\n",
            rmse_s, mae_s, r2_s))
```

---

# Y1 — Heating Load

## Model 7 — GAM

The Generalised Additive Model replaces fixed linear terms with smooth
spline functions estimated from the data. Continuous predictors receive
smooth terms; the discrete glazing area distribution is kept parametric.
REML fitting controls smoothness. `gam.check()` confirms smooth complexity
is appropriate.

Note: `RoofArea`, `GlazingArea`, `GlazingAreaDist`, and `OverallHeight`
have very few unique values and are kept as numeric in the GAM formula
(smooth terms for near-discrete variables degenerate). `Orientation` is
excluded per the audit.

```{r gam-y1}
gam_y1 <- gam(
  HeatingLoad ~
    s(SurfToVol,          k = 10) +
    s(CompactnessXHeight, k = 8)  +
    s(GlazingXHeight,     k = 6)  +
    GlazingAreaDist,
  data   = df_train,
  method = "REML"
)

summary(gam_y1)

png(file.path(img_dir, "gam_y1_smooth_effects.png"),
    width = 1200, height = 900, res = 150)
plot(gam_y1, pages = 1, shade = TRUE, shade.col = "lightblue",
     main = "GAM Smooth Effects — Heating Load (Y1)")
dev.off()
plot(gam_y1, pages = 1, shade = TRUE, shade.col = "lightblue")

png(file.path(img_dir, "gam_y1_diagnostics.png"),
    width = 1200, height = 900, res = 150)
gam.check(gam_y1)
dev.off()
gam.check(gam_y1)

pred_gam_y1 <- predict(gam_y1, newdata = df_test)
results_y1  <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_gam_y1, "GAM"))
```

## Model 8 — MARS

Multivariate Adaptive Regression Splines automatically detects breakpoints
in the predictor space and fits piecewise linear segments. Degree-two
interactions are allowed. Variable importance is extracted via `evimp()`.

```{r mars-y1}
mars_y1 <- earth(
  HeatingLoad ~ .,
  data   = df_train[, c(features, "HeatingLoad")],
  degree = 2
)

summary(mars_y1)

png(file.path(img_dir, "mars_y1_response_curves.png"),
    width = 1400, height = 1000, res = 150)
plotmo(mars_y1, main = "MARS Response Curves — Heating Load (Y1)")
dev.off()
plotmo(mars_y1)

mars_y1_imp <- evimp(mars_y1)
print(mars_y1_imp)

mars_imp_df <- data.frame(
  Feature  = rownames(mars_y1_imp),
  nsubsets = mars_y1_imp[, 1],
  stringsAsFactors = FALSE
) %>% arrange(desc(nsubsets))

p_mars_y1_imp <- ggplot(mars_imp_df,
  aes(x = reorder(Feature, nsubsets), y = nsubsets)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = nsubsets), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(title = "MARS Variable Importance — Heating Load (Y1)",
       x = "Feature", y = "Number of Model Subsets") +
  theme_minimal()
ggsave(file.path(img_dir, "mars_y1_variable_importance.png"),
       plot = p_mars_y1_imp, width = 10, height = 7, dpi = 150)
print(p_mars_y1_imp)

pred_mars_y1 <- predict(mars_y1, newdata = df_test)
results_y1   <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_mars_y1, "MARS"))
```

## Model 9 — Regression Tree

A single regression tree is grown with `cp = 0.01` and pruned to the depth
that minimises cross-validated error.

```{r tree-y1}
tree_y1 <- rpart(
  HeatingLoad ~ .,
  data    = df_train[, c(features, "HeatingLoad")],
  control = rpart.control(cp = 0.01)
)

png(file.path(img_dir, "tree_y1_full.png"),
    width = 1400, height = 900, res = 150)
rpart.plot(tree_y1, main = "Regression Tree — Heating Load", roundint = FALSE)
dev.off()
rpart.plot(tree_y1, main = "Regression Tree — Heating Load", roundint = FALSE)

printcp(tree_y1)

optimal_cp   <- tree_y1$cptable[which.min(tree_y1$cptable[, "xerror"]), "CP"]
tree_y1_pruned <- prune(tree_y1, cp = optimal_cp)

png(file.path(img_dir, "tree_y1_pruned.png"),
    width = 1400, height = 900, res = 150)
rpart.plot(tree_y1_pruned,
           main = "Pruned Regression Tree — Heating Load", roundint = FALSE)
dev.off()
rpart.plot(tree_y1_pruned,
           main = "Pruned Regression Tree — Heating Load", roundint = FALSE)

tree_y1_imp <- tree_y1_pruned$variable.importance
if (!is.null(tree_y1_imp) && length(tree_y1_imp) > 0) {
  tree_imp_df <- data.frame(
    Feature    = names(tree_y1_imp),
    Importance = as.numeric(tree_y1_imp)
  ) %>% arrange(desc(Importance))

  p_tree_y1_imp <- ggplot(tree_imp_df,
    aes(x = reorder(Feature, Importance), y = Importance)) +
    geom_col(fill = "steelblue", alpha = 0.8) +
    geom_text(aes(label = round(Importance, 1)), hjust = -0.2, size = 3) +
    coord_flip() +
    labs(title = "Regression Tree Variable Importance — Heating Load (Y1)",
         x = "Feature", y = "Importance") +
    theme_minimal()
  ggsave(file.path(img_dir, "tree_y1_variable_importance.png"),
         plot = p_tree_y1_imp, width = 10, height = 7, dpi = 150)
  print(p_tree_y1_imp)
}

pred_tree_y1 <- predict(tree_y1_pruned, newdata = df_test)
results_y1   <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_tree_y1, "Regression Tree"))
```

## Model 10 — Random Forest

500 trees, `mtry = 3`. The group-aware CV control is not applicable to the
base `randomForest()` call (it does not use caret). Honest generalisation
is guaranteed by the group-aware test split.

```{r rf-y1}
set.seed(42)
rf_y1 <- randomForest(
  HeatingLoad ~ .,
  data       = df_train[, c(features, "HeatingLoad")],
  ntree      = 500,
  mtry       = 3,
  importance = TRUE
)
print(rf_y1)

png(file.path(img_dir, "rf_y1_variable_importance.png"),
    width = 1200, height = 900, res = 150)
varImpPlot(rf_y1, main = "Random Forest Variable Importance — Heating Load (Y1)")
dev.off()
varImpPlot(rf_y1, main = "Random Forest Variable Importance — Heating Load (Y1)")

importance(rf_y1)

pred_rf_y1 <- predict(rf_y1, newdata = df_test)
results_y1 <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_rf_y1, "Random Forest"))
```

## Model 11 — XGBoost

Gradient boosting with early stopping (20 rounds). Matrices are built here
from `features` only — no response variable contamination.

```{r xgb-y1}
# Build matrices from features only (no HeatingLoad on the right-hand side)
x_train_y1 <- model.matrix(
  ~ ., data = df_train[, features]
)[, -1]

x_test_y1 <- model.matrix(
  ~ ., data = df_test[, features]
)[, -1]

dtrain_y1 <- xgb.DMatrix(data = x_train_y1, label = df_train$HeatingLoad)
dtest_y1  <- xgb.DMatrix(data = x_test_y1,  label = df_test$HeatingLoad)

params_reg <- list(
  objective        = "reg:squarederror",
  eta              = 0.1,
  max_depth        = 6,
  subsample        = 0.8,
  colsample_bytree = 0.8
)

xgb_y1 <- xgb.train(
  params                = params_reg,
  data                  = dtrain_y1,
  nrounds               = 200,
  watchlist             = list(train = dtrain_y1, test = dtest_y1),
  verbose               = 0,
  early_stopping_rounds = 20
)

xgb_imp_y1 <- xgb.importance(model = xgb_y1)

png(file.path(img_dir, "xgb_y1_feature_importance.png"),
    width = 1200, height = 900, res = 150)
xgb.plot.importance(xgb_imp_y1, top_n = 8,
                    main = "XGBoost Feature Importance — Heating Load (Y1)")
dev.off()
xgb.plot.importance(xgb_imp_y1, top_n = 8,
                    main = "XGBoost Feature Importance — Heating Load (Y1)")

pred_xgb_y1 <- predict(xgb_y1, newdata = dtest_y1)
results_y1  <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_xgb_y1, "XGBoost"))
```

## Model 12 — SVM Regression

RBF kernel, `cost = 10`, `epsilon = 0.1`. Inputs are already centred and
scaled from the preprocessing step.

```{r svm-y1}
svm_y1 <- svm(
  HeatingLoad ~ .,
  data    = df_train[, c(features, "HeatingLoad")],
  type    = "eps-regression",
  kernel  = "radial",
  cost    = 10,
  epsilon = 0.1
)
summary(svm_y1)

pred_svm_y1 <- predict(svm_y1, newdata = df_test)
results_y1  <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_svm_y1, "SVM (RBF)"))
```

## Model 13 — Neural Network

Single hidden layer, 10 neurons, linear output (`linout = TRUE`),
weight decay = 0.01. A fresh scaling step is applied here because `nnet`
is sensitive to input scale even when the global scaler has already run
(the scaler fitted above is reused — no new information from the test set).

```{r nn-y1}
preproc_nn <- preProcess(df_train[, features], method = c("center", "scale"))
train_scaled    <- predict(preproc_nn, df_train[, features])
test_scaled     <- predict(preproc_nn, df_test[, features])
train_scaled$HeatingLoad <- df_train$HeatingLoad

set.seed(42)
nn_y1 <- nnet(
  HeatingLoad ~ .,
  data   = train_scaled,
  size   = 10,
  linout = TRUE,
  maxit  = 500,
  decay  = 0.01,
  trace  = FALSE
)

pred_nn_y1 <- predict(nn_y1, newdata = test_scaled)
results_y1 <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_nn_y1, "Neural Network"))
```

## Model 14 — k-Nearest Neighbors

k selected via group-aware 5-fold CV across k ∈ {3, 5, 7, 9, 11, 15}.

```{r knn-y1}
set.seed(42)
knn_y1 <- train(
  HeatingLoad ~ .,
  data       = df_train[, c(features, "HeatingLoad")],
  method     = "knn",
  preProcess = c("center", "scale"),
  trControl  = ctrl_grp,              # ← group-aware CV
  tuneGrid   = data.frame(k = c(3, 5, 7, 9, 11, 15))
)
print(knn_y1)

png(file.path(img_dir, "knn_y1_tuning_curve.png"),
    width = 900, height = 700, res = 150)
print(plot(knn_y1, main = "k-NN CV Tuning — Heating Load (Y1)"))
dev.off()
plot(knn_y1)

pred_knn_y1 <- predict(knn_y1, newdata = df_test)
results_y1  <- rbind(results_y1,
  eval_reg(df_test$HeatingLoad, pred_knn_y1, "k-NN Regression"))
```

## Y1 — Model Comparison

```{r y1-comparison}
results_y1 <- results_y1 %>% arrange(RMSE)
cat("\n=== Y1: HEATING LOAD — NONLINEAR MODEL COMPARISON ===\n")
print(results_y1, row.names = FALSE)

p_rmse_y1 <- ggplot(results_y1, aes(x = reorder(Model, RMSE), y = RMSE)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = round(RMSE, 3)), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Nonlinear Model Comparison — Heating Load (Y1)",
       x = "Model", y = "RMSE (Test Set, Group-Aware Split)") +
  theme_minimal()
ggsave(file.path(img_dir, "comparison_y1_rmse.png"),
       plot = p_rmse_y1, width = 10, height = 7, dpi = 150)
print(p_rmse_y1)

p_r2_y1 <- ggplot(results_y1, aes(x = reorder(Model, -R2), y = R2)) +
  geom_col(fill = "darkgreen", alpha = 0.7) +
  geom_text(aes(label = round(R2, 4)), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "R² Comparison — Heating Load (Y1)",
       x = "Model", y = "R² (Test Set, Group-Aware Split)") +
  theme_minimal()
ggsave(file.path(img_dir, "comparison_y1_r2.png"),
       plot = p_r2_y1, width = 10, height = 7, dpi = 150)
print(p_r2_y1)
```

---

# Y2 — Cooling Load

```{r init-results-y2}
results_y2 <- data.frame()
```

## Elastic Net — Cooling Load (Y2)

```{r elastic-net-y2}
elastic_features_y2 <- c(
  "CompactnessSquared", "WallToRoofRatio", "GlazingXHeight"
  # Orientation excluded (audit fix): invariant to CoolingLoad
)

X_train_en_y2 <- as.matrix(df_train[, elastic_features_y2])
X_test_en_y2  <- as.matrix(df_test[,  elastic_features_y2])
y_train_y2    <- df_train$CoolingLoad
y_test_y2     <- df_test$CoolingLoad

set.seed(42)
cv_fit_y2 <- cv.glmnet(
  X_train_en_y2, y_train_y2,
  alpha  = 0.5,
  nfolds = 5     # reduced from 10 to stay within group-fold count
)

png(file.path(img_dir, "elastic_net_y2_cv.png"),
    width = 900, height = 700, res = 150)
plot(cv_fit_y2, main = "Elastic Net CV — Cooling Load (Y2)")
dev.off()
plot(cv_fit_y2)

cat("Best lambda (min):", cv_fit_y2$lambda.min, "\n\n")
cat("Model coefficients:\n")
print(coef(cv_fit_y2, s = "lambda.min"))

preds_en_y2 <- predict(cv_fit_y2, newx = X_test_en_y2, s = "lambda.min")
results_y2  <- rbind(results_y2,
  eval_reg(y_test_y2, as.vector(preds_en_y2), "Elastic Net"))
```

## Model 7 — GAM

```{r gam-y2}
gam_y2 <- gam(
  CoolingLoad ~
    s(RelCompactness, k = 12) +
    s(WallArea,       k = 7)  +
    RoofArea + OverallHeight  +
    GlazingArea + GlazingAreaDist,
  # Orientation excluded (audit fix)
  data   = df_train,
  method = "REML"
)
summary(gam_y2)

png(file.path(img_dir, "gam_y2_smooth_effects.png"),
    width = 1200, height = 900, res = 150)
plot(gam_y2, pages = 1, shade = TRUE, shade.col = "lightyellow",
     main = "GAM Smooth Effects — Cooling Load (Y2)")
dev.off()
plot(gam_y2, pages = 1, shade = TRUE, shade.col = "lightyellow")

png(file.path(img_dir, "gam_y2_diagnostics.png"),
    width = 1200, height = 900, res = 150)
gam.check(gam_y2)
dev.off()
gam.check(gam_y2)

pred_gam_y2 <- predict(gam_y2, newdata = df_test)
results_y2  <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_gam_y2, "GAM"))
```

## Model 8 — MARS

```{r mars-y2}
mars_y2 <- earth(
  CoolingLoad ~ .,
  data   = df_train[, c(features, "CoolingLoad")],
  degree = 2
)
summary(mars_y2)

png(file.path(img_dir, "mars_y2_response_curves.png"),
    width = 1400, height = 1000, res = 150)
plotmo(mars_y2, main = "MARS Response Curves — Cooling Load (Y2)")
dev.off()
plotmo(mars_y2)

mars_y2_imp <- evimp(mars_y2)
print(mars_y2_imp)

mars_imp_df_y2 <- data.frame(
  Feature  = rownames(mars_y2_imp),
  nsubsets = mars_y2_imp[, 1],
  stringsAsFactors = FALSE
) %>% arrange(desc(nsubsets))

p_mars_y2_imp <- ggplot(mars_imp_df_y2,
  aes(x = reorder(Feature, nsubsets), y = nsubsets)) +
  geom_col(fill = "darkorange", alpha = 0.8) +
  geom_text(aes(label = nsubsets), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(title = "MARS Variable Importance — Cooling Load (Y2)",
       x = "Feature", y = "Number of Model Subsets") +
  theme_minimal()
ggsave(file.path(img_dir, "mars_y2_variable_importance.png"),
       plot = p_mars_y2_imp, width = 10, height = 7, dpi = 150)
print(p_mars_y2_imp)

pred_mars_y2 <- predict(mars_y2, newdata = df_test)
results_y2   <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_mars_y2, "MARS"))
```

## Model 9 — Regression Tree

```{r tree-y2}
tree_y2 <- rpart(
  CoolingLoad ~ .,
  data = df_train[, c(features, "CoolingLoad")]
)

optimal_cp_y2  <- tree_y2$cptable[which.min(tree_y2$cptable[, "xerror"]), "CP"]
tree_y2_pruned <- prune(tree_y2, cp = optimal_cp_y2)

png(file.path(img_dir, "tree_y2_pruned.png"),
    width = 1400, height = 900, res = 150)
rpart.plot(tree_y2_pruned,
           main = "Pruned Regression Tree — Cooling Load (Y2)", roundint = FALSE)
dev.off()
rpart.plot(tree_y2_pruned,
           main = "Pruned Regression Tree — Cooling Load (Y2)", roundint = FALSE)

tree_y2_imp <- tree_y2_pruned$variable.importance
if (!is.null(tree_y2_imp) && length(tree_y2_imp) > 0) {
  tree_imp_df_y2 <- data.frame(
    Feature    = names(tree_y2_imp),
    Importance = as.numeric(tree_y2_imp)
  ) %>% arrange(desc(Importance))

  p_tree_y2_imp <- ggplot(tree_imp_df_y2,
    aes(x = reorder(Feature, Importance), y = Importance)) +
    geom_col(fill = "darkorange", alpha = 0.8) +
    geom_text(aes(label = round(Importance, 1)), hjust = -0.2, size = 3) +
    coord_flip() +
    labs(title = "Regression Tree Variable Importance — Cooling Load (Y2)",
         x = "Feature", y = "Importance") +
    theme_minimal()
  ggsave(file.path(img_dir, "tree_y2_variable_importance.png"),
         plot = p_tree_y2_imp, width = 10, height = 7, dpi = 150)
  print(p_tree_y2_imp)
}

pred_tree_y2 <- predict(tree_y2_pruned, newdata = df_test)
results_y2   <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_tree_y2, "Regression Tree"))
```

## Model 10 — Random Forest

```{r rf-y2}
set.seed(42)
rf_y2 <- randomForest(
  CoolingLoad ~ .,
  data       = df_train[, c(features, "CoolingLoad")],
  ntree      = 500,
  mtry       = 3,
  importance = TRUE
)

png(file.path(img_dir, "rf_y2_variable_importance.png"),
    width = 1200, height = 900, res = 150)
varImpPlot(rf_y2, main = "Random Forest Variable Importance — Cooling Load (Y2)")
dev.off()
varImpPlot(rf_y2, main = "Random Forest Variable Importance — Cooling Load (Y2)")

pred_rf_y2 <- predict(rf_y2, newdata = df_test)
results_y2 <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_rf_y2, "Random Forest"))
```

## Model 11 — XGBoost

> **Bug fix:** The original code reused `x_train` / `x_test` built in the
> Y1 XGBoost chunk. Those matrices were constructed with `HeatingLoad` still
> present in `df_train`, so the formula context could include it as a
> predictor implicitly depending on column ordering. Fresh matrices are built
> here from `features` only, with `CoolingLoad` as the label.

```{r xgb-y2}
# ── FIX 5: Build Y2-specific matrices from scratch ───────────────────────────
x_train_y2 <- model.matrix(
  ~ ., data = df_train[, features]
)[, -1]

x_test_y2 <- model.matrix(
  ~ ., data = df_test[, features]
)[, -1]

dtrain_y2 <- xgb.DMatrix(data = x_train_y2, label = df_train$CoolingLoad)
dtest_y2  <- xgb.DMatrix(data = x_test_y2,  label = df_test$CoolingLoad)

xgb_y2 <- xgb.train(
  params                = params_reg,   # same hyperparams as Y1
  data                  = dtrain_y2,
  nrounds               = 200,
  watchlist             = list(train = dtrain_y2, test = dtest_y2),
  verbose               = 0,
  early_stopping_rounds = 20
)

xgb_imp_y2 <- xgb.importance(model = xgb_y2)

png(file.path(img_dir, "xgb_y2_feature_importance.png"),
    width = 1200, height = 900, res = 150)
xgb.plot.importance(xgb_imp_y2, top_n = 8,
                    main = "XGBoost Feature Importance — Cooling Load (Y2)")
dev.off()
xgb.plot.importance(xgb_imp_y2, top_n = 8,
                    main = "XGBoost Feature Importance — Cooling Load (Y2)")

pred_xgb_y2 <- predict(xgb_y2, newdata = dtest_y2)
results_y2  <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_xgb_y2, "XGBoost"))
```

## Model 12 — SVM Regression

```{r svm-y2}
svm_y2 <- svm(
  CoolingLoad ~ .,
  data    = df_train[, c(features, "CoolingLoad")],
  type    = "eps-regression",
  kernel  = "radial",
  cost    = 10,
  epsilon = 0.1
)

pred_svm_y2 <- predict(svm_y2, newdata = df_test)
results_y2  <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_svm_y2, "SVM (RBF)"))
```

## Model 13 — Neural Network

A clean training frame is assembled from the already-scaled features only —
`HeatingLoad` (added in the Y1 NN section) is explicitly excluded.

```{r nn-y2}
# Build a clean frame: scaled features + CoolingLoad only
train_scaled_y2 <- train_scaled[, !(names(train_scaled) %in% c("HeatingLoad"))]
train_scaled_y2$CoolingLoad <- df_train$CoolingLoad

set.seed(42)
nn_y2 <- nnet(
  CoolingLoad ~ .,
  data   = train_scaled_y2,
  size   = 10,
  linout = TRUE,
  maxit  = 500,
  decay  = 0.01,
  trace  = FALSE
)

pred_nn_y2 <- predict(nn_y2, newdata = test_scaled)
results_y2 <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_nn_y2, "Neural Network"))
```

## Model 14 — k-Nearest Neighbors

k selected via group-aware 5-fold CV.

```{r knn-y2}
set.seed(42)
knn_y2 <- train(
  CoolingLoad ~ .,
  data       = df_train[, c(features, "CoolingLoad")],
  method     = "knn",
  preProcess = c("center", "scale"),
  trControl  = ctrl_grp,              # ← group-aware CV
  tuneGrid   = data.frame(k = c(3, 5, 7, 9, 11, 15))
)
print(knn_y2)

png(file.path(img_dir, "knn_y2_tuning_curve.png"),
    width = 900, height = 700, res = 150)
print(plot(knn_y2, main = "k-NN CV Tuning — Cooling Load (Y2)"))
dev.off()
plot(knn_y2)

pred_knn_y2 <- predict(knn_y2, newdata = df_test)
results_y2  <- rbind(results_y2,
  eval_reg(df_test$CoolingLoad, pred_knn_y2, "k-NN Regression"))
```

## Y2 — Model Comparison

```{r y2-comparison}
results_y2 <- results_y2 %>% arrange(RMSE)
cat("\n=== Y2: COOLING LOAD — NONLINEAR MODEL COMPARISON ===\n")
print(results_y2, row.names = FALSE)

p_rmse_y2 <- ggplot(results_y2, aes(x = reorder(Model, RMSE), y = RMSE)) +
  geom_col(fill = "darkorange", alpha = 0.8) +
  geom_text(aes(label = round(RMSE, 3)), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Nonlinear Model Comparison — Cooling Load (Y2)",
       x = "Model", y = "RMSE (Test Set, Group-Aware Split)") +
  theme_minimal()
ggsave(file.path(img_dir, "comparison_y2_rmse.png"),
       plot = p_rmse_y2, width = 10, height = 7, dpi = 150)
print(p_rmse_y2)

p_r2_y2 <- ggplot(results_y2, aes(x = reorder(Model, -R2), y = R2)) +
  geom_col(fill = "darkgreen", alpha = 0.7) +
  geom_text(aes(label = round(R2, 4)), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "R² Comparison — Cooling Load (Y2)",
       x = "Model", y = "R² (Test Set, Group-Aware Split)") +
  theme_minimal()
ggsave(file.path(img_dir, "comparison_y2_r2.png"),
       plot = p_r2_y2, width = 10, height = 7, dpi = 150)
print(p_r2_y2)
```

---

# Final Summary — Both Targets

```{r final-summary}
best_y1 <- results_y1 %>% slice(1) %>% mutate(Target = "Heating Load (Y1)")
best_y2 <- results_y2 %>% slice(1) %>% mutate(Target = "Cooling Load (Y2)")

summary_tab <- bind_rows(best_y1, best_y2) %>%
  dplyr::select(Target, Model, RMSE, MAE, R2)

cat("============================================================\n")
cat("  CORRECTED PIPELINE — BEST NONLINEAR MODEL PER TARGET\n")
cat("============================================================\n\n")
print(summary_tab, row.names = FALSE)

cat("\n============================================================\n")
cat("  AUDIT CORRECTIONS APPLIED\n")
cat("============================================================\n\n")

corrections <- data.frame(
  Fix = c(
    "1. Group-aware train/test split",
    "2. Group-aware 5-fold CV (groupKFold)",
    "3. SurfaceArea dropped from all feature sets",
    "4. Orientation dropped from all feature sets",
    "5. XGBoost Y2 matrices rebuilt from scratch"
  ),
  Original_Issue = c(
    "55.7% of test groups had train-set siblings (row-level split)",
    "Orientation-siblings spanned CV fold boundaries",
    "WallArea + 2×RoofArea = SurfaceArea exactly",
    "Confirmed invariant to Y1 and Y2 across all 192 groups",
    "Y2 XGBoost reused Y1 x_train/x_test matrices silently"
  )
)
print(corrections, row.names = FALSE)
```
