---
title: "ENB2012 — Feature Engineering & Preprocessing"
author: "Josh, Kosen, Kariuki"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 11, fig.height = 6)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 11, fig.height = 6)
```

---

# Package Installation & Loading

```{r packages}
pkgs <- c(
  "tidyverse", "caret", "MASS", "car", "moments",
  "readxl", "writexl", "ggplot2", "gridExtra",
  "corrplot", "GGally", "mgcv", "splines"
)

not_installed <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(not_installed) > 0) {
  install.packages(not_installed, dependencies = TRUE)
}

invisible(lapply(pkgs, library, character.only = TRUE))

cat("All packages loaded successfully.\n")
```

---

# Load Data from EDA Output

We load the post-EDA Excel file produced at the end of `EDA.Rmd`.
This file already has renamed columns, factor variants (`_f`), and
an `is_outlier` flag column.

```{r load-data}
df <- read_excel("~/Documents/strathmore/Modules/Module 3/linear_models/classwork/final_project/markdowns/new_data/ENB2012_post_EDA.xlsx")

cat("Dimensions:", nrow(df), "rows x", ncol(df), "columns\n")
cat("Columns:", paste(names(df), collapse = ", "), "\n\n")

glimpse(df)
```

---

# Restore Factor Variables

Excel does not preserve R factor types. We restore them from the raw
numeric columns that were carried over.

```{r restore-factors}
df$Orientation_f <- factor(
  df$Orientation,
  levels = 2:5,
  labels = c("N", "E", "S", "W")
)

df$OverallHeight_f <- factor(
  df$OverallHeight,
  levels = c(3.5, 7.0),
  labels = c("Low_3.5m", "High_7m")
)

df$GlazingArea_f <- factor(
  df$GlazingArea,
  levels = c(0, 0.1, 0.25, 0.4),
  labels = c("None", "Low_10pct", "Med_25pct", "High_40pct")
)

df$GlazingAreaDist_f <- factor(
  df$GlazingAreaDist,
  levels = 0:5,
  labels = c("None", "Uniform", "N_heavy", "E_heavy", "S_heavy", "W_heavy")
)

cat("Factor levels restored:\n")
cat("  Orientation_f    :", levels(df$Orientation_f), "\n")
cat("  OverallHeight_f  :", levels(df$OverallHeight_f), "\n")
cat("  GlazingArea_f    :", levels(df$GlazingArea_f), "\n")
cat("  GlazingAreaDist_f:", levels(df$GlazingAreaDist_f), "\n")
```

---

# Outlier Handling

The `is_outlier` flag was computed in EDA using the IQR method across
all numeric predictors and responses. We inspect the flagged rows and
decide on a strategy.

## Inspect flagged rows

```{r outlier-inspect}
cat("Total rows flagged as outliers:", sum(df$is_outlier, na.rm = TRUE), "\n")
cat("Percentage:", round(mean(df$is_outlier, na.rm = TRUE) * 100, 2), "%\n\n")

# Preview flagged rows
df %>%
  filter(is_outlier == TRUE) %>%
  dplyr::select(RelCompactness:CoolingLoad) %>%
  summary()
```

Since there are no outliers, we keep the data as it is.

# Response Variable Transformation

## Visual check: raw vs log

```{r log-check}
par(mfrow = c(2, 2))

hist(df_clean$HeatingLoad,      breaks = 30, col = "lightblue",
     main = "Y1: Heating Load (raw)", xlab = "HeatingLoad")
hist(log(df_clean$HeatingLoad), breaks = 30, col = "lightgreen",
     main = "log(Y1): Heating Load",  xlab = "log(HeatingLoad)")

hist(df_clean$CoolingLoad,      breaks = 30, col = "lightblue",
     main = "Y2: Cooling Load (raw)", xlab = "CoolingLoad")
hist(log(df_clean$CoolingLoad), breaks = 30, col = "lightgreen",
     main = "log(Y2): Cooling Load",  xlab = "log(CoolingLoad)")

par(mfrow = c(1, 1))
```

## Box-Cox optimal lambda — Y1 (Heating Load)

Box-Cox is fitted on a numeric-only subset (SurfaceArea excluded due
to perfect collinearity identified in EDA; Orientation and
GlazingAreaDist kept as numeric here for the Box-Cox linear model).

```{r boxcox-y1}
df_bc_y1 <- df_clean %>%
  dplyr::select(RelCompactness, WallArea, RoofArea,
         OverallHeight, Orientation, GlazingArea,
         GlazingAreaDist, HeatingLoad)

bc_y1      <- MASS::boxcox(lm(HeatingLoad ~ ., data = df_bc_y1),
                            lambda = seq(-2, 2, 0.05), plotit = TRUE)
lambda_y1  <- bc_y1$x[which.max(bc_y1$y)]

cat("Optimal Box-Cox lambda for Y1 (Heating Load):", lambda_y1, "\n")

# Apply transformation
if (abs(lambda_y1) < 0.05) {
  df_clean$HeatingLoad_bc <- log(df_clean$HeatingLoad)
  cat("Applied: log transformation (lambda ≈ 0)\n")
} else {
  df_clean$HeatingLoad_bc <- (df_clean$HeatingLoad ^ lambda_y1 - 1) / lambda_y1
  cat("Applied: Box-Cox with lambda =", lambda_y1, "\n")
}
```

## Box-Cox optimal lambda — Y2 (Cooling Load)

```{r boxcox-y2}
df_bc_y2 <- df_clean %>%
  dplyr::select(RelCompactness, WallArea, RoofArea,
         OverallHeight, Orientation, GlazingArea,
         GlazingAreaDist, CoolingLoad)

bc_y2      <- MASS::boxcox(lm(CoolingLoad ~ ., data = df_bc_y2),
                            lambda = seq(-2, 2, 0.05), plotit = TRUE)
lambda_y2  <- bc_y2$x[which.max(bc_y2$y)]

cat("Optimal Box-Cox lambda for Y2 (Cooling Load):", lambda_y2, "\n")

if (abs(lambda_y2) < 0.05) {
  df_clean$CoolingLoad_bc <- log(df_clean$CoolingLoad)
  cat("Applied: log transformation (lambda ≈ 0)\n")
} else {
  df_clean$CoolingLoad_bc <- (df_clean$CoolingLoad ^ lambda_y2 - 1) / lambda_y2
  cat("Applied: Box-Cox with lambda =", lambda_y2, "\n")
}
```

## QQ-plots after transformation

```{r qq-transformed}
par(mfrow = c(2, 2))

qqnorm(df_clean$HeatingLoad,    main = "Q-Q: Y1 raw")
qqline(df_clean$HeatingLoad,    col = "red", lwd = 2)

qqnorm(df_clean$HeatingLoad_bc, main = "Q-Q: Y1 Box-Cox transformed")
qqline(df_clean$HeatingLoad_bc, col = "steelblue", lwd = 2)

qqnorm(df_clean$CoolingLoad,    main = "Q-Q: Y2 raw")
qqline(df_clean$CoolingLoad,    col = "red", lwd = 2)

qqnorm(df_clean$CoolingLoad_bc, main = "Q-Q: Y2 Box-Cox transformed")
qqline(df_clean$CoolingLoad_bc, col = "steelblue", lwd = 2)

par(mfrow = c(1, 1))
```

---

# Multicollinearity & VIF Review

SurfaceArea was found in EDA to be perfectly collinear with
WallArea + RoofArea. We drop it from the modelling set and confirm
VIF scores are acceptable for the remaining predictors.

```{r vif}
vif_model <- lm(
  HeatingLoad ~ RelCompactness + WallArea + RoofArea +
                OverallHeight + Orientation + GlazingArea + GlazingAreaDist,
  data = df_clean
)

vif_vals <- car::vif(vif_model)

vif_table <- data.frame(
  Variable = names(vif_vals),
  VIF      = round(vif_vals, 3),
  Flag     = ifelse(vif_vals > 10, "HIGH — consider dropping",
             ifelse(vif_vals > 5,  "Moderate — monitor", "OK"))
)

print(vif_table)
```
We will keep them all, and accept the VIF. This is because Random Forest, XGBoost, and GAMs are not affected by multicollinearity at all. High VIF only damages OLS coefficient interpretation, not prediction accuracy.
---

# Polynomial Relationship Check (all predictors)

We extend the single-predictor check from `preprocessing.Rmd` to
cover all continuous predictors against both outcomes, so we know
which quadratic terms are worth including as engineered features.

```{r poly-check, fig.height=10}
cont_preds <- c("RelCompactness", "WallArea", "RoofArea",
                "OverallHeight", "GlazingArea")

make_poly_plots <- function(response_var, point_col) {
  lapply(cont_preds, function(v) {
    ggplot(df_clean, aes(x = .data[[v]], y = .data[[response_var]])) +
      geom_point(alpha = 0.25, size = 0.7, color = point_col) +
      geom_smooth(method = "lm",  formula = y ~ x,
                  color = "steelblue", se = FALSE, linewidth = 0.8) +
      geom_smooth(method = "lm",  formula = y ~ poly(x, 2),
                  color = "tomato", se = FALSE, linewidth = 0.8) +
      geom_smooth(method = "loess",
                  color = "forestgreen", se = FALSE,
                  linewidth = 0.8, linetype = "dashed") +
      labs(title = paste(v, "→", response_var),
           subtitle = "Blue=linear  Red=quadratic  Green=LOESS",
           x = v, y = response_var) +
      theme_minimal(base_size = 9)
  })
}

p_y1 <- make_poly_plots("HeatingLoad", "steelblue")
do.call(gridExtra::grid.arrange, c(p_y1, ncol = 3,
  list(top = "Polynomial check — Heating Load (Y1)")))

p_y2 <- make_poly_plots("CoolingLoad", "darkorange")
do.call(gridExtra::grid.arrange, c(p_y2, ncol = 3,
  list(top = "Polynomial check — Cooling Load (Y2)")))
```

---

# Feature Engineering

## Interaction and ratio features

Physically motivated engineered features based on the scatter plots
and domain interpretation from EDA.

```{r feature-engineering}
df_clean <- df_clean %>%
  mutate(
    # --- Original four from preprocessing.Rmd ---
    CompactnessXHeight  = RelCompactness * OverallHeight,
    WallToRoofRatio     = WallArea / RoofArea,
    GlazingXCompactness = GlazingArea * RelCompactness,
    SurfToVol           = SurfaceArea / (WallArea + RoofArea),

    # --- Additional physically motivated features ---
    GlazingXHeight      = GlazingArea * OverallHeight,
    # Tall buildings with more glazing lose/gain disproportionately more heat

    WallXGlazing        = WallArea * GlazingArea,
    # Interaction between exposed wall and window area

    CompactnessSquared  = RelCompactness^2,
    # Captures the nonlinear compactness-load relationship seen in poly plots

    HeightSquared       = OverallHeight^2,
    # OverallHeight is binary (3.5 / 7.0) so this is less critical,
    # but retained for completeness in polynomial model pipelines

    GlazingToWall       = GlazingArea / (WallArea + 1),
    # Facade transparency ratio: proportion of wall that is glass

    EnvelopeArea        = WallArea + RoofArea,
    # Total heat-exchange envelope (SurfaceArea minus floor, approx)

    WallPerHeight       = WallArea / (OverallHeight + 1)
    # Normalised wall exposure per unit height
  )

cat("New features added. Total columns now:", ncol(df_clean), "\n")
cat("Engineered feature names:\n")
cat(paste(" -",
  c("CompactnessXHeight","WallToRoofRatio","GlazingXCompactness",
    "SurfToVol","GlazingXHeight","WallXGlazing",
    "CompactnessSquared","HeightSquared","GlazingToWall",
    "EnvelopeArea","WallPerHeight")), sep = "\n")
```

## Correlation of engineered features with outcomes

Quick screen to confirm the new features are informative before
they enter modelling.

```{r feature-corr}
eng_features <- c(
  "CompactnessXHeight","WallToRoofRatio","GlazingXCompactness",
  "SurfToVol","GlazingXHeight","WallXGlazing",
  "CompactnessSquared","GlazingToWall","EnvelopeArea","WallPerHeight"
)

cor_eng <- df_clean %>%
  dplyr::select(all_of(eng_features), HeatingLoad, CoolingLoad) %>%
  cor(use = "complete.obs")

# Show only correlations with the two outcomes
cor_with_outcomes <- data.frame(
  Feature        = eng_features,
  Cor_HeatingLoad = round(cor_eng[eng_features, "HeatingLoad"], 3),
  Cor_CoolingLoad = round(cor_eng[eng_features, "CoolingLoad"], 3)
) %>%
  arrange(desc(abs(Cor_HeatingLoad)))

print(cor_with_outcomes)
```

---

# Train / Test Split

The split is done **before** scaling. The scaler is fitted only on
training data and then applied to the test set — this prevents data
leakage. We stratify on `OverallHeight_f` because it is the most
dominant predictor identified in EDA.

```{r train-test-split}
set.seed(42)

train_idx <- caret::createDataPartition(
  df_clean$OverallHeight_f,
  p    = 0.80,
  list = FALSE
)

df_train <- df_clean[ train_idx, ]
df_test  <- df_clean[-train_idx, ]

cat("Training set:", nrow(df_train), "rows\n")
cat("Test set    :", nrow(df_test),  "rows\n")

# Confirm stratification held
cat("\nOverallHeight distribution — train:\n")
print(prop.table(table(df_train$OverallHeight_f)))

cat("\nOverallHeight distribution — test:\n")
print(prop.table(table(df_test$OverallHeight_f)))
```

---

# Feature Scaling (Centre + Scale)

Scale-sensitive models (Ridge, Lasso, SVR, KNN, Neural Net) require
standardised predictors. Tree-based models (Random Forest, XGBoost)
do not require it but are not harmed by it, so we scale a single
unified numeric matrix.

Scaling is fitted on **training data only**.

```{r scaling}
# All numeric predictors — original + engineered
# Excludes: outcomes, factor columns, is_outlier flag, SurfaceArea (collinear)
scale_cols <- c(
  "RelCompactness", "WallArea", "RoofArea",
  "OverallHeight", "Orientation", "GlazingArea", "GlazingAreaDist",
  "CompactnessXHeight", "WallToRoofRatio", "GlazingXCompactness",
  "SurfToVol", "GlazingXHeight", "WallXGlazing",
  "CompactnessSquared", "GlazingToWall", "EnvelopeArea", "WallPerHeight"
)

# Fit scaler on training set
preproc_params <- caret::preProcess(
  df_train[, scale_cols],
  method = c("center", "scale")
)

# Apply to both sets
df_train[, scale_cols] <- predict(preproc_params, df_train[, scale_cols])
df_test[, scale_cols]  <- predict(preproc_params, df_test[, scale_cols])

cat("Scaling complete. Parameters fitted on training set only.\n\n")

# Confirm: training set should now have mean ≈ 0, sd ≈ 1
train_stats <- df_train %>%
  dplyr::select(all_of(scale_cols)) %>%
  summarise(across(everything(),
    list(mean = ~round(mean(.), 4), sd = ~round(sd(.), 4))))

cat("Sample check (first 3 features) — should be mean≈0, sd≈1:\n")
print(train_stats[, 1:6])
```

---

# Dummy Encoding for Regression Models

Some models (OLS, Ridge, Lasso, SVR) require fully numeric input.
We create a dummy-encoded version of the dataset for those models.
Tree-based and GAM models will use the factor columns directly.

```{r dummy-encode}
# Use model.matrix — this one-hot encodes all factor_f columns
# Drop the intercept column (first column from model.matrix)
dummy_formula <- ~ RelCompactness + WallArea + RoofArea +
                   OverallHeight + GlazingArea +
                   Orientation_f + GlazingAreaDist_f +
                   CompactnessXHeight + WallToRoofRatio + GlazingXCompactness +
                   SurfToVol + GlazingXHeight + WallXGlazing +
                   CompactnessSquared + GlazingToWall + EnvelopeArea + WallPerHeight

X_train_dummy <- model.matrix(dummy_formula, data = df_train)[, -1]
X_test_dummy  <- model.matrix(dummy_formula, data = df_test)[, -1]

cat("Dummy-encoded training matrix:", nrow(X_train_dummy), "x",
    ncol(X_train_dummy), "\n")
cat("Dummy-encoded test matrix    :", nrow(X_test_dummy),  "x",
    ncol(X_test_dummy),  "\n")
cat("Columns:", paste(colnames(X_train_dummy), collapse = ", "), "\n")
```

---

# Final Dataset Summary

```{r final-summary}
cat("============================================================\n")
cat("  PREPROCESSING COMPLETE — SUMMARY\n")
cat("============================================================\n\n")

cat("Source file   : ENB2012_post_EDA.xlsx\n")
cat("Outlier method: Winsorising (1st–99th pct) on continuous predictors\n")
cat("Y1 transform  : Box-Cox (lambda =", lambda_y1, ")\n")
cat("Y2 transform  : Box-Cox (lambda =", lambda_y2, ")\n")
cat("SurfaceArea   : DROPPED (perfectly collinear with WallArea + RoofArea)\n")
cat("Scaling       : Centre + scale, fit on train only\n\n")

cat("Engineered features (11):\n")
cat("  CompactnessXHeight, WallToRoofRatio, GlazingXCompactness,\n")
cat("  SurfToVol, GlazingXHeight, WallXGlazing, CompactnessSquared,\n")
cat("  GlazingToWall, EnvelopeArea, WallPerHeight, HeightSquared\n\n")

cat("Training rows:", nrow(df_train), "\n")
cat("Test rows    :", nrow(df_test),  "\n")
cat("Total columns in df_clean:", ncol(df_clean), "\n")
cat("Dummy-encoded predictor columns:", ncol(X_train_dummy), "\n\n")

cat("Objects available for modelling:\n")
cat("  df_train       — scaled training dataframe (factors intact)\n")
cat("  df_test        — scaled test dataframe (factors intact)\n")
cat("  X_train_dummy  — numeric matrix, dummy-encoded (for OLS/Ridge/SVR)\n")
cat("  X_test_dummy   — numeric matrix, dummy-encoded (for OLS/Ridge/SVR)\n")
cat("  preproc_params — caret preProcess object (for inverse-transforming)\n")
cat("  lambda_y1      — Box-Cox lambda for HeatingLoad\n")
cat("  lambda_y2      — Box-Cox lambda for CoolingLoad\n")
```

---

# Export

```{r export}
library(writexl)

write_xlsx(
  list(
    full_clean = df_clean,
    train      = df_train,
    test       = df_test
  ),
  "new_data/ENB2012_preprocessed.xlsx"
)

# Also save the dummy matrices as CSV for Python interoperability
write.csv(as.data.frame(X_train_dummy), "new_data/X_train_dummy.csv", row.names = FALSE)
write.csv(as.data.frame(X_test_dummy),  "new_data/X_test_dummy.csv",  row.names = FALSE)

# Save the scaler params so they can be re-applied to new data
saveRDS(preproc_params, "new_data/preproc_params.rds")
saveRDS(list(lambda_y1 = lambda_y1, lambda_y2 = lambda_y2), "new_data/boxcox_lambdas.rds")

cat("Exported:\n")
cat("  ENB2012_preprocessed.xlsx  — full_clean / train / test sheets\n")
cat("  X_train_dummy.csv          — dummy-encoded train matrix\n")
cat("  X_test_dummy.csv           — dummy-encoded test matrix\n")
cat("  preproc_params.rds         — caret scaler object\n")
cat("  boxcox_lambdas.rds         — Box-Cox lambda values\n")
```
